### 2018-07-05 manuscript.tex
initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% LIVECOMS ARTICLE TEMPLATE FOR BEST PRACTICES GUIDE
%%% ADAPTED FROM ELIFE ARTICLE TEMPLATE (8/10/2017)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE
\documentclass[9pt,tutorial,lineno,onehalfspacing]{livecoms}
% Use the 'onehalfspacing' option for 1.5 line spacing
% Use the 'doublespacing' option for 2.0 line spacing
% Use the 'lineno' option for adding line numbers.
% The 'bestpractices' option for indicates that this is a best practices guide.
% Omit the bestpractices option to remove the marking as a LiveCoMS paper.
% Please note that these options may affect formatting.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORTANT USER CONFIGURATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\versionnumber}{0.1}
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Learning PyEMMA : v\versionnumber}

\author[1\authfn{1}*]{AUTH0}
\author[1\authfn{1}*]{AUTH1}
\author[1*]{AUTH2}
\affil[1]{AFFILIATION}

\corr{auth0@email}{A0}
\corr{auth1@email}{A1}
\corr{auth2@email}{A2}

\contrib[\authfn{1}]{These authors contributed equally to this work}

\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}
\maketitle

\begin{abstract}
This tutorial provides an introduction to the PyEMMA project. Using Jupyter notebooks and accompanying screencasts, we will guide you through the basic functionality as well as the more common advanced mechanisms. Short exercises to self check your learning progress and a notebook on troubleshooting complete this basic introduction.
\end{abstract}

\end{frontmatter}

\section{Introduction}

PyEMMA~\cite{pyemma} (http://emma-project.org) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models; the package is written in Python (\url{http://python.org}) with heavy usage of numpy~\cite{numpy} and compatible to scikit-learn~\cite{sklearn} package.

\subsection{Scope}

In this tutorial, we assume familiarity with the theory behind the MSM approach and focus solely on usage of PyEMMA. In the first four sessions/notebooks, we introduce the basic components of the PyEMMA analysis pipeline, i.e., data I/O and featurization, dimension reduction and discretization; MSM estimation and validation, and MSM coase graining and analysis. In three further sessions/notebooks we introduce special topics, i.e., use of hidden MSMs, working with expectations and observables, and finding good features using the VAMP score approach. In the eigths and last session/notebook, we show common problems with data and how PyEMMA reacts to these specific situations.

\section{Prerequisites}

\subsection{Background knowledge}

MSM introduction~\cite{msm-brooke}, MSM theory~\cite{msm-jhp,msm-book}, TICA~\cite{tica,tica3,tica2}.

\subsection{Software/system requirements}

\section{Content and links}

This tutorial guides you through the elementary pipeline of buildiung, estimating, validating, and analysing MSMs (notebook 00). The individual steps are explained on simple systems in specialised notebooks (01-08).

Here we show briefly the form a standard MSM pipeline:

data-i/o + featurisation (number of features or system parameters)

dimension reduction and discretization (features plot, free energy or density fot tics[0,1]?)

estimation and validation (its, ck)

pcca + flux

Then we discuss common problems with realistic dataset and show how to spot them.

All notebooks are available on \githubrepository where you also might find additional materials and instalation instructions.

\section{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
For a more detailed description of author contributions, see the GitHub issue tracking and changelog at \githubrepository.

\section{Other Contributions}
%%%%%%%%%%%%%%%
% You should include all people who have filed issues that were
% accepted into the paper, or that upon discussion altered what was in the paper.
% Multiple significant contributions might mean that the contributor
% should be moved to authorship at the discretion of the a
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io for
% more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%
For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.

\section{Potentially Conflicting Interests}
%%%%%%%
%Declare any potentially competing interests, financial or otherwise
%%%%%%%

\section{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%

\bibliography{literature}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% APPENDICES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\appendix

\end{document}

### 2018-07-05 conftest.py
[ci] store output of notebooks as artifacts for later inspection

@@ -1,4 +1,8 @@
-import pytest, os, hashlib
+import pytest
+import os
+
+all_notebooks = None
+
 
 def pytest_collection_modifyitems(session, config, items):
     for i in items:
@@ -16,6 +20,11 @@ def pytest_collection_modifyitems(session, config, items):
     by_parents = defaultdict(list)
     for index, item in enumerate(items):
         by_parents[item.parent].append(item)
+
+    # store notebooks for later access
+    global all_notebooks
+    all_notebooks = [(x.name, x.nb) for x in by_parents.keys()]
+
     for i, p in enumerate(by_parents.keys()):
         if i % circle_node_total != circle_node_index:
             deselected.extend(by_parents[p])
@@ -44,3 +53,24 @@ def pytest_report_header(config):
     circle_node_total, circle_node_index = read_circleci_env_variables()
     return "CircleCI total nodes: {}, this node index: {}".format(circle_node_total, circle_node_index)
 
+
+def pytest_sessionfinish(session, exitstatus):
+    """ we store all notebooks in variable 'all_notebooks' to a given path and convert them to html """
+    import nbformat as nbf
+    import tempfile
+    out_dir = os.getenv('NBVAL_OUTPUT', tempfile.mkdtemp(prefix='pyemma_tut_test_output'))
+    print('write html output to', os.path.abspath(out_dir))
+    out_files = []
+    assert all_notebooks is not None
+    for name, nb in all_notebooks:
+        out_file = os.path.join(out_dir, os.path.basename(name))
+        with open(out_file, 'x') as fh:
+            nbf.write(nb, fh)
+        out_files.append(out_file)
+
+    import subprocess
+    subprocess.check_output(['jupyter', 'nbconvert', '--to=html', ' '.join(out_files)])
+
+    # delete source output notebooks
+    #for f in out_files:
+    #    os.unlink(f)

### 2018-07-05 conftest.py
[conftest] only upload executed notebooks (needed to check for deselections)

@@ -1,7 +1,7 @@
 import pytest
 import os
 
-all_notebooks = None
+executed_notebooks = None
 
 
 def pytest_collection_modifyitems(session, config, items):
@@ -10,27 +10,31 @@ def pytest_collection_modifyitems(session, config, items):
         excercise_2_cell = m.get('solution2_first', False)
         skip = m.get('skip', False)
         if excercise_2_cell or skip:
-           i.add_marker(pytest.mark.skip('solution stub'))
+            i.add_marker(pytest.mark.skip('solution stub'))
 
     circle_node_total, circle_node_index = read_circleci_env_variables()
-    deselected = []
-    # round robbin:
-    # first group by parent, then skip by ci node index
+    # store notebooks for later access
+    global executed_notebooks
     from collections import defaultdict
     by_parents = defaultdict(list)
     for index, item in enumerate(items):
         by_parents[item.parent].append(item)
 
-    # store notebooks for later access
-    global all_notebooks
-    all_notebooks = [(x.name, x.nb) for x in by_parents.keys()]
+    if circle_node_total == 1:
+        executed_notebooks = by_parents.keys()
+    else:
+        deselected = []
+        # round robbin: by notebook file and ci node index
+        for i, p in enumerate(by_parents.keys()):
+            if i % circle_node_total != circle_node_index:
+                deselected.extend(by_parents[p])
+        for d in deselected:
+            items.remove(d)
 
-    for i, p in enumerate(by_parents.keys()):
-        if i % circle_node_total != circle_node_index:
-            deselected.extend(by_parents[p])
-    for d in deselected:
-        items.remove(d)
-    config.hook.pytest_deselected(items=deselected)
+
+        executed_notebooks = [(nb.name, nb.nb) for nb in
+                              set(x.parent for x in set(items) - set(deselected))]
+        config.hook.pytest_deselected(items=deselected)
 
 
 class CircleCIError(Exception):
@@ -55,21 +59,23 @@ def pytest_report_header(config):
 
 
 def pytest_sessionfinish(session, exitstatus):
-    """ we store all notebooks in variable 'all_notebooks' to a given path and convert them to html """
+    """ we store all notebooks in variable 'executed_notebooks' to a given path and convert them to html """
     import nbformat as nbf
     import tempfile
     out_dir = os.getenv('NBVAL_OUTPUT', tempfile.mkdtemp(prefix='pyemma_tut_test_output'))
[…]

### 2018-07-05 manuscript.tex
added software section, fixed typos

@@
 
-\newcommand{\versionnumber}{0.1}
+\newcommand{\versionnumber}{0.1} % TODO: can we extract this number from a git tag?
 \newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}
@@
 \author[1\authfn{1}*]{AUTH0}
-\author[1\authfn{1}*]{AUTH1}
+\author[1\authfn{1}*]{Martin K. Scherer}
 \author[1*]{AUTH2}
@@
 \corr{auth0@email}{A0}
-\corr{auth1@email}{A1}
+\corr{m.scherer@fu-berlin.de}{A1}
 \corr{auth2@email}{A2}
@@
 
-\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
+\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, 
+please visit the GitHub repository and participate via the issue tracker.}
 
@@
 \begin{abstract}
-This tutorial provides an introduction to the PyEMMA project.
-Using Jupyter notebooks and accompanying screencasts, we will guide you through the basic functionality as well as the more common advanced mechanisms.
+This tutorial provides an introduction to the PyEMMA project. Using Jupyter notebooks and accompanying screencasts, 
+we will guide you through the basic functionality as well as the more common advanced mechanisms. 
 Short exercises to self check your learning progress and a notebook on troubleshooting complete this basic introduction.
@@
 
-PyEMMA~\cite{pyemma} (http://emma-project.org) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models; the package is written in Python (\url{http://python.org}) with heavy usage of numpy~\cite{numpy} and compatible to scikit-learn~\cite{sklearn} package.
+PyEMMA~\cite{pyemma} (http://emma-project.org) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models;
+the package is written in Python (\url{http://python.org}) with heavy usage of NumPy~\cite{numpy} and compatible to scikit-learn~\cite{sklearn} package.
 
@@
 In this tutorial, we assume familiarity with the theory behind the MSM approach and focus solely on usage of PyEMMA. 
-In the first four sessions/notebooks, we introduce the basic components of the PyEMMA analysis pipeline, i.e., data I/O and featurization, dimension reduction and discretization; MSM estimation and validation, and MSM coase graining and analysis.
-In three further sessions/notebooks we introduce special topics, i.e., use of hidden MSMs, working with expectations and observables, and finding good features using the VAMP score approach.
+In the first four sessions/notebooks, we introduce the basic components of the PyEMMA analysis pipeline, i.e., data I/O and featurization, dimension reduction and discretization; 
+MSM estimation and validation, and MSM coarse graining and analysis. In three further sessions/notebooks we introduce special topics, i.e., 
+use of hidden MSMs, working with expectations and observables, and finding good features using the VAMP score approach.
 In the eigths and last session/notebook, we show common problems with data and how PyEMMA reacts to these specific situations.
@@
 
+We utilize Jupyter notebooks to show code examples along with figures and interactive widgets to display molecules.
+The user can install all needed packages with only one command, if he or she uses the conda command provided by the Anaconda Python stack.
+We rely on Anaconda, because it resolves and installs dependencies and provides pre-compiled versions of hard to build packages like SciPy.
+The installation will contain a launcher command to start the Jupyter notebook server
+as well as the notebook files. The data for the demonstrated test systems is downloaded upon the first usage and is cached for future invocations of the tutorial.
+The underlying software stack for running the tutorial importantly consists out of:
+\begin{itemize}
+ \item NumPy - number crunching
+ \item SciPy - linear algebra, sparse computations
+ \item Matplotlib - Visualization
+ \item \textbf{PyEMMA} - MSM/HMM estimation, analysis, validation and visualization
+ \item mdtraj - reading MD simulation data, calculate observables from MD
+\end{itemize}
+
+The software is currently supported for Python versions 3.5 and 3.6 on the operating systems Linux, OSX and Windows.
+
+If one does not want to use Anaconda, he or she can also perform a manual installation via the pip installer or use the Binder to look and execute the tutorials online in any browser.
+
 \section{Content and links}
 
-This tutorial guides you through the elementary pipeline of buildiung, estimating, validating, and analysing MSMs (notebook 00).
-The individual steps are explained on simple systems in specialised notebooks (01-08).
+This tutorial guides you through the elementary pipeline of building, estimating, validating, and analyzing MSMs (notebook 00). 
+The individual steps are explained on simple systems in specialized notebooks (01-08).
 
@@
 
-data-i/o + featurisation (number of features or system parameters)
+data-i/o + featurization (number of features or system parameters)
 
-dimension reduction and discretization (features plot, free energy or density fot tics[0,1]?)
+dimension reduction and discretization (features plot, free energy or density for tics[0,1]?)
 
@@
 
-All notebooks are available on \githubrepository where you also might find additional materials and instalation instructions.
+All notebooks are available on \githubrepository where you also might find additional materials and installation instructions.
 

### 2018-07-06 conftest.py
[conftest] fix conversion

@@ -21,7 +21,7 @@ def pytest_collection_modifyitems(session, config, items):
         by_parents[item.parent].append(item)
 
     if circle_node_total == 1:
-        executed_notebooks = by_parents.keys()
+        executed_notebooks = [(nb.name, nb.nb) for nb in  by_parents.keys()]
     else:
         deselected = []
         # round robbin: by notebook file and ci node index
@@ -73,7 +73,7 @@ def pytest_sessionfinish(session, exitstatus):
         out_files.append(out_file)
 
     import subprocess
-    cmd = ['jupyter', 'nbconvert', '--to=html', ' '.join(out_files)]
+    cmd = ['jupyter', 'nbconvert', '--to=html'] + out_files
     print('converting via cmd:', cmd)
     subprocess.check_output(cmd)
 

### 2018-07-06 manuscript.tex
adding first two figures and draft pdf

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Learning PyEMMA : v\versionnumber}
+\title{First steps with PyEMMA : v\versionnumber}
 
-\author[1\authfn{1}*]{AUTH0}
+\author[1\authfn{1}*]{Christoph Wehmeyer}
 \author[1\authfn{1}*]{Martin K. Scherer}
-\author[1*]{AUTH2}
-\affil[1]{AFFILIATION}
+\author[1*]{Tim Hempel}
+\author[1*]{Simon Olsson}
+\author[1*]{Frank Noé}
+\affil[1]{Department of Mathematics and Computer Science, Freie Universität Berlin, Arnimallee 6, 14195 Berlin, Germany}
 
-\corr{auth0@email}{A0}
-\corr{m.scherer@fu-berlin.de}{A1}
-\corr{auth2@email}{A2}
+\corr{christoph.wehmeyer@fu-berlin.de}{CW}
+\corr{m.scherer@fu-berlin.de}{MS}
+\corr{tim.hempel@fu-berlin.de}{TH}
+\corr{simon.olsson@fu-berlin.de}{SO}
+\corr{frank.noe@fu-berlin.de}{FN}
 
@@
 
-\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, 
+\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository{}; to provide feedback, suggestions, or help improve it, 
 please visit the GitHub repository and participate via the issue tracker.}
@@
 \begin{abstract}
-This tutorial provides an introduction to the PyEMMA project. Using Jupyter notebooks and accompanying screencasts, 
-we will guide you through the basic functionality as well as the more common advanced mechanisms. 
+This tutorial provides an introduction to the PyEMMA project.
+Using Jupyter notebooks, we will guide you through the basic functionality as well as the more common advanced mechanisms.
 Short exercises to self check your learning progress and a notebook on troubleshooting complete this basic introduction.
@@
 
-PyEMMA~\cite{pyemma} (http://emma-project.org) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models;
-the package is written in Python (\url{http://python.org}) with heavy usage of NumPy~\cite{numpy} and compatible to scikit-learn~\cite{sklearn} package.
+PyEMMA~\cite{pyemma} (http://emma-project.org) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models; the package is written in Python (\url{http://python.org}) with heavy usage of NumPy~\cite{numpy} and compatible to scikit-learn~\cite{sklearn} package.
 
@@
 In this tutorial, we assume familiarity with the theory behind the MSM approach and focus solely on usage of PyEMMA.
-In the first four sessions/notebooks, we introduce the basic components of the PyEMMA analysis pipeline, i.e., data I/O and featurization, dimension reduction and discretization; 
-MSM estimation and validation, and MSM coarse graining and analysis. In three further sessions/notebooks we introduce special topics, i.e., 
-use of hidden MSMs, working with expectations and observables, and finding good features using the VAMP score approach.
+In the first four sessions/notebooks, we introduce the basic components of the PyEMMA analysis pipeline, i.e., data I/O and featurization, dimension reduction and discretization; MSM estimation and validation, and MSM coarse graining and analysis.
+In three further sessions/notebooks we introduce special topics, i.e., use of hidden MSMs, working with expectations and observables, and finding good features using the VAMP score approach.
 In the eigths and last session/notebook, we show common problems with data and how PyEMMA reacts to these specific situations.
@@
 We rely on Anaconda, because it resolves and installs dependencies and provides pre-compiled versions of hard to build packages like SciPy.
-The installation will contain a launcher command to start the Jupyter notebook server
-as well as the notebook files. The data for the demonstrated test systems is downloaded upon the first usage and is cached for future invocations of the tutorial.
+The installation will contain a launcher command to start the Jupyter notebook server as well as the notebook files.
+The data for the demonstrated test systems is downloaded upon the first usage and is cached for future invocations of the tutorial.
 The underlying software stack for running the tutorial importantly consists out of:
@@
 
+\begin{figure}
+\includegraphics{figure_1}
+\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The VAMP2 score indicates which of the tested featurization contains the most kinetic variance.
+(b) The first two independent components of a TICA projection at lag time $\tau=0.5$ ns show rare transition events.
+(c) Projection of the sample distribution and 200 $k$-means cluster centers onto the first two independent TICA components.
+(d) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM; the shaded areas indicate $95\%$ confidence intervals.}
+\label{fig:io-to-its}
+\end{figure}
+
+Figure~\ref{fig:io-to-its}
+
 data-i/o + featurization (number of features or system parameters)
@@
 
+\begin{figure}
+\includegraphics{figure_2}
+\caption{A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
+\label{fig:ck}
+\end{figure}
+
+Figure~\ref{fig:ck}
+
 pcca + flux
@@
 
-All notebooks are available on \githubrepository where you also might find additional materials and installation instructions.
+All notebooks are available on \githubrepository{} where you also might find additional materials and installation instructions.
 
@@
 %%%%%%%%%%%%%%%%
-For a more detailed description of author contributions, see the GitHub issue tracking and changelog at \githubrepository.
+For a more detailed description of author contributions, see the GitHub issue tracking and changelog at \githubrepository{}.
 
@@
 %%%%%%%%%%%%%%%
-For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository.
+For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository{}.
 

### 2018-07-09 conftest.py
group notebook dependencies.

@@ -1,10 +1,21 @@
+from glob import glob
+
 import pytest
 import os
 
 executed_notebooks = None
 
-from collections import defaultdict
+notebook_groups = [
+    (glob('notebooks/*msm-estimation*.ipynb')[0],
+     glob('notebooks/*msm-analysis*.ipynb')[0],
+     glob('notebooks/*pcca-tpt*.ipynb')[0],
+     ),
+]
 
+assert notebook_groups
+
+
+from collections import defaultdict
 timings = defaultdict(int)
 
 def pytest_runtest_logreport(report):
@@ -38,6 +49,18 @@ def pytest_collection_modifyitems(session, config, items):
     if circle_node_total == 1:
         executed_notebooks = [(nb.name, nb.nb) for nb in  by_parents.keys()]
     else:
+        # merge grouped parents
+        for n in notebook_groups:
+            items_to_group = []
+            keys_to_merge = []
+            for p in by_parents:
+                if p.name in n:
+                    items_to_group.extend(by_parents[p])
+                    keys_to_merge.append(p)
+            for k in keys_to_merge:
+                del by_parents[k]
+            by_parents[tuple(keys_to_merge)] = items_to_group
+
         deselected = []
         # round robbin: by notebook file and ci node index
         for i, p in enumerate(by_parents.keys()):
@@ -45,15 +68,9 @@ def pytest_collection_modifyitems(session, config, items):
                 deselected.extend(by_parents[p])
         for d in deselected:
             items.remove(d)
-
-
         executed_notebooks = [(nb.name, nb.nb) for nb in
                               set(x.parent for x in set(items) - set(deselected))]
         config.hook.pytest_deselected(items=deselected)
-
-
-class CircleCIError(Exception):
-    """Raised for problems running the CirleCI py.test plugin"""
 
 
 def read_circleci_env_variables():
@@ -62,7 +79,7 @@ def read_circleci_env_variables():
     circle_node_index = int(os.environ.get("CIRCLE_NODE_INDEX", "0").strip() or "0")
 
     if circle_node_index >= circle_node_total:
-        raise CircleCIError("CIRCLE_NODE_INDEX={} >= CIRCLE_NODE_TOTAL={}, should be less".format(circle_node_index, circle_node_total))
+        raise RuntimeError("CIRCLE_NODE_INDEX={} >= CIRCLE_NODE_TOTAL={}, should be less".format(circle_node_index, circle_node_total))
 
     return circle_node_total, circle_node_index
 

### 2018-07-09 conftest.py
[conftest] report notebook execution times

@@ -3,6 +3,21 @@ import os
 
 executed_notebooks = None
 
+from collections import defaultdict
+
+timings = defaultdict(int)
+
+def pytest_runtest_logreport(report):
+    if report.when == "call":
+        key = report.location[0]
+        timings[key] += report.duration
+
+def pytest_terminal_summary(terminalreporter, exitstatus):
+    from operator import itemgetter
+    s = sorted(timings.items(), key=lambda x: x[1])
+    for nb, total in s:
+        terminalreporter.write_line('%s took %.1f seconds' % (nb, total))
+
 
 def pytest_collection_modifyitems(session, config, items):
     for i in items:

### 2018-07-09 conftest.py
[conftest] do not use glob for pattern matching

@@ -6,15 +6,13 @@ import os
 executed_notebooks = None
 
 notebook_groups = [
-    (glob('notebooks/*msm-estimation*.ipynb')[0],
-     glob('notebooks/*msm-analysis*.ipynb')[0],
-     glob('notebooks/*pcca-tpt*.ipynb')[0],
+    (('msm-estimation'),
+     ('msm-analysis'),
+     ('pcca-tpt'),
      ),
 ]
 
-assert notebook_groups
-
-
+### execution timing ##########################################################
 from collections import defaultdict
 timings = defaultdict(int)
 
@@ -29,7 +27,10 @@ def pytest_terminal_summary(terminalreporter, exitstatus):
     for nb, total in s:
         terminalreporter.write_line('%s took %.1f seconds' % (nb, total))
 
+###############################################################################
 
+
+#### Circle CI parallel execution #############################################
 def pytest_collection_modifyitems(session, config, items):
     for i in items:
         m = i.cell['metadata']
@@ -88,6 +89,8 @@ def pytest_report_header(config):
     """Add CircleCI information to report"""
     circle_node_total, circle_node_index = read_circleci_env_variables()
     return "CircleCI total nodes: {}, this node index: {}".format(circle_node_total, circle_node_index)
+
+###############################################################################
 
 
 def pytest_sessionfinish(session, exitstatus):

### 2018-07-09 conftest.py
intermediate

@@ -42,22 +42,22 @@ def pytest_collection_modifyitems(session, config, items):
     circle_node_total, circle_node_index = read_circleci_env_variables()
     # store notebooks for later access
     global executed_notebooks
-    from collections import defaultdict
     by_parents = defaultdict(list)
     for index, item in enumerate(items):
         by_parents[item.parent].append(item)
 
     if circle_node_total == 1:
-        executed_notebooks = [(nb.name, nb.nb) for nb in  by_parents.keys()]
+        executed_notebooks = [(nb.name, nb.nb) for nb in by_parents.keys()]
     else:
         # merge grouped parents
         for n in notebook_groups:
             items_to_group = []
             keys_to_merge = []
             for p in by_parents:
-                if p.name in n:
-                    items_to_group.extend(by_parents[p])
-                    keys_to_merge.append(p)
+                for nb in n:
+                    if nb in p.name:
+                       items_to_group.extend(by_parents[p])
+                       keys_to_merge.append(p)
             for k in keys_to_merge:
                 del by_parents[k]
             by_parents[tuple(keys_to_merge)] = items_to_group
@@ -71,6 +71,7 @@ def pytest_collection_modifyitems(session, config, items):
             items.remove(d)
         executed_notebooks = [(nb.name, nb.nb) for nb in
                               set(x.parent for x in set(items) - set(deselected))]
+	print('will execute:', executed_notebooks)
         config.hook.pytest_deselected(items=deselected)
 
 

### 2018-07-09 conftest.py
pep8

@@ -16,10 +16,12 @@ notebook_groups = [
 from collections import defaultdict
 timings = defaultdict(int)
 
+
 def pytest_runtest_logreport(report):
     if report.when == "call":
         key = report.location[0]
         timings[key] += report.duration
+
 
 def pytest_terminal_summary(terminalreporter, exitstatus):
     from operator import itemgetter
@@ -56,8 +58,8 @@ def pytest_collection_modifyitems(session, config, items):
             for p in by_parents:
                 for nb in n:
                     if nb in p.name:
-                       items_to_group.extend(by_parents[p])
-                       keys_to_merge.append(p)
+                        items_to_group.extend(by_parents[p])
+                        keys_to_merge.append(p)
             for k in keys_to_merge:
                 del by_parents[k]
             by_parents[tuple(keys_to_merge)] = items_to_group
@@ -71,17 +73,20 @@ def pytest_collection_modifyitems(session, config, items):
             items.remove(d)
         executed_notebooks = [(nb.name, nb.nb) for nb in
                               set(x.parent for x in set(items) - set(deselected))]
-	print('will execute:', executed_notebooks)
+        print('will execute:', executed_notebooks)
         config.hook.pytest_deselected(items=deselected)
 
 
 def read_circleci_env_variables():
     """Read and convert CIRCLE_* environment variables"""
-    circle_node_total = int(os.environ.get("CIRCLE_NODE_TOTAL", "1").strip() or "1")
-    circle_node_index = int(os.environ.get("CIRCLE_NODE_INDEX", "0").strip() or "0")
+    circle_node_total = int(os.environ.get(
+        "CIRCLE_NODE_TOTAL", "1").strip() or "1")
+    circle_node_index = int(os.environ.get(
+        "CIRCLE_NODE_INDEX", "0").strip() or "0")
 
     if circle_node_index >= circle_node_total:
-        raise RuntimeError("CIRCLE_NODE_INDEX={} >= CIRCLE_NODE_TOTAL={}, should be less".format(circle_node_index, circle_node_total))
+        raise RuntimeError("CIRCLE_NODE_INDEX={} >= CIRCLE_NODE_TOTAL={}, should be less".format(
+            circle_node_index, circle_node_total))
 
     return circle_node_total, circle_node_index
 
@@ -98,7 +103,8 @@ def pytest_sessionfinish(session, exitstatus):
[…]

### 2018-07-09 manuscript.tex
minor text and figure updates

@@
 \includegraphics{figure_2}
-\caption{A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
+\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: a Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
 \label{fig:ck}
@@
 pcca + flux
+
+\begin{figure}
+\includegraphics{figure_3}
+\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The reweighted free energy surface in the projection of the first two independent components exhibits five minima which (b) PCCA++ identifies as four metastable states.
+(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state (0) and the two most probable states (2, 3), whereas (d) the commitor $1\to2$ indicates that the most probable state 3 acts as a transition state for transitions between states 1 and 2.}
+\label{fig:msm-analysis}
+\end{figure}
+
+Figure~\ref{fig:msm-analysis}
 
@@
 %%%%%%%%%%%%%%%%
-For a more detailed description of author contributions, see the GitHub issue tracking and changelog at \githubrepository{}.
+For a more detailed description of author contributions, see the GitHub issue tracking and changelog at\\\githubrepository{}.
 

### 2018-07-10 conftest.py
minor changes

@@ -6,16 +6,15 @@ import os
 executed_notebooks = None
 
 notebook_groups = [
-    (('msm-estimation'),
-     ('msm-analysis'),
-     ('pcca-tpt'),
+    ('msm-estimation',
+     'msm-analysis',
+     'pcca-tpt',
      ),
 ]
 
 ### execution timing ##########################################################
 from collections import defaultdict
 timings = defaultdict(int)
-
 
 def pytest_runtest_logreport(report):
     if report.when == "call":
@@ -24,7 +23,7 @@ def pytest_runtest_logreport(report):
 
 
 def pytest_terminal_summary(terminalreporter, exitstatus):
-    from operator import itemgetter
+    terminalreporter.section('Notebook timings')
     s = sorted(timings.items(), key=lambda x: x[1])
     for nb, total in s:
         terminalreporter.write_line('%s took %.1f seconds' % (nb, total))
@@ -73,7 +72,7 @@ def pytest_collection_modifyitems(session, config, items):
             items.remove(d)
         executed_notebooks = [(nb.name, nb.nb) for nb in
                               set(x.parent for x in set(items) - set(deselected))]
-        print('will execute:', executed_notebooks)
+        print('Notebooks to execute:', [x[0] for x in executed_notebooks])
         config.hook.pytest_deselected(items=deselected)
 
 
@@ -115,7 +114,9 @@ def pytest_sessionfinish(session, exitstatus):
         out_files.append(out_file)
 
     import subprocess
-    cmd = ['jupyter', 'nbconvert', '--to=html'] + out_files
+    import sys
+    
+    cmd = [sys.executable, '-m', 'jupyter', 'nbconvert', '--to=html'] + out_files
     print('converting via cmd:', cmd)
     subprocess.check_output(cmd)
 

### 2018-07-10 setup.py

@@ -7,9 +7,13 @@ from setuptools import setup
 def copy_notebooks():
     import shutil
     dest = os.path.join('pyemma_tutorials', 'notebooks')
+    ignored_patterns = ('data', '*.xtc', '*.pyemma', '*.pdb', '*.dcd')
+    import fnmatch
+    def ignore(src, fn):
+        return any(fnmatch.fnmatch(fn, p) for p in ignored_patterns)
     try:
         shutil.rmtree(dest, ignore_errors=True)
-        shutil.copytree('notebooks', dest)
+        shutil.copytree('notebooks', dest, ignore=ignore)
         print('moved notebooks into pkg')
     except OSError:
         pass

### 2018-07-10 setup.py
fix

@@ -7,13 +7,9 @@ from setuptools import setup
 def copy_notebooks():
     import shutil
     dest = os.path.join('pyemma_tutorials', 'notebooks')
-    ignored_patterns = ('data', '*.xtc', '*.pyemma', '*.pdb', '*.dcd')
-    import fnmatch
-    def ignore(src, fn):
-        return any(fnmatch.fnmatch(fn, p) for p in ignored_patterns)
     try:
         shutil.rmtree(dest, ignore_errors=True)
-        shutil.copytree('notebooks', dest, ignore=ignore)
+        shutil.copytree('notebooks', dest)
         print('moved notebooks into pkg')
     except OSError:
         pass

### 2018-07-10 conftest.py
[conftest] fix output storage

@@ -3,7 +3,7 @@ from glob import glob
 import pytest
 import os
 
-executed_notebooks = None
+executed_notebooks = set()
 
 notebook_groups = [
     ('msm-estimation',
@@ -15,6 +15,7 @@ notebook_groups = [
 ### execution timing ##########################################################
 from collections import defaultdict
 timings = defaultdict(int)
+
 
 def pytest_runtest_logreport(report):
     if report.when == "call":
@@ -41,15 +42,11 @@ def pytest_collection_modifyitems(session, config, items):
             i.add_marker(pytest.mark.skip('solution stub'))
 
     circle_node_total, circle_node_index = read_circleci_env_variables()
-    # store notebooks for later access
-    global executed_notebooks
-    by_parents = defaultdict(list)
-    for index, item in enumerate(items):
-        by_parents[item.parent].append(item)
+    if circle_node_total > 1:
+        by_parents = defaultdict(list)
+        for index, item in enumerate(items):
+            by_parents[item.parent].append(item)
 
-    if circle_node_total == 1:
-        executed_notebooks = [(nb.name, nb.nb) for nb in by_parents.keys()]
-    else:
         # merge grouped parents
         for n in notebook_groups:
             items_to_group = []
@@ -70,7 +67,7 @@ def pytest_collection_modifyitems(session, config, items):
                 deselected.extend(by_parents[p])
         for d in deselected:
             items.remove(d)
-        executed_notebooks = [(nb.name, nb.nb) for nb in
+        executed_notebooks = [nb.name for nb in
                               set(x.parent for x in set(items) - set(deselected))]
         print('Notebooks to execute:', [x[0] for x in executed_notebooks])
         config.hook.pytest_deselected(items=deselected)
@@ -98,6 +95,10 @@ def pytest_report_header(config):
 ###############################################################################
 
 
+def pytest_runtest_protocol(item, nextitem):
+    executed_notebooks.add(item.parent)
+
+
 def pytest_sessionfinish(session, exitstatus):
     """ we store all notebooks in variable 'executed_notebooks' to a given path and convert them to html """
     import nbformat as nbf
@@ -106,17 +107,18 @@ def pytest_sessionfinish(session, exitstatus):
         prefix='pyemma_tut_test_output'))
     print('write html output to', os.path.abspath(out_dir))
     out_files = []
-    assert executed_notebooks is not None
[…]

### 2018-07-11 manuscript.tex
merging figure_2 into figure_1

@@
 \includegraphics{figure_1}
-\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The VAMP2 score indicates which of the tested featurization contains the most kinetic variance.
-(b) The first two independent components of a TICA projection at lag time $\tau=0.5$ ns show rare transition events.
-(c) Projection of the sample distribution and 200 $k$-means cluster centers onto the first two independent TICA components.
-(d) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM; the shaded areas indicate $95\%$ confidence intervals.}
+\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The studied pentapeptide.
+(b) The VAMP2 score indicates which of the tested featurization contains the most kinetic variance.
+(c) The first two independent components of a TICA projection at lag time $\tau=0.5$ ns show rare transition events.
+(d) Projection of the sample distribution and 200 $k$-means cluster centers onto the first two independent TICA components.
+(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM; the shaded areas indicate $95\%$ confidence intervals.
+(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
 \label{fig:io-to-its}
@@
 
-\begin{figure}
-\includegraphics{figure_2}
-\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: a Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
-\label{fig:ck}
-\end{figure}
-
-Figure~\ref{fig:ck}
-
 pcca + flux
@@
 \begin{figure}
-\includegraphics{figure_3}
+\includegraphics{figure_2}
 \caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The reweighted free energy surface in the projection of the first two independent components exhibits five minima which (b) PCCA++ identifies as four metastable states.

### 2018-07-11 conftest.py
properly map test outputs to input notebook cells

@@ -1,9 +1,6 @@
-from glob import glob
-
 import pytest
 import os
 
-executed_notebooks = set()
 
 notebook_groups = [
     ('msm-estimation',
@@ -30,16 +27,18 @@ def pytest_terminal_summary(terminalreporter, exitstatus):
         terminalreporter.write_line('%s took %.1f seconds' % (nb, total))
 
 ###############################################################################
-
+def cell_skipped(cell_metadata):
+    excercise_2_cell = cell_metadata.get('solution2_first', False)
+    skip = cell_metadata.get('skip', False)
+    if excercise_2_cell or skip:
+        return True
+    return False
 
 #### Circle CI parallel execution #############################################
 def pytest_collection_modifyitems(session, config, items):
     for i in items:
-        m = i.cell['metadata']
-        excercise_2_cell = m.get('solution2_first', False)
-        skip = m.get('skip', False)
-        if excercise_2_cell or skip:
-            i.add_marker(pytest.mark.skip('solution stub'))
+        if cell_skipped(i.cell['metadata']):
+            i.add_marker(pytest.mark.skip('solution stub or metadata["skip"]=True'))
 
     circle_node_total, circle_node_index = read_circleci_env_variables()
     if circle_node_total > 1:
@@ -69,7 +68,7 @@ def pytest_collection_modifyitems(session, config, items):
             items.remove(d)
         executed_notebooks = [nb.name for nb in
                               set(x.parent for x in set(items) - set(deselected))]
-        print('Notebooks to execute:', [x[0] for x in executed_notebooks])
+        print('Notebooks to execute:', executed_notebooks)
         config.hook.pytest_deselected(items=deselected)
 
 
@@ -92,13 +91,12 @@ def pytest_report_header(config):
     circle_node_total, circle_node_index = read_circleci_env_variables()
     return "CircleCI total nodes: {}, this node index: {}".format(circle_node_total, circle_node_index)
 
+cells_per_notebook = defaultdict(list)
+def pytest_runtest_call(item):
+    cells_per_notebook[item.parent].append(item)
+
 ###############################################################################
-
-
-def pytest_runtest_protocol(item, nextitem):
-    executed_notebooks.add(item.parent)
-
-
+@pytest.hookimpl(hookwrapper=True)
 def pytest_sessionfinish(session, exitstatus):
[…]

### 2018-07-11 conftest.py
fix

@@ -91,12 +91,15 @@ def pytest_report_header(config):
     circle_node_total, circle_node_index = read_circleci_env_variables()
     return "CircleCI total nodes: {}, this node index: {}".format(circle_node_total, circle_node_index)
 
+###############################################################################
+
 cells_per_notebook = defaultdict(list)
+
+
 def pytest_runtest_call(item):
     cells_per_notebook[item.parent].append(item)
 
-###############################################################################
-@pytest.hookimpl(hookwrapper=True)
+
 def pytest_sessionfinish(session, exitstatus):
     """ we store all notebooks in variable 'executed_notebooks' to a given path and convert them to html """
     import nbformat as nbf
@@ -130,5 +133,3 @@ def pytest_sessionfinish(session, exitstatus):
     # delete source output notebooks
     for f in out_files:
         os.unlink(f)
-
-    return exitstatus

### 2018-07-12 conftest.py
disable progress bars for testing session

@@ -12,6 +12,14 @@ notebook_groups = [
      ),
 ]
 
+@pytest.fixture(scope='session')
+def no_progress_bars():
+    """ disables progress bars during testing """
+    import pyemma
+    pyemma.config.show_progress_bars = False
+    yield
+
+
 ### execution timing ##########################################################
 from collections import defaultdict
 timings = defaultdict(int)

### 2018-07-14 conftest.py
dbg

@@ -112,10 +112,9 @@ def pytest_sessionfinish(session, exitstatus):
     for ipynbfile in ipynbfiles:
         out_file = os.path.join(out_dir, os.path.basename(ipynbfile.name))
         # map output cells
-        cells_with_output = filter(lambda c: hasattr(c, 'outputs'), ipynbfile.nb.cells)
-        cells_with_output_not_skipped = filter(lambda c: not cell_skipped(c.metadata), cells_with_output)
-        for cell, ipynbcell in zip(cells_with_output_not_skipped, cells_per_notebook[ipynbfile]):
-            assert hasattr(cell, 'outputs')
+        cells_with_non_skipped_output = (c for c in ipynbfile.nb.cells if hasattr(c, 'outputs') and not cell_skipped(c.metadata))
+        for cell, ipynbcell in zip(cells_with_non_skipped_output, cells_per_notebook[ipynbfile]):
+            print(cell, ipynbcell)
             cell.outputs = ipynbcell.test_outputs
 
         with open(out_file, 'x') as fh:

### 2018-07-17 conftest.py
[conftest] ungroup notebook, which has no dependency

@@ -6,8 +6,7 @@ import os
 executed_notebooks = set()
 
 notebook_groups = [
-    ('msm-estimation',
-     'msm-analysis',
+     ('msm-analysis',
      'pcca-tpt',
      ),
 ]

### 2018-07-18 setup.py
ship static folder
Fixes #118

@@ -21,6 +21,7 @@ metadata=dict(
     cmdclass=versioneer.get_cmdclass(),
     packages=['pyemma_tutorials'],
     package_data={'pyemma_tutorials': ['notebooks/*',
+                                       'notebooks/static/*',
                                        'jupyter_notebook_config.py',
                                        'jupyter_notebook_config.json',
                                        ]},

### 2018-07-20 manuscript.tex
figure and literature update
draft of the content section

@@
 %%% PREAMBLE
-\documentclass[9pt,tutorial,lineno,onehalfspacing]{livecoms}
+\documentclass[9pt,tutorial]{livecoms}
 % Use the 'onehalfspacing' option for 1.5 line spacing
@@
 \corr{christoph.wehmeyer@fu-berlin.de}{CW}
-\corr{m.scherer@fu-berlin.de}{MS}
+\corr{m.scherer@fu-berlin.de}{MKS}
 \corr{tim.hempel@fu-berlin.de}{TH}
@@
 
-\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository{}; to provide feedback, suggestions, or help improve it, 
-please visit the GitHub repository and participate via the issue tracker.}
+\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository{}; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
 
@@
 
-We utilize Jupyter notebooks to show code examples along with figures and interactive widgets to display molecules.
+We utilize Jupyter~\cite{jupyter} notebooks to show code examples along with figures and interactive widgets to display molecules.
 The user can install all needed packages with only one command, if he or she uses the conda command provided by the Anaconda Python stack.
@@
 
-This tutorial guides you through the elementary pipeline of building, estimating, validating, and analyzing MSMs (notebook 00).
+% This tutorial guides you through the elementary pipeline of building, estimating, validating, and analyzing MSMs (notebook 00, \url{https://git.io/fNkjJ}).
 The individual steps are explained on simple systems in specialized notebooks (01-08).
 
-Here we show briefly the form a standard MSM pipeline:
+This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
+The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow.
+The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics from the workflow, an introduction to hidden Markov state models (HMMs), and guidelines how to deal with common problems during an MSM estimation.
+
+\subsection{The PyEMMA workflow}
+
+In short, the workflow consists of extracting molecular features from the raw data (01) which we transform into a suitable, low-dimensional subspace and discretize (02).
+Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajetcories and perform validation tests (03).
+We can then analize the stationary and kinetic properties of the MSM (04), apply transition path theory (TPT) to identify the pathways of conformation change (05), and compute expectation values for experimental observables (06).
+
+In the showcase notebook, we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns.~\cite{pyemma}
+
+\subsection{Estimation and validation}
 
@@
 (f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
-\label{fig:io-to-its}
+\label{fig:io-to-ck}
 \end{figure}
 
-Figure~\ref{fig:io-to-its}
+As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we start with a broad systematic analysis: we compute the (cross-validated) VAMP-2~\cite{vamp-preprint} score at lag time $0.5$ ns for three different featurizations and find that backbone torsions contain more kinetic variance than the backbone's heavy atoms' positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
 
-data-i/o + featurization (number of features or system parameters)
+A subsequent TICA~\cite{tica,kinetic-maps} transformation of the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace.
+The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima and a time series visualization of said components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
+We thus assume that our transformed feature selection describes one or more metastable processes.
 
-dimension reduction and discretization (features plot, free energy or density for tics[0,1]?)
+We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$.~\cite{swope-its} The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM's transition matrix via
+\begin{equation}
+\label{eq:its}
+t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}
+\end{equation}
+with $\tau$ being the lag time.
+In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme.~\cite{noe-tmat-sampling} In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
+We can further see a converged fourth process (cyan line) which, however, decays for lag times above $2.5$ ns and might not be well resolved in the final model.
 
-estimation and validation (its, ck)
+Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test which compares the right and the left side of the Chapman-Kolmogorov equation
+\begin{equation}
+\label{eq:ck}
+T(k \tau) = T^k(\tau)
+\end{equation}
+with $T$ being the MSM's transition matrix.
+PyEMMA automatically estimates a new MSM transition matrix at lag time $k \tau$ and propagates the original transition matrix by the $k^\textrm{th}$ power.
+As we have resolved three slow timescales in the ITS test, we assume to have four metastable states present in the data and the subsequent CK test shows that our model accurately indeed predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
 
-pcca + flux
+\subsection{Analyzing the MSM}
 
@@
 
-Figure~\ref{fig:msm-analysis}
+We can now directly extract several thermodynamic and kinetic properties from the estimated and validated model.
+An example of the former is the free energy surface in the projection of the first two TICA components (Fig.~\ref{fig:msm-analysis}a) reweighted based on the MSM's stationary distribution.
 
-Then we discuss common problems with realistic dataset and show how to spot them.
+A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=0,\dots,3$, for which we then approximate the absolute probabilities and free energies
+\[ \begin{array}{ccc}
+\textrm{macrostate } i & \pi_{\mathcal{S}_i} & G_i / \textrm{k}_\textrm{B} T \\
+\hline
+0 & 0.004 & 5.565 \\
+1 & 0.013 & 4.314 \\
+2 & 0.021 & 3.864 \\
+3 & 0.962 & 0.039
+\end{array}\]
+the latter via
+\begin{equation}
+\label{eq:fe}
+G_i = - \textrm{k}_\textrm{B} T \ln\left(\sum\limits_{j\in \mathcal{S}_i} \pi_j\right),
+\end{equation}
+where $\pi_j$ denotes the MSM's stationary weight of the $j^\textrm{th}$ microstate.
 
-All notebooks are available on \githubrepository{} where you also might find additional materials and installation instructions.
+To understand which slow process occur, we refer to the the MSM's (right) eigenvectors as they contain information about what configurational changes are happening on which timescales.
+The first right eigenvector corresponds to the stationary process and is constant $1$.
+The second right eigenvector, however, corresponds the the slowest process and its minimal and maximal components indicate the microstates between the process shifts probability.
+The relaxation timescale of this exchange process is exactly the implied timescale.
+In the usual projection, we identify the slowest process as a probability shift between macrostate $0$ and the rest of the system, with macrostates $2$ and $3$ in particular (Fig.~\ref{fig:msm-analysis}c).
+
+The mean first passage times (MFPTs) for the slowest process compute to
+\[ \begin{array}{crcr}
+\textrm{direction} & \textrm{mean / ns} && \textrm{std / ns} \\
+\hline
+\mathcal{S}_0 \to \mathcal{S}_{(1,2,3)} & 8.9 & \pm & 1.8 \\
+\mathcal{S}_{(1,2,3)} \to \mathcal{S}_0 & 2487.5 & \pm &  504.7
+\end{array}\]
+using the Bayesian MSM.
+
+As an example for TPT, we compute the flux between macrostates $1$ and $2$ (Fig.~\ref{fig:msm-analysis}d).
+The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
+Transition regions (macrostate $3$) can be identified by committor values $\approx \frac{1}{2}$.
+
+Next we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
+The SASA is a reasonable approximation for the instantaneous tryptophan flourescence signal and the auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.
+In our usual projection, we observe a high variance of the SASA in all macrostates except $0$ (Fig.~\ref{fig:msm-analysis}e), indicating that Trp-1 has only a low solvent exposure in this macrostate.
+The ACF itself yields only a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
+If, however, we prepare the system in macrostate $0$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
+
+Finally, we sample $100$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g).
+We observe that, in macrostates $0$ and $1$, the backbone appears to shield the Trp-1 (shown in red) from the solvent.
+In the more probable macrostates $2$ and $3$, the Trp-1 appears to be much more exposed.
+
+\subsection{Summary}
+
+We have summarized in this section how to conduct an MSM-based analysis of molecular dynamcis data using PyEMMA.
+For the full analysis, please refer to the first notebook (00).
+All notebooks as well as detailed installation instructions are available on \githubrepository{}.
 
@@
 %%%%%%%%%%%%%%%
+We are grateful to Brooke Husic for her thorough review of the notebooks in this tutorial as well as Camilla Ventura Santos and the entire computational molecular biology group for valuable discussion and feedback.
+
 For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository{}.
@@
 %%%%%%%
+TH acknowledges financial support by SFB/TRR 186. SO acknowledges a postdoctral fellowship from the Alexander von Humboldt Stiftung.
 

### 2018-07-20 manuscript.tex
draft of the scope section

@@
 
-PyEMMA~\cite{pyemma} (http://emma-project.org) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models; the package is written in Python (\url{http://python.org}) with heavy usage of NumPy~\cite{numpy} and compatible to scikit-learn~\cite{sklearn} package.
+PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models\cite{schuette-msm,singhal-msm-naming} (MSMs).
+The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.
 
@@
 
-In this tutorial, we assume familiarity with the theory behind the MSM approach and focus solely on usage of PyEMMA.
-In the first four sessions/notebooks, we introduce the basic components of the PyEMMA analysis pipeline, i.e., data I/O and featurization, dimension reduction and discretization; MSM estimation and validation, and MSM coarse graining and analysis.
-In three further sessions/notebooks we introduce special topics, i.e., use of hidden MSMs, working with expectations and observables, and finding good features using the VAMP score approach.
-In the eigths and last session/notebook, we show common problems with data and how PyEMMA reacts to these specific situations.
+In this tutorial, we assume the reader's familiarity with the basic theory behind the MSM approach (see Sec.~\ref{sec:background}) and focus on usage of PyEMMA.
+Nevertheless will we mention important theoretical concepts throughout the tutorial.
+
+The tutorial is divided into lessons on specific topics, each accompanied by a Jupyter~\cite{jupyter} notebook containing code, instructions, and exercises.
+The lessons start with a showcase of the PyEMMA workflow and follow up with in-depth lessons on specific topics.
 
@@
 
+In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial as well as the software requirements to work through the lessons.
+
 \subsection{Background knowledge}
+\label{sec:background}
 
-MSM introduction~\cite{msm-brooke}, MSM theory~\cite{msm-jhp,msm-book}, TICA~\cite{tica,tica3,tica2}.
+For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a comprehensive overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the mathematical theory.
+
+In addition to the literature on theory and application of MArkov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
 
@@
 We utilize Jupyter~\cite{jupyter} notebooks to show code examples along with figures and interactive widgets to display molecules.
-The user can install all needed packages with only one command, if he or she uses the conda command provided by the Anaconda Python stack.
-We rely on Anaconda, because it resolves and installs dependencies and provides pre-compiled versions of hard to build packages like SciPy.
+The user can install all needed packages with only one command if they use the \texttt{conda} command provided by the Anaconda Python stack (\url{https://anaconda.com}).
+We rely on Anaconda, because it resolves and installs dependencies, and provides pre-compiled versions of hard to build packages like SciPy.
+
 The installation will contain a launcher command to start the Jupyter notebook server as well as the notebook files.
-The data for the demonstrated test systems is downloaded upon the first usage and is cached for future invocations of the tutorial.
-The underlying software stack for running the tutorial importantly consists out of:
+The data for the demonstrated test systems is downloaded upon the first use and is cached for future invocations of the tutorial.
+
+The underlying software stack for running the tutorial importantly consists of:
 \begin{itemize}
- \item NumPy - number crunching
- \item SciPy - linear algebra, sparse computations
- \item Matplotlib - Visualization
- \item \textbf{PyEMMA} - MSM/HMM estimation, analysis, validation and visualization
- \item mdtraj - reading MD simulation data, calculate observables from MD
+\item NumPy -- number crunching~\cite{numpy}
+\item SciPy -- linear algebra, sparse computations~\cite{scipy}
+\item Matplotlib -- Visualization~\cite{matplotlib}
+\item \textbf{PyEMMA} -- MSM/HMM estimation, validation, analysis, and visualization~\cite{pyemma}
+\item mdtraj -- reading MD simulation data, calculate observables from MD~\cite{mdtraj}
 \end{itemize}
 
-The software is currently supported for Python versions 3.5 and 3.6 on the operating systems Linux, OSX and Windows.
+The software is currently supported for Python versions $3.5$ and $3.6$ on the operating systems Linux, OSX, and Windows.
 
-If one does not want to use Anaconda, he or she can also perform a manual installation via the pip installer or use the Binder to look and execute the tutorials online in any browser.
+If one does not want to use Anaconda, they can also perform a manual installation via the pip installer or use the Binder service to view and run the tutorials online in any browser.
 
@@
 
-% This tutorial guides you through the elementary pipeline of building, estimating, validating, and analyzing MSMs (notebook 00, \url{https://git.io/fNkjJ}).
-The individual steps are explained on simple systems in specialized notebooks (01-08).
-
 This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
 The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow.
-The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics from the workflow, an introduction to hidden Markov state models (HMMs), and guidelines how to deal with common problems during an MSM estimation.
+The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics from the workflow, an introduction to HMMs, and guidelines how to deal with common problems during an MSM estimation.
 
@@
 %%%%%%%%%%%%%%%
-We are grateful to Brooke Husic for her thorough review of the notebooks in this tutorial as well as Camilla Ventura Santos and the entire computational molecular biology group for valuable discussion and feedback.
+We are grateful to Brooke E.
+Husic for her thorough review of the notebooks in this tutorial as well as Camilla Ventura Santos and the entire computational molecular biology group for valuable discussion and feedback.
 

### 2018-07-20 manuscript.tex
minor changes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{First steps with PyEMMA : v\versionnumber}
+\title{PyEMMA: learning the ropes of Markov state modeling --- v\versionnumber}
 
@@
 \author[1\authfn{1}*]{Martin K. Scherer}
-\author[1*]{Tim Hempel}
+\author[1\authfn{1}*]{Tim Hempel}
 \author[1*]{Simon Olsson}
@@
 
-In addition to the literature on theory and application of MArkov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
+In addition to the literature on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
 

### 2018-07-20 manuscript.tex
completed manuscript draft

@@
 
-PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a framework for the analysis of molecular dynamics (MD) simulations using Markov state models\cite{schuette-msm,singhal-msm-naming} (MSMs).
+PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming} (MSMs).
 The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.
@@
 In this tutorial, we assume the reader's familiarity with the basic theory behind the MSM approach (see Sec.~\ref{sec:background}) and focus on usage of PyEMMA.
-Nevertheless will we mention important theoretical concepts throughout the tutorial.
+Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
 
@@
 
-In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial as well as the software requirements to work through the lessons.
+In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial. Then, we address the software requirements to work through the lessons.
 
@@
 
-In addition to the literature on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
+In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
 
@@
 We utilize Jupyter~\cite{jupyter} notebooks to show code examples along with figures and interactive widgets to display molecules.
-The user can install all needed packages with only one command if they use the \texttt{conda} command provided by the Anaconda Python stack (\url{https://anaconda.com}).
+The user can install all necessary packages in one step using the \texttt{conda} command provided by the Anaconda Python stack (\url{https://anaconda.com}).
 We rely on Anaconda, because it resolves and installs dependencies, and provides pre-compiled versions of hard to build packages like SciPy.
@@
 \begin{itemize}
-\item NumPy -- number crunching~\cite{numpy}
+\item NumPy -- vectorized operations on multi-dimensional arrays~\cite{numpy}
 \item SciPy -- linear algebra, sparse computations~\cite{scipy}
@@
 
-If one does not want to use Anaconda, they can also perform a manual installation via the pip installer or use the Binder service to view and run the tutorials online in any browser.
+Should the user prefer to not use Anaconda, a manual installation via the pip installer is possible. Alternatively, one can use the Binder service to view and run the tutorials online in any browser.
 
@@
 The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow.
-The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics from the workflow, an introduction to HMMs, and guidelines how to deal with common problems during an MSM estimation.
+The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics, an introduction to HMMs, and guidelines how to deal with common problems during an MSM estimation.
 
@@
 In short, the workflow consists of extracting molecular features from the raw data (01) which we transform into a suitable, low-dimensional subspace and discretize (02).
-Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajetcories and perform validation tests (03).
-We can then analize the stationary and kinetic properties of the MSM (04), apply transition path theory (TPT) to identify the pathways of conformation change (05), and compute expectation values for experimental observables (06).
+Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajectories and perform validation tests (03).
+We can then analize the stationary and kinetic properties of the MSM (04), find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05), and compute expectation values for experimental observables (06).
 
-In the showcase notebook, we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns.~\cite{pyemma}
+In the showcase notebook, we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns~\cite{pyemma}.
 
@@
 \includegraphics{figure_1}
-\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The studied pentapeptide.
-(b) The VAMP2 score indicates which of the tested featurization contains the most kinetic variance.
-(c) The first two independent components of a TICA projection at lag time $\tau=0.5$ ns show rare transition events.
-(d) Projection of the sample distribution and 200 $k$-means cluster centers onto the first two independent TICA components.
+\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation.
+(b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
+(c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and (d) the timeseries of the first two independent components show rare transition events.
 (e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM; the shaded areas indicate $95\%$ confidence intervals.
@@
 
-As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we start with a broad systematic analysis: we compute the (cross-validated) VAMP-2~\cite{vamp-preprint} score at lag time $0.5$ ns for three different featurizations and find that backbone torsions contain more kinetic variance than the backbone's heavy atoms' positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
+As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we start with a broad systematic analysis: we compute the (cross-validated) VAMP-2~\cite{vamp-preprint} score at lag time $0.5$ ns for three different featurizations and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
 
@@
 
-We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$.~\cite{swope-its} The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM's transition matrix via
+We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$~\cite{swope-its}.
+The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
 \begin{equation}
@@
 with $\tau$ being the lag time.
-In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme.~\cite{noe-tmat-sampling} In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
+In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{noe-tmat-sampling}.
+In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
 We can further see a converged fourth process (cyan line) which, however, decays for lag times above $2.5$ ns and might not be well resolved in the final model.
@@
 PyEMMA automatically estimates a new MSM transition matrix at lag time $k \tau$ and propagates the original transition matrix by the $k^\textrm{th}$ power.
-As we have resolved three slow timescales in the ITS test, we assume to have four metastable states present in the data and the subsequent CK test shows that our model accurately indeed predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
+As we have resolved three slow timescales in the ITS test, we assume to have four metastable states present in the data and the subsequent CK test shows that indeed our model predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
 
@@
 \includegraphics{figure_2}
-\caption{Exemplary analysis of a pentapeptide's backbone conformational dynamics: (a) The reweighted free energy surface in the projection of the first two independent components exhibits five minima which (b) PCCA++ identifies as four metastable states.
-(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state (0) and the two most probable states (2, 3), whereas (d) the commitor $1\to2$ indicates that the most probable state 3 acts as a transition state for transitions between states 1 and 2.}
+\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which (b) PCCA++ identifies as four metastable states.
+(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_0$) and the two most probable states ($\mathcal{S}_2$, $\mathcal{S}_3$), whereas (d) the commitor $\mathcal{S}_1\to\mathcal{S}_2$ indicates that the most probable state $\mathcal{S}_3$ acts as a transition state for transitions between states $\mathcal{S}_1$ and $\mathcal{S}_2$.
+(e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_0$ in which the Trp-1 shows the lowest exposure to the solvent.
+(f) The Trp-1 SASA auto-correlation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the non-equilibrium condition $\mathcal{S}_0$ (bottom).
+(g) Super-imposed cartoon representations of $100$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(0,1)}$ but less so in $\mathcal{S}_{(2,3)}$.}
 \label{fig:msm-analysis}
@@
 We can now directly extract several thermodynamic and kinetic properties from the estimated and validated model.
-An example of the former is the free energy surface in the projection of the first two TICA components (Fig.~\ref{fig:msm-analysis}a) reweighted based on the MSM's stationary distribution.
+An example of the former is the free energy surface in the projection onto the first two TICA components (Fig.~\ref{fig:msm-analysis}a) reweighted based on the MSM stationary distribution.
 
@@
 \[ \begin{array}{ccc}
-\textrm{macrostate } i & \pi_{\mathcal{S}_i} & G_i / \textrm{k}_\textrm{B} T \\
+\textrm{macrostate } \mathcal{S}_i & \pi_{\mathcal{S}_i} & G_{\mathcal{S}_i} / \textrm{k}_\textrm{B} T \\
 \hline
-0 & 0.004 & 5.565 \\
-1 & 0.013 & 4.314 \\
-2 & 0.021 & 3.864 \\
-3 & 0.962 & 0.039
+\mathcal{S}_0 & 0.004 & 5.565 \\
+\mathcal{S}_1 & 0.013 & 4.314 \\
+\mathcal{S}_2 & 0.021 & 3.864 \\
+\mathcal{S}_3 & 0.962 & 0.039
 \end{array}\]
-the latter via
+using the relation
 \begin{equation}
 \label{eq:fe}
-G_i = - \textrm{k}_\textrm{B} T \ln\left(\sum\limits_{j\in \mathcal{S}_i} \pi_j\right),
+G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln\left(\sum\limits_{j\in \mathcal{S}_i} \pi_j\right),
 \end{equation}
-where $\pi_j$ denotes the MSM's stationary weight of the $j^\textrm{th}$ microstate.
+where $\pi_j$ denotes the MSM stationary weight of the $j^\textrm{th}$ microstate.
 
-To understand which slow process occur, we refer to the the MSM's (right) eigenvectors as they contain information about what configurational changes are happening on which timescales.
+To understand which slow process occur, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening on which timescales.
 The first right eigenvector corresponds to the stationary process and is constant $1$.
-The second right eigenvector, however, corresponds the the slowest process and its minimal and maximal components indicate the microstates between the process shifts probability.
-The relaxation timescale of this exchange process is exactly the implied timescale.
-In the usual projection, we identify the slowest process as a probability shift between macrostate $0$ and the rest of the system, with macrostates $2$ and $3$ in particular (Fig.~\ref{fig:msm-analysis}c).
+The second right eigenvector, however, corresponds to the slowest process.
+Its minimal and maximal components indicate the microstates between which the process shifts probability.
+The relaxation timescale of this exchange process is exactly the corresponding implied timescale.
+In the projection onto the first two TICA components, we identify the slowest process as a probability shift between macrostate $\mathcal{S}_0$ and the rest of the system, with macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ in particular (Fig.~\ref{fig:msm-analysis}c).
 
-The mean first passage times (MFPTs) for the slowest process compute to
+The mean first passage times (MFPTs) out and into the macrostate $\mathcal{S}_0$ compute to
 \[ \begin{array}{crcr}
@@
 
-As an example for TPT, we compute the flux between macrostates $1$ and $2$ (Fig.~\ref{fig:msm-analysis}d).
+As an example for TPT, we compute the flux between macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$ (Fig.~\ref{fig:msm-analysis}d).
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
-Transition regions (macrostate $3$) can be identified by committor values $\approx \frac{1}{2}$.
+Transition regions (macrostate $\mathcal{S}_3$) can be identified by committor values $\approx \frac{1}{2}$.
 
-Next we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
-The SASA is a reasonable approximation for the instantaneous tryptophan flourescence signal and the auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.
-In our usual projection, we observe a high variance of the SASA in all macrostates except $0$ (Fig.~\ref{fig:msm-analysis}e), indicating that Trp-1 has only a low solvent exposure in this macrostate.
-The ACF itself yields only a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
-If, however, we prepare the system in macrostate $0$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
+To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
+The SASA is a reasonable approximation for the instantaneous tryptophan flourescence signal.
+The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.
+
+In the projection onto the first two TICA components, we observe a high variance of the SASA in all macrostates except $\mathcal{S}_0$ (Fig.~\ref{fig:msm-analysis}e).
+This indicates that Trp-1 has only a low solvent exposure in this macrostate.
+The ACF itself yields a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
+If, however, we prepare the system in macrostate $\mathcal{S}_0$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
 
 Finally, we sample $100$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g).
-We observe that, in macrostates $0$ and $1$, the backbone appears to shield the Trp-1 (shown in red) from the solvent.
-In the more probable macrostates $2$ and $3$, the Trp-1 appears to be much more exposed.
+We observe that, in macrostates $\mathcal{S}_0$ and $\mathcal{S}_1$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
+In the (more probable) macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$, the Trp-1 appears to be much more exposed.
 
@@
 
-We have summarized in this section how to conduct an MSM-based analysis of molecular dynamcis data using PyEMMA.
+In this section, we have summarized how to conduct an MSM-based analysis of molecular dynamcis data using PyEMMA.
 For the full analysis, please refer to the first notebook (00).
@@
 %%%%%%%%%%%%%%%%
+Most notebooks in this tutorial are in design and implementation a collaborative effort by CW, TH, and MKS; SO devised and implemented notebook 06 and contributed to notebook 00.
+CW and MKS wrote the manuscript; TH, SO, and FN revised the manuscript.
+The infrastructure for continuous integration and conda packaging was designed and implemented by MKS.
+
 For a more detailed description of author contributions, see the GitHub issue tracking and changelog at\\\githubrepository{}.
@@
 We are grateful to Brooke E.
-Husic for her thorough review of the notebooks in this tutorial as well as Camilla Ventura Santos and the entire computational molecular biology group for valuable discussion and feedback.
+Husic for her thorough review of the notebooks in this tutorial, Nuria Plattner for providing the pentapeptide simulation data, and Camilla Ventura Santos as well as the entire computational molecular biology group for valuable discussion and feedback.
 
@@
 %%%%%%%
+The authors declare no conflicting interests.
 

### 2018-07-23 manuscript.tex
macrostate relabeling and figure update

@@
 \caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which (b) PCCA++ identifies as four metastable states.
-(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_0$) and the two most probable states ($\mathcal{S}_2$, $\mathcal{S}_3$), whereas (d) the commitor $\mathcal{S}_1\to\mathcal{S}_2$ indicates that the most probable state $\mathcal{S}_3$ acts as a transition state for transitions between states $\mathcal{S}_1$ and $\mathcal{S}_2$.
-(e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_0$ in which the Trp-1 shows the lowest exposure to the solvent.
-(f) The Trp-1 SASA auto-correlation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the non-equilibrium condition $\mathcal{S}_0$ (bottom).
-(g) Super-imposed cartoon representations of $100$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(0,1)}$ but less so in $\mathcal{S}_{(2,3)}$.}
+(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas (d) the commitor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
+(e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
+(f) The Trp-1 SASA auto-correlation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the non-equilibrium condition $\mathcal{S}_1$ (bottom).
+(g) Super-imposed cartoon representations of $200$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(1,2)}$ but less so in $\mathcal{S}_{(3,4)}$.}
 \label{fig:msm-analysis}
@@
 
-A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=0,\dots,3$, for which we then approximate the absolute probabilities and free energies
+A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the absolute probabilities and free energies
 \[ \begin{array}{ccc}
@@
 \hline
-\mathcal{S}_0 & 0.004 & 5.565 \\
-\mathcal{S}_1 & 0.013 & 4.314 \\
-\mathcal{S}_2 & 0.021 & 3.864 \\
-\mathcal{S}_3 & 0.962 & 0.039
+\mathcal{S}_1 & 0.004 & 5.565 \\
+\mathcal{S}_2 & 0.013 & 4.314 \\
+\mathcal{S}_3 & 0.021 & 3.864 \\
+\mathcal{S}_4 & 0.962 & 0.039
 \end{array}\]
@@
 The relaxation timescale of this exchange process is exactly the corresponding implied timescale.
-In the projection onto the first two TICA components, we identify the slowest process as a probability shift between macrostate $\mathcal{S}_0$ and the rest of the system, with macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ in particular (Fig.~\ref{fig:msm-analysis}c).
+In the projection onto the first two TICA components, we identify the slowest process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$ in particular (Fig.~\ref{fig:msm-analysis}c).
 
-The mean first passage times (MFPTs) out and into the macrostate $\mathcal{S}_0$ compute to
+The mean first passage times (MFPTs) out and into the macrostate $\mathcal{S}_1$ compute to
 \[ \begin{array}{crcr}
@@
 \hline
-\mathcal{S}_0 \to \mathcal{S}_{(1,2,3)} & 8.9 & \pm & 1.8 \\
-\mathcal{S}_{(1,2,3)} \to \mathcal{S}_0 & 2487.5 & \pm &  504.7
+\mathcal{S}_1 \to \mathcal{S}_{(2,3,4)} & 8.9 & \pm & 1.8 \\
+\mathcal{S}_{(2,3,4)} \to \mathcal{S}_1 & 2487.5 & \pm &  504.7
 \end{array}\]
@@
 
-As an example for TPT, we compute the flux between macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$ (Fig.~\ref{fig:msm-analysis}d).
+As an example for TPT, we compute the flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
-Transition regions (macrostate $\mathcal{S}_3$) can be identified by committor values $\approx \frac{1}{2}$.
+Transition regions (macrostate $\mathcal{S}_4$) can be identified by committor values $\approx \frac{1}{2}$.
 
@@
 
-In the projection onto the first two TICA components, we observe a high variance of the SASA in all macrostates except $\mathcal{S}_0$ (Fig.~\ref{fig:msm-analysis}e).
+In the projection onto the first two TICA components, we observe a high variance of the SASA in all macrostates except $\mathcal{S}_1$ (Fig.~\ref{fig:msm-analysis}e).
 This indicates that Trp-1 has only a low solvent exposure in this macrostate.
 The ACF itself yields a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
-If, however, we prepare the system in macrostate $\mathcal{S}_0$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
+If, however, we prepare the system in macrostate $\mathcal{S}_1$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
 
-Finally, we sample $100$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g).
-We observe that, in macrostates $\mathcal{S}_0$ and $\mathcal{S}_1$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
-In the (more probable) macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$, the Trp-1 appears to be much more exposed.
+Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g).
+We observe that, in macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
+In the (more probable) macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$, the Trp-1 appears to be much more exposed.
 

### 2018-07-23 manuscript.tex
inserted line breaks at sentences breaks or reasonable punctuation marks
add pyemma workflow flowchart

@@
 
-In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial. Then, we address the software requirements to work through the lessons.
+In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial.
+Then, we address the software requirements to work through the lessons.
 
@@
 
-In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
+In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
+we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
+hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
 
 \subsection{Software/system requirements}
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{figures/workflowchart.pdf}
+\caption{The PyEMMA workflow: The MD trajectories are processed and discretized (first row).
+A Markov model is estimated from the resulting discrete trajectories and validated (middle row).
+By iterating between data processing and MSM estimation / validation, a valid model is obtained that can be analyzed (last row).}
+\label{fig:workflowchart}
+\end{figure}
 
@@
 
-Should the user prefer to not use Anaconda, a manual installation via the pip installer is possible. Alternatively, one can use the Binder service to view and run the tutorials online in any browser.
+Should the user prefer to not use Anaconda, a manual installation via the pip installer is possible.
+Alternatively, one can use the Binder service to view and run the tutorials online in any browser.
 
@@
 (b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and (d) the timeseries of the first two independent components show rare transition events.
-(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM; the shaded areas indicate $95\%$ confidence intervals.
-(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales; the shaded areas indicate $95\%$ confidence intervals.}
+(c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and
+(d) the timeseries of the first two independent components show rare transition events.
+(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM;
+the shaded areas indicate $95\%$ confidence intervals.
+(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales;
+the shaded areas indicate $95\%$ confidence intervals.}
 \label{fig:io-to-ck}
@@
 
-As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we start with a broad systematic analysis: we compute the (cross-validated) VAMP-2~\cite{vamp-preprint} score at lag time $0.5$ ns for three different featurizations and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
+As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we start with a broad systematic analysis:
+we compute the (cross-validated) VAMP-2~\cite{vamp-preprint} score at lag time $0.5$ ns for three different featurizations and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
 
@@
 \includegraphics{figure_2}
-\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which (b) PCCA++ identifies as four metastable states.
-(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas (d) the commitor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
+\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
+(a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
+(b) PCCA++ identifies as four metastable states.
+(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
+(d) the commitor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
 (e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.

### 2018-07-23 manuscript.tex

--- 2018-07-23 manuscript.tex
@@
 \begin{abstract}
-This tutorial provides an introduction to the PyEMMA project.
+This tutorial provides an introduction to the PyEMMA project, a Python application program interface for constructing Markov state models and related analyses from biomolecular simulation data.
 Using Jupyter notebooks, we will guide you through the basic functionality as well as the more common advanced mechanisms.
@@
 Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
+We also assume the reader is familiar with molecular dynamics analyses of proteins and peptides and commonly used structural components of these systems; for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}.
 
@@
 
-In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks}, we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt}, hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
+In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
+we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
+hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
 
@@
 The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow.
+The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
 The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics, an introduction to HMMs, and guidelines how to deal with common problems during an MSM estimation.
@@
 
-As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we start with a broad systematic analysis: we compute the (cross-validated) VAMP-2~\cite{vamp-preprint} score at lag time $0.5$ ns for three different featurizations and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
+To begin, we start with the raw trajectory data, and we seek to constuct and validate an MSM, which is represented by a transition matrix.
+As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we perform a broad systematic analysis using the variational approach for Markov processes (VAMP)~\cite{vamp-preprint}.
+We utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
+To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
 
 A subsequent TICA~\cite{tica,kinetic-maps} transformation of the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace.
-The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima and a time series visualization of said components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
+The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima, and a time series visualization of said components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
 We thus assume that our transformed feature selection describes one or more metastable processes.
 
-We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$~\cite{swope-its}.
+We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$~\cite{swope-its}
+(we note that the number of cluster centers can additionally be chosen using the VAMP score).
 The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
@@
 In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
-We can further see a converged fourth process (cyan line) which, however, decays for lag times above $2.5$ ns and might not be well resolved in the final model.
+We can further see a converged fourth process (cyan line); however, this process decays for lag times above $2.5$ ns and thus might not be well resolved in the final model.
 
-Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test which compares the right and the left side of the Chapman-Kolmogorov equation
+Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test, which compares the right and the left side of the Chapman-Kolmogorov equation
 \begin{equation}
@@
 with $T$ being the MSM's transition matrix.
-PyEMMA automatically estimates a new MSM transition matrix at lag time $k \tau$ and propagates the original transition matrix by the $k^\textrm{th}$ power.
-As we have resolved three slow timescales in the ITS test, we assume to have four metastable states present in the data and the subsequent CK test shows that indeed our model predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
+In practice, we utilize a coarse-grained approximation of $T$ for this test, obtained from PCCA++~\cite{pcca++}.
+To perform this test, PyEMMA automatically estimates a new MSM transition matrix at lag time $k \tau$ and propagates the original transition matrix by the $k^\textrm{th}$ power.
+As we have resolved three slow timescales in the ITS test, we assume that four metastable states are present in the data.
+The subsequent CK test shows that indeed our model predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
 
@@
 To understand which slow process occur, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening on which timescales.
-The first right eigenvector corresponds to the stationary process and is constant $1$.
+The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue $1$.
 The second right eigenvector, however, corresponds to the slowest process.
@@
 The relaxation timescale of this exchange process is exactly the corresponding implied timescale.
-In the projection onto the first two TICA components, we identify the slowest process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$ in particular (Fig.~\ref{fig:msm-analysis}c).
+In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$ in particular (Fig.~\ref{fig:msm-analysis}c).
 
@@
 
-As an example for TPT, we compute the flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
+As an example TPT analysis, we compute the flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
@@
 
-Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g).
+Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g) and render them using VMD~\cite{humphrey1996vmd}.
 We observe that, in macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
@@
 
-In this section, we have summarized how to conduct an MSM-based analysis of molecular dynamcis data using PyEMMA.
+In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamcis data using PyEMMA.
 For the full analysis, please refer to the first notebook (00).

### 2018-07-23 manuscript.tex

@@
 \subsection{Software/system requirements}
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{figures/workflowchart.pdf}
+\caption{The PyEMMA workflow: The MD trajectories are processed and discretized (first row).
+A Markov model is estimated from the resulting discrete trajectories and validated (middle row).
+By iterating between data processing and MSM estimation / validation, a valid model is obtained that can be analyzed (last row).}
+\label{fig:workflowchart}
+\end{figure}
 

### 2018-07-23 manuscript.tex
title update
merging internal review

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{PyEMMA: learning the ropes of Markov state modeling --- v\versionnumber}
+\title{Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber}
 
 \author[1\authfn{1}*]{Christoph Wehmeyer}
-\author[1\authfn{1}*]{Martin K. Scherer}
-\author[1\authfn{1}*]{Tim Hempel}
-\author[1*]{Simon Olsson}
+\author[1\authfn{1}]{Martin K. Scherer}
+\author[1\authfn{1}]{Tim Hempel}
+\author[1]{Simon Olsson}
+\author[2]{Brooke E. Husic}
 \author[1*]{Frank Noé}
 \affil[1]{Department of Mathematics and Computer Science, Freie Universität Berlin, Arnimallee 6, 14195 Berlin, Germany}
+\affil[2]{Department of Chemistry, Stanford University, Stanford, California 94305, USA}
 
 \corr{christoph.wehmeyer@fu-berlin.de}{CW}
-\corr{m.scherer@fu-berlin.de}{MKS}
-\corr{tim.hempel@fu-berlin.de}{TH}
-\corr{simon.olsson@fu-berlin.de}{SO}
+%\corr{m.scherer@fu-berlin.de}{MKS}
+%\corr{tim.hempel@fu-berlin.de}{TH}
+%\corr{simon.olsson@fu-berlin.de}{SO}
+%\corr{brookehusic@gmail.com}{BEH}
 \corr{frank.noe@fu-berlin.de}{FN}
@@
 \begin{abstract}
-This tutorial provides an introduction to the PyEMMA project, a Python application program interface for constructing Markov state models and related analyses from biomolecular simulation data.
-Using Jupyter notebooks, we will guide you through the basic functionality as well as the more common advanced mechanisms.
-Short exercises to self check your learning progress and a notebook on troubleshooting complete this basic introduction.
+This tutorial provides an introduction to the construction of Markov models of molecular kinetics from molecular dynamics (MD) trajectory data.
+Using Jupyter notebooks, we will guide the user through the basic functionality as well as the more common advanced mechanisms.
+Short exercises to self check the learning progress and a notebook on troubleshooting complete this basic introduction.
 \end{abstract}
@@
 
-For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a comprehensive overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the mathematical theory.
+For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail..
 
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{figures/workflowchart.pdf}
+\includegraphics[width=0.48\textwidth]{figure_1}
 \caption{The PyEMMA workflow: The MD trajectories are processed and discretized (first row).
@@
 
-The installation will contain a launcher command to start the Jupyter notebook server as well as the notebook files.
+The installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
 The data for the demonstrated test systems is downloaded upon the first use and is cached for future invocations of the tutorial.
@@
 Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajectories and perform validation tests (03).
-We can then analize the stationary and kinetic properties of the MSM (04), find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05), and compute expectation values for experimental observables (06).
+We can then analize the stationary and kinetic properties of the MSM (04),
+find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05),
+and compute expectation values for experimental observables (06).
 
-In the showcase notebook, we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns~\cite{pyemma}.
+In order to guide the user through the process before going into technical details,
+the first notebook goes through the full analysis pipeline for,
+we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns~\cite{pyemma}.
 
@@
 \begin{figure}
-\includegraphics{figure_1}
+\includegraphics{figure_2}
 \caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation.
@@
 To begin, we start with the raw trajectory data, and we seek to constuct and validate an MSM, which is represented by a transition matrix.
-As it is unknown which molecular features are best at representing the slow dynamics of the pentapeptide, we perform a broad systematic analysis using the variational approach for Markov processes (VAMP)~\cite{vamp-preprint}.
-We utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
+
+It is initially unknown which features (e.g.
+distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
+A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of nonequilibrium data, VAMP~\cite{vamp-preprint}.
+VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
+This score is computed via cross-validation in order to balance underfitting (too few features) and overfitting (too many features)~\cite{gmrq}.
+Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
 To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
+We note that recently, deep learning approaches for feature selection have been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.
 
-A subsequent TICA~\cite{tica,kinetic-maps} transformation of the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace.
+Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space to a low-dimensional space that can be discretized with higher resolution and better statistical efficiency.
+TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics.
+Here, TICA of the $\cos/\sin$-transformed backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff.
 The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima, and a time series visualization of said components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
@@
 with $\tau$ being the lag time.
-In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{noe-tmat-sampling}.
+In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling}.
 In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
@@
 \end{equation}
-with $T$ being the MSM's transition matrix.
+with $T$ being the MSM transition matrix.
 In practice, we utilize a coarse-grained approximation of $T$ for this test, obtained from PCCA++~\cite{pcca++}.
-To perform this test, PyEMMA automatically estimates a new MSM transition matrix at lag time $k \tau$ and propagates the original transition matrix by the $k^\textrm{th}$ power.
-As we have resolved three slow timescales in the ITS test, we assume that four metastable states are present in the data.
-The subsequent CK test shows that indeed our model predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
+To perform this test, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
+As we have identified three slow relaxation timescales in the ITS test,
+we perform the CK test on the four metastable states whose kinetics involves these three relaxation timescales.
+The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
 
@@
 \begin{figure}
-\includegraphics{figure_2}
+\includegraphics{figure_3}
 \caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
@@
 
-A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the absolute probabilities and free energies
+A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the absolute probabilities and relative free energies (defined up to an additive constant)
 \[ \begin{array}{ccc}
@@
 \label{eq:fe}
-G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln\left(\sum\limits_{j\in \mathcal{S}_i} \pi_j\right),
+G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln \sum\limits_{j\in \mathcal{S}_i} \pi_j,
 \end{equation}
@@
 
-To understand which slow process occur, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening on which timescales.
+In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening on which timescales.
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue $1$.
@@
 
-As an example TPT analysis, we compute the flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
+TPT~\cite{weinan-tpt,metzner-msm-tpt} is method to analyze the statistics of transition pathways.
+The TPT version of~\cite{noe-folding-pathways} can be conveniently applied to the estimated MSM.
+Here, we compute the TPT flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
@@
 
-Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g) and render them using VMD~\cite{humphrey1996vmd}.
+Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g) and render them using VMD~\cite{vmd}.
 We observe that, in macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
@@
 %%%%%%%%%%%%%%%%
-Most notebooks in this tutorial are in design and implementation a collaborative effort by CW, TH, and MKS; SO devised and implemented notebook 06 and contributed to notebook 00.
-CW and MKS wrote the manuscript; TH, SO, and FN revised the manuscript.
-The infrastructure for continuous integration and conda packaging was designed and implemented by MKS.
+CW, MKS, TH, SO, and FN designed research.
+CW, MKS, TH, SO, and BHE developed and tested notebooks.
+MKS developed the software infrastructure, test, and install environment.
+CW, MKS, TH, BEH, and FN wrote the manuscript.
 
@@
 %%%%%%%%%%%%%%%
-We are grateful to Brooke E.
-Husic for her thorough review of the notebooks in this tutorial, Nuria Plattner for providing the pentapeptide simulation data, and Camilla Ventura Santos as well as the entire computational molecular biology group for valuable discussion and feedback.
+We are grateful to Nuria Plattner for providing the pentapeptide simulation data and Camilla Ventura Santos as well as the entire computational molecular biology group for valuable discussion and feedback.
 

### 2018-07-24 manuscript.tex

@@
 with $T$ being the MSM transition matrix.
-In practice, we utilize a coarse-grained approximation of $T$ for this test, obtained from PCCA++~\cite{pcca++}.
-To perform this test, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
+To perform this test, we first choose a number of metatstable states appropriate for our system.
+Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
 As we have identified three slow relaxation timescales in the ITS test,

### 2018-07-24 manuscript.tex

@@
 
-In the following, we summarize the recommended theoretical background knowlegde of Markov state modeling for this tutorial.
+In the following, we summarize the recommended theoretical background knowledge of Markov state modeling for this tutorial.
 Then, we address the software requirements to work through the lessons.
@@
 Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajectories and perform validation tests (03).
-We can then analize the stationary and kinetic properties of the MSM (04),
+We can then analyze the stationary and kinetic properties of the MSM (04),
 find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05),
@@
 (c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and
-(d) the timeseries of the first two independent components show rare transition events.
+(d) the time series of the first two independent components show rare transition events.
 (e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM;
@@
 
-To begin, we start with the raw trajectory data, and we seek to constuct and validate an MSM, which is represented by a transition matrix.
+To begin, we start with the raw trajectory data, and we seek to construct and validate an MSM, which is represented by a transition matrix.
 
@@
 distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
-A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of nonequilibrium data, VAMP~\cite{vamp-preprint}.
+A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of non-equilibrium data, VAMP~\cite{vamp-preprint}.
 VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
@@
 with $T$ being the MSM transition matrix.
-To perform this test, we first choose a number of metatstable states appropriate for our system.
+To perform this test, we first choose a number of metastable states appropriate for our system.
 Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
@@
 (c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
-(d) the commitor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
+(d) the committor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
 (e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
@@
 To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
-The SASA is a reasonable approximation for the instantaneous tryptophan flourescence signal.
+The SASA is a reasonable approximation for the instantaneous tryptophan florescence signal.
 The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.
@@
 
-In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamcis data using PyEMMA.
+In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
 For the full analysis, please refer to the first notebook (00).

### 2018-07-24 manuscript.tex
some internal review comments

--- 2018-07-24 manuscript.tex
@@
 %\corr{simon.olsson@fu-berlin.de}{SO}
-%\corr{brookehusic@gmail.com}{BEH}
+%\corr{bhusic@stanford.edu}{BEH}
 \corr{frank.noe@fu-berlin.de}{FN}
@@
 \begin{itemize}
-\item NumPy -- vectorized operations on multi-dimensional arrays~\cite{numpy}
-\item SciPy -- linear algebra, sparse computations~\cite{scipy}
-\item Matplotlib -- Visualization~\cite{matplotlib}
-\item \textbf{PyEMMA} -- MSM/HMM estimation, validation, analysis, and visualization~\cite{pyemma}
-\item mdtraj -- reading MD simulation data, calculate observables from MD~\cite{mdtraj}
+% \item NumPy -- vectorized operations on multi-dimensional arrays~\cite{numpy}
+% \item SciPy -- linear algebra, sparse computations~\cite{scipy}
+% \item Matplotlib -- Visualization~\cite{matplotlib}
+\item \textbf{PyEMMA} -- MSM/HMM estimation, validation, analysis, and visualization, and its dependencies~\cite{pyemma}
+\item mdshare -- A downloader for MD data from a public server
+\item notebook -- The Jupyter notebook tool used for running the tutorials~\cite{jupyter}, along with extension packages jupyter_contrib_nbextensions and nbexamples
+\item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nguyen2017nglview}
+% \item mdtraj -- reading MD simulation data, calculate observables from MD~\cite{mdtraj}
 \end{itemize}
@@
 The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
-The eight subsequent notebooks (01--08) provide in-depth lessons on specific topics, an introduction to HMMs, and guidelines how to deal with common problems during an MSM estimation.
+The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics, and the last notebook (08) contains guidelines on how to deal with common problems during an MSM estimation.
 
@@
 
-In short, the workflow consists of extracting molecular features from the raw data (01) which we transform into a suitable, low-dimensional subspace and discretize (02).
-Then, we estimate a maximum likelihood or Bayesian MSM from the discrete trajectories and perform validation tests (03).
-We can then analize the stationary and kinetic properties of the MSM (04),
-find metastable macrostates and apply transition path theory (TPT) to identify the pathways of conformation change (05),
-and compute expectation values for experimental observables (06).
+In short, the workflow consists of:
+\begin{itemize}
+	\item extracting molecular features from the raw data (01),
+	\item transforming those features into a suitable, low-dimensional subspace (02),
+	\item discretizing the low-dimensional subsets into a state decomposition (02),
+	\item estimating a maximum likelihood or Bayesian MSM from the discrete trajectories and performing validation tests (03),
+	\item analyzing the stationary and kinetic properties of the MSM (04),
+	\item finding metastable macrostates and applying transition path theory (TPT) to identify the pathways of conformation change (05),
+	\item computing expectation values for experimental observables (06), and
+	\item coarse-graining the model using a hidden Markov model approach (07).
+\end{itemize}
 
@@
 
-\subsection{Estimation and validation}
+\subsection{Feature selection}
 
@@
 
+\subsection{Dimensionality reduction}
+
 Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space to a low-dimensional space that can be discretized with higher resolution and better statistical efficiency.
@@
 We thus assume that our transformed feature selection describes one or more metastable processes.
+
+\subsection{Discretization}
 
@@
 We can further see a converged fourth process (cyan line); however, this process decays for lag times above $2.5$ ns and thus might not be well resolved in the final model.
+
+\subsection{MSM estimation and validation}
 
@@
 Transition regions (macrostate $\mathcal{S}_4$) can be identified by committor values $\approx \frac{1}{2}$.
+
+\subsection{Connecting the MSM with experimental data}
 

### 2018-07-24 manuscript.tex
more edits

@@
 \author[1\authfn{1}]{Tim Hempel}
+\author[2]{Brooke E. Husic}
 \author[1]{Simon Olsson}
-\author[2]{Brooke E. Husic}
 \author[1*]{Frank Noé}
@@
 \includegraphics{figure_2}
-\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation.
+\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{humphrey1996vmd}.
 (b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c) The  sample density projected of the onto the first two independent components of a TICA transformation at lag time $\tau=0.5$ ns shows multiple density maxima and
-(d) the time series of the first two independent components show rare transition events.
-(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for estimating an MSM;
-the shaded areas indicate $95\%$ confidence intervals.
-(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales;
-the shaded areas indicate $95\%$ confidence intervals.}
+(c) The  sample density projected of the onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
+(d) the time series of the first two ICs show rare transition events.
+(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
+(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales.
+In (e) and (f), the shaded areas indicate $95\%$ confidence intervals.}
 \label{fig:io-to-ck}
@@
 The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
+We note that our MSM estimator results in a model that adheres to detailed balance.
 
@@
 The second right eigenvector, however, corresponds to the slowest process.
+The eigenvector components are real because of the detailed balance constraint enforced during MSM estimation.
 Its minimal and maximal components indicate the microstates between which the process shifts probability.

### 2018-07-24 manuscript.tex
a few latex changes; updated pdf

@@
 \item mdshare -- A downloader for MD data from a public server
-\item notebook -- The Jupyter notebook tool used for running the tutorials~\cite{jupyter}, along with extension packages jupyter_contrib_nbextensions and nbexamples
+\item notebook -- The Jupyter notebook tool used for running the tutorials~\cite{jupyter}, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
 \item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nguyen2017nglview}
@@
 CW, MKS, TH, SO, and FN designed research.
-CW, MKS, TH, SO, and BHE developed and tested notebooks.
+CW, MKS, TH, BEH, and SO developed and tested notebooks.
 MKS developed the software infrastructure, test, and install environment.

### 2018-07-24 manuscript.tex
brooke`s comments, fixes #3

@@
 The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow. 
+The figure that will be displayed in the following are created there and can be easily reproduced.
 The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
@@
 (f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales.
-In (e) and (f), the shaded areas indicate $95\%$ confidence intervals.}
+In (e) and (f), the shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
 \label{fig:io-to-ck}
@@
 with $\tau$ being the lag time.
-In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling}.
+In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
 In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
@@
 
-A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the absolute probabilities and relative free energies (defined up to an additive constant)
+A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
 \[ \begin{array}{ccc}

### 2018-07-24 manuscript.tex
fix citation

@@
 \includegraphics{figure_2}
-\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{humphrey1996vmd}.
+\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
 (b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.

### 2018-07-24 manuscript.tex
address and hacky fix to overflow

--- 2018-07-24 manuscript.tex
@@
 \affil[1]{Department of Mathematics and Computer Science, Freie Universität Berlin, Arnimallee 6, 14195 Berlin, Germany}
-\affil[2]{Department of Chemistry, Stanford University, Stanford, California 94305, USA}
+\affil[2]{Department of Chemistry, Stanford University, 333 Campus Drive, Stanford, California 94305, USA}
 
@@
 	\item extracting molecular features from the raw data (01),
-	\item transforming those features into a suitable, low-dimensional subspace (02),
+	\item transforming those features into a suitable, low dimensional subspace (02),
 	\item discretizing the low-dimensional subsets into a state decomposition (02),

### 2018-07-25 devtools/create_bibliography_nb.py
fix umlaut

@@ -1,0 +1,30 @@
+# encoding: utf-8
+
+import nbformat
+import os
+import re
+
+def export(in_file, out_file):
+    nb = nbformat.v4.new_notebook()
+    with open(in_file, encoding='utf8') as fh:
+        bibtex = fh.read()
+
+    src = '<!--bibtex\n{bibtex}\n-->'.format(bibtex=bibtex)
+
+    # remove bibdesk comments
+    src = re.sub(pattern='@comment.*\{.*\}\}', repl='', string=src, flags=re.DOTALL)
+    src = re.sub(pattern=r"\{\\'\{e\}\}", repl='é', string=src)
+    src = re.sub(pattern=r"\{\\'\{a\}\}", repl='á', string=src)
+    src = re.sub(pattern=r'\\"\{o\}', repl='ö', string=src)
+    cell = nbformat.v4.new_markdown_cell(src)
+    nb.cells.append(cell)
+
+    with open(out_file, 'w', encoding='utf-8') as fh:
+        nbformat.write(nb, fh)
+
+
+if __name__ == '__main__':
+    devtools_dir = os.path.abspath(os.path.dirname(__file__))
+    in_file = os.path.join(devtools_dir, '../manuscript/literature.bib')
+    out_file = os.path.join(devtools_dir, '../notebooks/Bibliography.ipynb')
+    export(in_file, out_file)

### 2018-07-25 manuscript.tex
text cleanup

@@
 
-In order to guide the user through the process before going into technical details,
-the first notebook goes through the full analysis pipeline for,
-we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns~\cite{pyemma}.
-
+For the remainder of this manuscript we will walk through the first notebook (00).
+In notebook 00 we analyse a data-set of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of 25 independent MD trajectories conducted in implicit solvent with a saving interval of 0.1ns.
+We present the results obtained in the notebook and thereby give an example how results generated using PyEMMA can be integrated into research publications.
 \subsection{Feature selection}
@@
 
-To begin, we start with the raw trajectory data, and we seek to construct and validate an MSM, which is represented by a transition matrix.
+In Markov state modelling our objective is to model the slow dynamics of a molecular process.
+In order to approximate the slow dynamics in a statistically efficient manner, a lower dimensional representation our simulation data is necessary.
+However, the features (fx torsion angles, distances or contacts) which best represent is the slow dynamic modes of a given molecular system is unknown a priori~\cite{NoeClementiReview}.
+Fortunately, the variation principle of molecular dynamics~\cite{noe-vac,nueske-vamk} and its more general form VAMP~\cite{vamp-preprint} provides a systematic means to gauge multiple different representations of the molecular dynamics against each other.
+In particular, we can use the VAMP score to directly compare the ability of certain features to capture slow dynamic modes in a particular molecular system.
+We here use the VAMP score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq}.
 
-It is initially unknown which features (e.g.
-distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
-A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of non-equilibrium data, VAMP~\cite{vamp-preprint}.
-VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
-This score is computed via cross-validation in order to balance underfitting (too few features) and overfitting (too many features)~\cite{gmrq}.
 Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
@@
 \subsection{Discretization}
+TICA gives us a reduced dimensionality of our molecular simulation data, which can greatly facilitate the definition of discrete Markov states necessary for MSM estimation.
+Here, we use the $k$-means algorithm to segment the four dimensional TICA into $k=75$ cluster centers.
+The number of cluster centers has been chosen as to optimize the VAMP score as in a manner identical to how the feature selection was carried out above.
 
-We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$~\cite{swope-its}
-(we note that the number of cluster centers can additionally be chosen using the VAMP score).
+\subsection{MSM estimation and validation}
+When estimating a MSM it is critical to chose a lag-time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space.
+Simultaneously we want to keep $\tau$ short to conserve as much of our expensive MD data as possible.
+Plotting the implied timescales (ITS) as a function $\tau$ can be a helpful diagnostic when selecting the MSM lag-time~\cite{swope-its}.
 The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
@@
 with $\tau$ being the lag time.
-In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
-In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
-We can further see a converged fourth process (cyan line); however, this process decays for lag times above $2.5$ ns and thus might not be well resolved in the final model.
+A necessary condition for Markovian dynamics in our reduced space is the ITS being approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfils this condition within the model uncertainty.
+The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
+In our example, we find the four slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
+Using this lag-time we can now estimate a (Bayesian) MSM with $\tau=0.5\mathrm{ns}$.
 
-\subsection{MSM estimation and validation}
-
-Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test, which compares the right and the left side of the Chapman-Kolmogorov equation
+To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
 \begin{equation}
@@
 with $T$ being the MSM transition matrix.
-To perform this test, we first choose a number of metastable states appropriate for our system.
-Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
-As we have identified three slow relaxation timescales in the ITS test,
-we perform the CK test on the four metastable states whose kinetics involves these three relaxation timescales.
-The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
-We note that our MSM estimator results in a model that adheres to detailed balance.
+The left-hand side of the equation corresponds to a MSM estimated at lag-time $k\tau$ with $k$ being an integer larger than 1, whereas the right-hand-side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
+Visualizing the full transition probability matrices $T$ is difficult, we therefore coarse-grain $T$ into a number of meta-stable states before performing the test.
+The number of meta-stable states can be identified using by identifying a large gap in the ITS plot.
+In our case we have multiple gaps and therefore multiple different options as to chose the number of meta-stable states.
+We chose 5 meta-stable states for the presentation here, as it is the largest number which only includes processes which are well above the MSM lag time.
+Given the number of meta-stable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$, in a coarse-grained representation.
+The CK test shows that the MSM we have estimated at lag time $\tau=0.5\,\mathrm{ns}$ indeed predicts the long-time time-scale behavior of our system (Fig.~\ref{fig:io-to-ck}f).
 
@@
 
-A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
+A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
 \[ \begin{array}{ccc}
@@
 \subsection{Connecting the MSM with experimental data}
+Markov state models can also be connected to experimental data.
+Connecting MSM analysis to experimental data can serve both as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
+Since we have both the stationary and dynamic properties of the molecular system encoded into the transition probability matrix, we can compute observables which involve both stationary ensemble averages as well as correlation functions.
 
-To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
-The SASA is a reasonable approximation for the instantaneous tryptophan florescence signal.
-The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.
+Here we look at florescence correlation of tryptophan residues as an example, since this is a realistic experimental observable for our penta-peptide system.
+In order to compute florescence correlation functions we have to have a microscopic, instantaneous value of our of the tryptophan florescence for each Markov state.
+These values are computed using a 'forward model'.
+We use the mdtraj library~\cite{mdtraj} to compute the surface accessible surface area (SASA) of Trp-1 in the penta-peptide as an approximation of the florescence signal.
 
-In the projection onto the first two TICA components, we observe a high variance of the SASA in all macrostates except $\mathcal{S}_1$ (Fig.~\ref{fig:msm-analysis}e).
-This indicates that Trp-1 has only a low solvent exposure in this macrostate.
-The ACF itself yields a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
-If, however, we prepare the system in macrostate $\mathcal{S}_1$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
-
-Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g) and render them using VMD~\cite{vmd}.
-We observe that, in macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
-In the (more probable) macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$, the Trp-1 appears to be much more exposed.
+With an approximation of the florescence in each of our Markov states we can use PyEMMA to compute the florescence auto-correlation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
+Note how the computed ACF has a very small response, or signal amplitude.
+Using PyEMMA can we simulate the relaxation of an observable if we had we prepared our molecular system in a non-equilibrium initial condition.
+The experimental counter-part of such prediction could be temperature or pressure jump experiments or a stopped flow assay.
+To illustrate such an experiment we initialize our molecular ensemble as the meta-stable distribution of $\mathcal{S}_1$ and follow the predicted florescence signal as it relaxes to equilibrium (\ref{fig:msm-analysis}e, lower).
+We see that the predicted relaxation signal has a much larger amplitude, making it more likely to be experimentally measurable.
 
@@
 MKS developed the software infrastructure, test, and install environment.
-CW, MKS, TH, BEH, and FN wrote the manuscript.
+CW, MKS, TH, BEH, SO, and FN wrote the manuscript.
 
@@
 %%%%%%%
-TH acknowledges financial support by SFB/TRR 186. SO acknowledges a postdoctral fellowship from the Alexander von Humboldt Stiftung.
+TH acknowledges financial support by SFB/TRR 186. SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.
 

### 2018-07-25 devtools/create_bibliography_nb.py
added references
Fixes #105

@@ -16,6 +16,8 @@ def export(in_file, out_file):
     src = re.sub(pattern=r"\{\\'\{e\}\}", repl='é', string=src)
     src = re.sub(pattern=r"\{\\'\{a\}\}", repl='á', string=src)
     src = re.sub(pattern=r'\\"\{o\}', repl='ö', string=src)
+    src = re.sub(pattern=r'\\"\{u\}', repl='ü', string=src)
+    src = re.sub(pattern='pcca\+\+', repl='pcca_plus_plus', string=src)
     cell = nbformat.v4.new_markdown_cell(src)
     nb.cells.append(cell)
 

### 2018-07-25 manuscript.tex

--- 2018-07-25 manuscript.tex
@@
 Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
-We also assume the reader is familiar with molecular dynamics analyses of proteins and peptides and commonly used structural components of these systems; for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}.
+We also assume the reader is familiar with molecular dynamics analyzes of proteins and peptides and commonly used structural components of these systems; for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}.
 
@@
 
-In order to guide the user through the process before going into technical details,
-the first notebook goes through the full analysis pipeline for,
-we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns~\cite{pyemma}.
-
+For the remainder of this manuscript we will walk through the first notebook (00).
+In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of 25 independent MD trajectories conducted in implicit solvent with a saving interval of 0.1~ns.
+We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
 \subsection{Feature selection}
@@
 
-To begin, we start with the raw trajectory data, and we seek to construct and validate an MSM, which is represented by a transition matrix.
+In Markov state modeling our objective is to model the slow dynamics of a molecular process.
+In order to approximate the slow dynamics in a statistically efficient manner, a lower dimensional representation of our simulation data is necessary.
+However, the features (e.g.
+torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system is unknown a priori~\cite{NoeClementiReview}.
+Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk} and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint} provides a systematic means to quantitatively compare multiple representations of the dataset.
+In particular, we can use the VAMP score to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
+%Here, we use the VAMP score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq}.
 
-It is initially unknown which features (e.g.
-distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
-A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of non-equilibrium data, VAMP~\cite{vamp-preprint}.
-VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
-This score is computed via cross-validation in order to balance underfitting (too few features) and overfitting (too many features)~\cite{gmrq}.
 Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
+We should always evaluate the score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
 To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
@@
 \subsection{Discretization}
+TICA yields a representation of our molecular simulation data with a reduced dimensionality, which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
+Here, we use the $k$-means algorithm to segment the four dimensional TICA into $k=75$ cluster centers.
+The number of cluster centers has been chosen to optimize the VAMP-2 score as in a manner identical to how the feature selection was carried out above, but is not presented in the showcase notebook.
 
-We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$~\cite{swope-its}
-(we note that the number of cluster centers can additionally be chosen using the VAMP score).
+\subsection{MSM estimation and validation}
+When estimating a MSM it is critical to chose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space.
+Simultaneously we want to keep $\tau$ short to conserve as much of our MD data as possible.
+Plotting the implied timescales (ITS) as a function $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
 The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
@@
 with $\tau$ being the lag time.
-In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
-In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
-We can further see a converged fourth process (cyan line); however, this process decays for lag times above $2.5$ ns and thus might not be well resolved in the final model.
+A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
+The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
+In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:io-to-ck}e).
+Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5\mathrm{ns}$.
 
-\subsection{MSM estimation and validation}
-
-Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test, which compares the right and the left side of the Chapman-Kolmogorov equation
+To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
 \begin{equation}
@@
 with $T$ being the MSM transition matrix.
-To perform this test, we first choose a number of metastable states appropriate for our system.
-Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
-As we have identified three slow relaxation timescales in the ITS test,
-we perform the CK test on the four metastable states whose kinetics involves these three relaxation timescales.
-The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
-We note that our MSM estimator results in a model that adheres to detailed balance.
+The left-hand side of the equation corresponds to a MSM estimated at lag time $k\tau$ with $k$ being an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
+Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a number of metastable states before performing the test.
+An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
+In our case we have multiple gaps and therefore various options choose from.
+We chose 5 metastable states for this analysis, as it is the largest number that only includes processes which are well above the MSM lag time.
+Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
+The CK test shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system (Fig.~\ref{fig:io-to-ck}f).
 
@@
 
-A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
+A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse-grain the $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
 \[ \begin{array}{ccc}
@@
 \mathcal{S}_3 & 0.021 & 3.864 \\
-\mathcal{S}_4 & 0.962 & 0.039
+\mathcal{S}_4 & 0.962 & 0.039 \\
+\mathcal{S}_5 & \text{TODO} & \text{TODO} \\
 \end{array}\]
@@
 \subsection{Connecting the MSM with experimental data}
+MSMs can also be connected to experimental observables.
+Connecting MSM analysis to experimental data can both serve as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
+Since we have both the stationary and dynamic properties of the molecular system encoded into the transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.
 
-To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
-The SASA is a reasonable approximation for the instantaneous tryptophan florescence signal.
-The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.
+As an example, here we look at the fluorescence correlation of the tryptophan in our system, Trp-1, since this is a realistic experimental observable for our pentapeptide system.
+In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each Markov state.
+%These values are computed using a 'forward model'.
+To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
 
-In the projection onto the first two TICA components, we observe a high variance of the SASA in all macrostates except $\mathcal{S}_1$ (Fig.~\ref{fig:msm-analysis}e).
-This indicates that Trp-1 has only a low solvent exposure in this macrostate.
-The ACF itself yields a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
-If, however, we prepare the system in macrostate $\mathcal{S}_1$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).
-
-Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g) and render them using VMD~\cite{vmd}.
-We observe that, in macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
-In the (more probable) macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$, the Trp-1 appears to be much more exposed.
+Now that we have an approximation of the fluorescence in each of our Markovian states, we can use PyEMMA to compute the fluorescence auto-correlation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
+Note how the computed ACF has a very small response (i.e., signal amplitude).
+Using PyEMMA, we can simulate the relaxation of an observable if we had we prepared our molecular system in a non-equilibrium initial condition.
+The experimental counterpart of such prediction could be temperature or pressure jump experiments or a stopped flow assay.
+To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-analysis}e, lower).
+We see that the predicted relaxation signal has a much larger amplitude for the non-equilibrium initialization, making it more likely to be experimentally measurable.
 
@@
 
-In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
+In this section, we have summarized how to conduct a MSM-based analysis of biomolecular dynamics data using PyEMMA.
 For the full analysis, please refer to the first notebook (00).
@@
 %%%%%%%%%%%%%%%%
-% This section mustt describe the actual contributions of
+% This section must describe the actual contributions of
 % author. Since this is an electronic-only journal, there is

### 2018-07-25 manuscript.tex
full read through;

@@
 \begin{abstract}
-This tutorial provides an introduction to the construction of Markov models of molecular kinetics from molecular dynamics (MD) trajectory data.
+This tutorial provides an introduction to the construction of Markov models of molecular kinetics from molecular dynamics trajectory data.
 Using Jupyter notebooks, we will guide the user through the basic functionality as well as the more common advanced mechanisms.
@@
 Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
-We also assume the reader is familiar with molecular dynamics analyzes of proteins and peptides and commonly used structural components of these systems; for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}.
+We also assume the reader is familiar with MD analyses of proteins and peptides and commonly used structural components of these systems (for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}).
 
@@
 In the following, we summarize the recommended theoretical background knowledge of Markov state modeling for this tutorial.
-Then, we address the software requirements to work through the lessons.
+Then, we address the software required to work through the lessons.
 
@@
 
-For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail..
+For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
 
@@
 \includegraphics[width=0.48\textwidth]{figure_1}
-\caption{The PyEMMA workflow: The MD trajectories are processed and discretized (first row).
-A Markov model is estimated from the resulting discrete trajectories and validated (middle row).
-By iterating between data processing and MSM estimation / validation, a valid model is obtained that can be analyzed (last row).}
+\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
+A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
+By iterating between data processing and MSM estimation/validation, a dynamical model is obtained that can be analyzed (last row).}
 \label{fig:workflowchart}
@@
 The user can install all necessary packages in one step using the \texttt{conda} command provided by the Anaconda Python stack (\url{https://anaconda.com}).
-We rely on Anaconda, because it resolves and installs dependencies, and provides pre-compiled versions of hard to build packages like SciPy.
+We recommend Anaconda because it resolves and installs dependencies as well as provides pre-compiled versions of common packages.
 
-The installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
+The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
 The data for the demonstrated test systems is downloaded upon the first use and is cached for future invocations of the tutorial.
 
-The underlying software stack for running the tutorial importantly consists of:
+The underlying software stack for running the tutorial consists of:
 \begin{itemize}
@@
 
-The software is currently supported for Python versions $3.5$ and $3.6$ on the operating systems Linux, OSX, and Windows.
+The tutorial software is currently supported for Python versions $3.5$ and $3.6$ on the operating systems Linux, OSX, and Windows.
 
@@
 This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
-The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow. 
-The figure that will be displayed in the following are created there and can be easily reproduced.
+The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow for a small example system. 
+% The figure that will be displayed in the following are created there and can be easily reproduced.
 The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
-The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics, and the last notebook (08) contains guidelines on how to deal with common problems during an MSM estimation.
+The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics, and the last notebook (08) contains guidelines on how to deal with common problems during MSM estimation.
 
@@
 	\item transforming those features into a suitable, low dimensional subspace (02),
-	\item discretizing the low-dimensional subsets into a state decomposition (02),
+	\item discretizing the low dimensional subsets into a state decomposition (02),
 	\item estimating a maximum likelihood or Bayesian MSM from the discrete trajectories and performing validation tests (03),
 	\item analyzing the stationary and kinetic properties of the MSM (04),
-	\item finding metastable macrostates and applying transition path theory (TPT) to identify the pathways of conformation change (05),
+	\item finding metastable macrostates and applying transition path theory (TPT) to identify the pathways of conformational change (05),
 	\item computing expectation values for experimental observables (06), and
-	\item coarse-graining the model using a hidden Markov model approach (07).
+	\item coarse-graining the MSM using a hidden Markov model approach (07).
 \end{itemize}
@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of 25 independent MD trajectories conducted in implicit solvent with a saving interval of 0.1~ns.
+In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of $25$ independent MD trajectories conducted in implicit solvent with a frames saved at an interval of 0.1~ns.
 We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
+The figure that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
+
 \subsection{Feature selection}
@@
 \includegraphics{figure_2}
-\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a) The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
-(b) The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c) The  sample density projected of the onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
-(d) the time series of the first two ICs show rare transition events.
-(e) The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
-(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of four metastable states accurately predicts the kinetic behavior on longer timescales.
-In (e) and (f), the shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
+\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
+(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
+(c)~The  sample density projected of the onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
+(d)~the time series of the first two ICs show rare transition events.
+(e)~The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
+(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of five metastable states accurately predicts the kinetic behavior on longer timescales.
+In~(e) and~(f), the shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
 \label{fig:io-to-ck}
@@
 In order to approximate the slow dynamics in a statistically efficient manner, a lower dimensional representation of our simulation data is necessary.
-However, the features (e.g.
-torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system is unknown a priori~\cite{NoeClementiReview}.
-Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk} and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint} provides a systematic means to quantitatively compare multiple representations of the dataset.
-In particular, we can use the VAMP score to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
-%Here, we use the VAMP score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq}.
+However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system is unknown a priori~\cite{NoeClementiReview}.
+Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk} and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint} provide a systematic means to quantitatively compare multiple representations of the simulation data.
+In particular, we can use a scalar score obtained using the VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
 
 Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
-We should always evaluate the score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
-To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
-We note that recently, deep learning approaches for feature selection have been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.
+We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
+To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-ck}b).
+We note that deep learning approaches for feature selection have recently been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.
 
@@
 
-Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space to a low-dimensional space that can be discretized with higher resolution and better statistical efficiency.
-TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics.
-Here, TICA of the $\cos/\sin$-transformed backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff.
-The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima, and a time series visualization of said components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
-We thus assume that our transformed feature selection describes one or more metastable processes.
+Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space, which typically contains many degrees of freedom, to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
+TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
+Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
+The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima, and a time series visualization of these components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
+We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.
 
@@
 TICA yields a representation of our molecular simulation data with a reduced dimensionality, which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
-Here, we use the $k$-means algorithm to segment the four dimensional TICA into $k=75$ cluster centers.
-The number of cluster centers has been chosen to optimize the VAMP-2 score as in a manner identical to how the feature selection was carried out above, but is not presented in the showcase notebook.
+Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
+The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above, but is not presented in the showcase notebook.
 
 \subsection{MSM estimation and validation}
-When estimating a MSM it is critical to chose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space.
-Simultaneously we want to keep $\tau$ short to conserve as much of our MD data as possible.
+When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to conserve as much of our MD data as possible.
 Plotting the implied timescales (ITS) as a function $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
-The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
+The ITS $t_i$ approximates the decorrelation time of the $i$\th process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
 \begin{equation}
@@
 \end{equation}
-with $\tau$ being the lag time.
+where $\tau$ is the MSM lag time.
 A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
@@
 In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:io-to-ck}e).
-Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5\mathrm{ns}$.
+Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.
 
@@
 \end{equation}
-with $T$ being the MSM transition matrix.
-The left-hand side of the equation corresponds to a MSM estimated at lag time $k\tau$ with $k$ being an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
-Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a number of metastable states before performing the test.
+where $T$ is the MSM transition matrix.
+The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k$\th power.
+Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
 An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
@@
 We chose 5 metastable states for this analysis, as it is the largest number that only includes processes which are well above the MSM lag time.
-Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
-The CK test shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system (Fig.~\ref{fig:io-to-ck}f).
+% Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
+The CK test shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).
 
@@
 (a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
-(b) PCCA++ identifies as four metastable states.
-(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
+(b) PCCA++ identifies as five metastable states.
+(c) TODO:~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
 (d) the committor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
 (e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
-(f) The Trp-1 SASA auto-correlation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the non-equilibrium condition $\mathcal{S}_1$ (bottom).
-(g) Super-imposed cartoon representations of $200$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(1,2)}$ but less so in $\mathcal{S}_{(3,4)}$.}
+(f) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).
+(g) Superimposed cartoon representations of $200$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(1,2)}$ but less so in $\mathcal{S}_{(3,4)}$.}
 \label{fig:msm-analysis}
@@
 We can now directly extract several thermodynamic and kinetic properties from the estimated and validated model.
-An example of the former is the free energy surface in the projection onto the first two TICA components (Fig.~\ref{fig:msm-analysis}a) reweighted based on the MSM stationary distribution.
+An example of the former is the free energy surface in the projection onto the first two TICA components (Fig.~\ref{fig:msm-analysis}a) reweighted by the MSM stationary distribution.
 
-A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse-grain the $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
+A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
 \[ \begin{array}{ccc}
@@
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue $1$.
-The second right eigenvector, however, corresponds to the slowest process.
-The eigenvector components are real because of the detailed balance constraint enforced during MSM estimation.
-Its minimal and maximal components indicate the microstates between which the process shifts probability.
-The relaxation timescale of this exchange process is exactly the corresponding implied timescale.
-In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$ in particular (Fig.~\ref{fig:msm-analysis}c).
+The second right eigenvector, however, corresponds to the slowest process 
+(the eigenvector components are real because of the detailed balance constraint enforced during MSM estimation).
+The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.
+The relaxation timescale of this exchange process is exactly the corresponding implied timescale, which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
+TODO:~In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$ in particular (Fig.~\ref{fig:msm-analysis}c).
 
-The mean first passage times (MFPTs) out and into the macrostate $\mathcal{S}_1$ compute to
+TODO:~The mean first passage times (MFPTs) out of and into the macrostate $\mathcal{S}_1$ compute to
 \[ \begin{array}{crcr}
@@
 
-TPT~\cite{weinan-tpt,metzner-msm-tpt} is method to analyze the statistics of transition pathways.
+TPT~\cite{weinan-tpt,metzner-msm-tpt} is a method used to analyze the statistics of transition pathways.
 The TPT version of~\cite{noe-folding-pathways} can be conveniently applied to the estimated MSM.
-Here, we compute the TPT flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
+TODO:~Here, we compute the TPT flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
@@
 \subsection{Connecting the MSM with experimental data}
-MSMs can also be connected to experimental observables.
+MSMs can also be analyzed in the context of experimental observables.
 Connecting MSM analysis to experimental data can both serve as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
-Since we have both the stationary and dynamic properties of the molecular system encoded into the transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.
+Since we have both the stationary and dynamic properties of the molecular system encoded in the MSM transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.
 
-As an example, here we look at the fluorescence correlation of the tryptophan in our system, Trp-1, since this is a realistic experimental observable for our pentapeptide system.
-In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each Markov state.
+As an example, here we look at the fluorescence correlation Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
+In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the five macrostates.
 %These values are computed using a 'forward model'.
 To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
+Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
+Note how the computed ACF has a very small response (i.e., signal amplitude).
 
-Now that we have an approximation of the fluorescence in each of our Markovian states, we can use PyEMMA to compute the fluorescence auto-correlation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
-Note how the computed ACF has a very small response (i.e., signal amplitude).
 Using PyEMMA, we can simulate the relaxation of an observable if we had we prepared our molecular system in a non-equilibrium initial condition.
-The experimental counterpart of such prediction could be temperature or pressure jump experiments or a stopped flow assay.
+The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
 To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-analysis}e, lower).
@@
 
-In this section, we have summarized how to conduct a MSM-based analysis of biomolecular dynamics data using PyEMMA.
+In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
 For the full analysis, please refer to the first notebook (00).

### 2018-07-26 manuscript.tex
resolve TODOs related to macrostate numbers

@@
 (b) PCCA++ identifies as five metastable states.
-(c) TODO:~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the two most probable states ($\mathcal{S}_3$, $\mathcal{S}_4$), whereas
-(d) the committor $\mathcal{S}_2\to\mathcal{S}_3$ indicates that the most probable state $\mathcal{S}_4$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_3$.
+(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
+(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that the most probable state $\mathcal{S}_5$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_4$.
 (e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
@@
 \hline
-\mathcal{S}_1 & 0.004 & 5.565 \\
-\mathcal{S}_2 & 0.013 & 4.314 \\
-\mathcal{S}_3 & 0.021 & 3.864 \\
-\mathcal{S}_4 & 0.962 & 0.039 \\
-\mathcal{S}_5 & \text{TODO} & \text{TODO} \\
+\mathcal{S}_1 & 0.004 & 5.567 \\
+\mathcal{S}_2 & 0.014 & 4.293 \\
+\mathcal{S}_3 & 0.021 & 3.841 \\
+\mathcal{S}_4 & 0.021 & 3.875 \\
+\mathcal{S}_5 & 0.940 & 0.062 \\
 \end{array}\]
@@
 The relaxation timescale of this exchange process is exactly the corresponding implied timescale, which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
-TODO:~In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$ in particular (Fig.~\ref{fig:msm-analysis}c).
+In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_4$ and $\mathcal{S}_5$ in particular (Fig.~\ref{fig:msm-analysis}c).
 
-TODO:~The mean first passage times (MFPTs) out of and into the macrostate $\mathcal{S}_1$ compute to
+The mean first passage times (MFPTs) out of and into the macrostate $\mathcal{S}_1$ compute to
 \[ \begin{array}{crcr}
@@
 \hline
-\mathcal{S}_1 \to \mathcal{S}_{(2,3,4)} & 8.9 & \pm & 1.8 \\
-\mathcal{S}_{(2,3,4)} \to \mathcal{S}_1 & 2487.5 & \pm &  504.7
+\mathcal{S}_1 \to \mathcal{S}_{(2,3,4,5)} & 9.0 & \pm & 1.9 \\
+\mathcal{S}_{(2,3,4,5)} \to \mathcal{S}_1 & 2496.4 & \pm &  470.0
 \end{array}\]
@@
 The TPT version of~\cite{noe-folding-pathways} can be conveniently applied to the estimated MSM.
-TODO:~Here, we compute the TPT flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_3$ (Fig.~\ref{fig:msm-analysis}d).
+Here, we compute the TPT flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_4$ (Fig.~\ref{fig:msm-analysis}d).
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.

### 2018-07-26 manuscript.tex
figure 3 adjust caption + macrostate numbers

@@
 (d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that the most probable state $\mathcal{S}_5$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_4$.
-(e) The Trp-1 SASA shows high variance for all macrostates except $\mathcal{S}_1$ in which the Trp-1 shows the lowest exposure to the solvent.
-(f) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).
-(g) Superimposed cartoon representations of $200$ molecular conformations sampled from each of the four macrostates show that the backbone appears to obstruct the Trp-1 exposure to the solvent in $\mathcal{S}_{(1,2)}$ but less so in $\mathcal{S}_{(3,4)}$.}
+(e) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).}
 \label{fig:msm-analysis}
@@
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
-Transition regions (macrostate $\mathcal{S}_4$) can be identified by committor values $\approx \frac{1}{2}$.
+Transition regions (macrostates $\mathcal{S}_{(1,5)}$) can be identified by committor values $\approx \frac{1}{2}$.
 

### 2018-07-26 manuscript.tex
added fig 4 and related text

@@
 (c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
-(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that the most probable state $\mathcal{S}_5$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_4$.
+(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$. 
+% the most probable state $\mathcal{S}_5$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_4$. 
 (e) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).}
@@
 Transition regions (macrostates $\mathcal{S}_{(1,5)}$) can be identified by committor values $\approx \frac{1}{2}$.
+
+The transition network can be additionally visualized by plotting representative structures of the five metastable states $\mathcal{S}_{(1-5)}$ according to their committor probability (Fig.~\ref{fig:tpt-network}).
+It is easy to see from this depiction that the dominant pahtway from $\mathcal{S}_2$ to $\mathcal{S}_4$ proceeds through $\mathcal{S}_5$.
+
+\begin{figure}
+\includegraphics{figure_4}
+\caption{Visualization of the transition paths from $\mathcal{S}_2$ to $\mathcal{S}_4$:
+Metastable states $\mathcal{S}_{(1-5)}$ are represented by an ensemble of representative structures and are arranged along the horizonal axis according to their committor probabilities.
+The three main transition pathways starting from $\mathcal{S}_2$ and ending in $\mathcal{S}_4$ are depicted by gray arrows with thickness proportional to the transition flux.
+The dominant pathway proceeds through $\mathcal{S}_5$.}
+\label{fig:tpt-network}
+\end{figure}
 

### 2018-07-26 manuscript.tex
fixes from tim

@@
 \begin{abstract}
-This tutorial provides an introduction to the construction of Markov models of molecular kinetics from molecular dynamics trajectory data.
-Using Jupyter notebooks, we will guide the user through the basic functionality as well as the more common advanced mechanisms.
+This tutorial provides an introduction to the construction of Markov models of molecular kinetics from molecular dynamics trajectory data with the PyEMMA software.
+Using tutorial notebooks, we will guide the user through the basic functionality as well as the more common advanced mechanisms.
 Short exercises to self check the learning progress and a notebook on troubleshooting complete this basic introduction.
@@
 
-In short, the workflow consists of:
+In short, the workflow for a full analysis of an MD dataset might consist of,
 \begin{itemize}
@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of $25$ independent MD trajectories conducted in implicit solvent with a frames saved at an interval of 0.1~ns.
+In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of 0.1~ns.
 We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
-The figure that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
+The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
 
@@
 (b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The  sample density projected of the onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
+(c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
 (d)~the time series of the first two ICs show rare transition events.
@@
 In order to approximate the slow dynamics in a statistically efficient manner, a lower dimensional representation of our simulation data is necessary.
-However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system is unknown a priori~\cite{NoeClementiReview}.
+However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
 Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk} and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint} provide a systematic means to quantitatively compare multiple representations of the simulation data.
-In particular, we can use a scalar score obtained using the VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
+In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
 
-Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
+Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
@@
 Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
-The sample density projected onto the first two independent components (Fig.~\ref{fig:io-to-ck}c) exhibits several maxima, and a time series visualization of these components from the first trajectory (Fig.~\ref{fig:io-to-ck}d) shows discrete jumps between these density maxima.
+The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-ck}c).
+Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-ck}d).
+ % time series visualization of these components from the first trajectory shows discrete jumps between these maxima (Fig.~\ref{fig:io-to-ck}d).
 We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.
@@
 Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
-The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above, but is not presented in the showcase notebook.
+The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above, which is shown in the showcase notebook (00).
 
 \subsection{MSM estimation and validation}
-When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to conserve as much of our MD data as possible.
-Plotting the implied timescales (ITS) as a function $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
-The ITS $t_i$ approximates the decorrelation time of the $i$\th process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
+When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
+Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
+The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
 \begin{equation}
 \label{eq:its}
-t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}
+t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
 \end{equation}
-where $\tau$ is the MSM lag time.
 A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
@@
 where $T$ is the MSM transition matrix.
-The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k$\th power.
+The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
 Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
 An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
-In our case we have multiple gaps and therefore various options choose from.
-We chose 5 metastable states for this analysis, as it is the largest number that only includes processes which are well above the MSM lag time.
+% In our case we have multiple gaps and therefore various options to choose from.
+For this analysis, we chose 5 metastable states.
+%, as it is the largest number that only includes processes which are well above the MSM lag time.
 % Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
-The CK test shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).
+The CK test confirms that this is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).
 
@@
 
-In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening on which timescales.
+In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening and their timescales.
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue $1$.
@@
 The committor projection onto the first two TICA components shows that it is constant within the metastable states defined above.
-Transition regions (macrostates $\mathcal{S}_{(1,5)}$) can be identified by committor values $\approx \frac{1}{2}$.
+Transition regions (macrostates $\mathcal{S}_{(1,3,5)}$) can be identified by committor values $\approx \frac{1}{2}$.
 
 The transition network can be additionally visualized by plotting representative structures of the five metastable states $\mathcal{S}_{(1-5)}$ according to their committor probability (Fig.~\ref{fig:tpt-network}).
-It is easy to see from this depiction that the dominant pahtway from $\mathcal{S}_2$ to $\mathcal{S}_4$ proceeds through $\mathcal{S}_5$.
+It is easy to see from this depiction that the dominant pathway from $\mathcal{S}_2$ to $\mathcal{S}_4$ proceeds through $\mathcal{S}_5$.
 
@@
 
-As an example, here we look at the fluorescence correlation Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
+As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
 In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the five macrostates.
@@
 
-Using PyEMMA, we can simulate the relaxation of an observable if we had we prepared our molecular system in a non-equilibrium initial condition.
+Using PyEMMA, we can simulate the relaxation of an observable if we had prepared our molecular system in a nonequilibrium initial condition.
 The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
 To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-analysis}e, lower).
-We see that the predicted relaxation signal has a much larger amplitude for the non-equilibrium initialization, making it more likely to be experimentally measurable.
+We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization, making it more likely to be experimentally measurable.
 

### 2018-07-26 manuscript/manuscript.tex

--- 2018-07-26 manuscript.tex
@@
 As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
-In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the five macrostates.
+In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original $75$ MSM microstates.
 %These values are computed using a 'forward model'.

### 2018-07-26 manuscript/manuscript.tex
version 0.3
citation check
submission version and minor changes

@@
 
-\newcommand{\versionnumber}{0.1} % TODO: can we extract this number from a git tag?
+\newcommand{\versionnumber}{1.0} % TODO: can we extract this number from a git tag?
 \newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}
@@
 \begin{itemize}
-% \item NumPy -- vectorized operations on multi-dimensional arrays~\cite{numpy}
-% \item SciPy -- linear algebra, sparse computations~\cite{scipy}
-% \item Matplotlib -- Visualization~\cite{matplotlib}
 \item \textbf{PyEMMA} -- MSM/HMM estimation, validation, analysis, and visualization, and its dependencies~\cite{pyemma}
@@
 \item notebook -- The Jupyter notebook tool used for running the tutorials~\cite{jupyter}, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
-\item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nguyen2017nglview}
-% \item mdtraj -- reading MD simulation data, calculate observables from MD~\cite{mdtraj}
+\item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nglview}
 \end{itemize}
@@
 
-Should the user prefer to not use Anaconda, a manual installation via the pip installer is possible.
+Should the user prefer not to use Anaconda, a manual installation via the pip installer is possible.
 Alternatively, one can use the Binder service to view and run the tutorials online in any browser.
@@
 The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow for a small example system.
-% The figure that will be displayed in the following are created there and can be easily reproduced.
 The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of 0.1~ns.
+In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns.
 We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
@@
 To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-ck}b).
+
 We note that deep learning approaches for feature selection have recently been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.

### 2018-07-27 manuscript/manuscript.tex
adding Frank's change requests

@@
 
-\newcommand{\versionnumber}{1.0} % TODO: can we extract this number from a git tag?
+\newcommand{\versionnumber}{0.3} % TODO: can we extract this number from a git tag?
 \newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}
@@
 For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
+Two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
 
-In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
+In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
 we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
@@
 
-A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
+A spectral clustering using the PCCA++ algorithm~\cite{pcca++,Deuflhard2005-pcca,Kube2007-pcca+} allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
 \[ \begin{array}{ccc}
@@
 
+\subsection{Advanced Methods}
+
+While the present tutorial covers Markov State Modeling 101, we encourage the user to explore other, more recent extensions of the methodology.
+Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram} can be used to combine unbiased and biased simulations so as to probe kinetics of very rare events~\cite{trammbar}; MEMMs are implemented in PyEMMA.
+Recently, there have been steps towards replacing the traditional user-directed pipeline (involving featurizing, reducing dimension, discretizing, MSM estimation and coarse-graining) by a single end-to-end deep learning method such as VAMPnets~\cite{vampnet}.
+Other deep learning methods for performing the dimension reduction~\cite{tae}, finding reaction coordinates for enhanced sampling~\cite{hernandez-vde,Sultan2018-vde-enhanced-sampling,Ribeiro2018-rave}, and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research on its own right.
+Implementations of some of these methods are available or is under development in the package deeptime \url{github.com/markovmodel/deeptime}.
+
 \section{Author Contributions}

### 2018-07-27 manuscript/manuscript.tex
adding missing reference to fig.2a [ci skip]

@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns.
+In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-ck}a), consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns.
 We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.

### 2018-07-27 manuscript/manuscript.tex
tex cleanup [ci skip]
added amm citation [ci skip]

@@
 
-\usepackage{lipsum} % Required to insert dummy text
 \usepackage[version=4]{mhchem}
@@
 
-\newcommand{\versionnumber}{0.3} % TODO: can we extract this number from a git tag?
+\newcommand{\versionnumber}{0.3}
 \newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}
@@
 \corr{christoph.wehmeyer@fu-berlin.de}{CW}
-%\corr{m.scherer@fu-berlin.de}{MKS}
-%\corr{tim.hempel@fu-berlin.de}{TH}
-%\corr{simon.olsson@fu-berlin.de}{SO}
-%\corr{bhusic@stanford.edu}{BEH}
 \corr{frank.noe@fu-berlin.de}{FN}
@@
 
-In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
+In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
 we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
@@
 Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-ck}d).
- % time series visualization of these components from the first trajectory shows discrete jumps between these maxima (Fig.~\ref{fig:io-to-ck}d).
 We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.
@@
 An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
-% In our case we have multiple gaps and therefore various options to choose from.
 For this analysis, we chose 5 metastable states.
-%, as it is the largest number that only includes processes which are well above the MSM lag time.
-% Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
 The CK test confirms that this is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).
@@
 (d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.
-% the most probable state $\mathcal{S}_5$ acts as a transition state for transitions between states $\mathcal{S}_2$ and $\mathcal{S}_4$. 
 (e) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).}
@@
 \subsection{Connecting the MSM with experimental data}
+
 MSMs can also be analyzed in the context of experimental observables.
@@
 In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original $75$ MSM microstates.
-%These values are computed using a 'forward model'.
 To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.

### 2018-07-27 manuscript/manuscript.tex
addressing review comments [ci skip]

@@
 For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
-Two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
+Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
 
@@
 
-While the present tutorial covers Markov State Modeling 101, we encourage the user to explore other, more recent extensions of the methodology.
+While the present tutorial is intended to cover Markov State Modeling 101, we encourage the user to explore other, more recent extensions of the methodology.
 Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram} can be used to combine unbiased and biased simulations so as to probe kinetics of very rare events~\cite{trammbar}; MEMMs are implemented in PyEMMA.
@@
 Other deep learning methods for performing the dimension reduction~\cite{tae}, finding reaction coordinates for enhanced sampling~\cite{hernandez-vde,Sultan2018-vde-enhanced-sampling,Ribeiro2018-rave}, and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research on its own right.
-Implementations of some of these methods are available or is under development in the package deeptime \url{github.com/markovmodel/deeptime}.
+Implementations of some of these methods are available or are under development in the deeptime package \url{github.com/markovmodel/deeptime}.
 

### 2018-08-27 manuscript/manuscript.tex
Amending manuscript (fixes #143)

@@
 (c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
-(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.
-(e) The Trp-1 SASA autocorrelation function yields a weak signal (top) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (bottom).}
+(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
 \label{fig:msm-analysis}
@@
 
+\begin{figure}
+\includegraphics{figure_5}
+\caption{The Trp-1 SASA autocorrelation function yields a weak signal (a) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (b).
+The solid/orange lines denote the maximum likelihood MSM result; the dashed/blue lines and the the shaded areas indicate sample means and $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
+\label{fig:msm-exp-obs}
+\end{figure}
+
 MSMs can also be analyzed in the context of experimental observables.
@@
 To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
-Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
+Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-exp-obs}a).
 Note how the computed ACF has a very small response (i.e., signal amplitude).
@@
 The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
-To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-analysis}e, lower).
+To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
 We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization, making it more likely to be experimentally measurable.

### 2018-08-27 manuscript/manuscript.tex
minor caption changes (#137)

@@
 \includegraphics{figure_2}
-\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone: (a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
 (b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
@@
 \includegraphics{figure_3}
-\caption{Exemplary analysis of the conformational dynamics of a pentapeptide backbone:
-(a) The reweighted free energy surface projected onto the first two independent components exhibits five minima which
-(b) PCCA++ identifies as five metastable states.
-(c) The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
-(d) the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
+(b)~PCCA++ identifies as five metastable states.
+(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
+(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
 \label{fig:msm-analysis}
@@
 \includegraphics{figure_4}
-\caption{Visualization of the transition paths from $\mathcal{S}_2$ to $\mathcal{S}_4$:
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+visualization of the transition paths from $\mathcal{S}_2$ to $\mathcal{S}_4$.
 Metastable states $\mathcal{S}_{(1-5)}$ are represented by an ensemble of representative structures and are arranged along the horizonal axis according to their committor probabilities.
@@
 \includegraphics{figure_5}
-\caption{The Trp-1 SASA autocorrelation function yields a weak signal (a) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (b).
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: the Trp-1 SASA autocorrelation function yields a weak signal (a) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (b).
 The solid/orange lines denote the maximum likelihood MSM result; the dashed/blue lines and the the shaded areas indicate sample means and $95\%$ confidence intervals computed with a Bayesian sampling procedure.}

### 2018-08-27 manuscript/manuscript.tex
cite fig.1 (#137)
mentioning+citing matplotlib (#137)

@@
 \item mdshare -- A downloader for MD data from a public server
-\item notebook -- The Jupyter notebook tool used for running the tutorials~\cite{jupyter}, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
+\item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
+\item matplotlib -- A plotting library~\cite{matplotlib}
 \item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nglview}
@@
 
-In short, the workflow for a full analysis of an MD dataset might consist of,
+In short, the workflow (Fig.~\ref{fig:workflowchart}) for a full analysis of an MD dataset might consist of,
 \begin{itemize}

### 2018-08-27 manuscript/manuscript.tex
adding Binder link (fixes #137)
minor change [ci skip]

@@
 Should the user prefer not to use Anaconda, a manual installation via the pip installer is possible.
-Alternatively, one can use the Binder service to view and run the tutorials online in any browser.
+Alternatively, one can use the Binder service (\url{https://mybinder.org}) to view and run the tutorials online in any browser.
 
@@
 \includegraphics{figure_5}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: the Trp-1 SASA autocorrelation function yields a weak signal (a) which, however, can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$ (b).
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however, (b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
 The solid/orange lines denote the maximum likelihood MSM result; the dashed/blue lines and the the shaded areas indicate sample means and $95\%$ confidence intervals computed with a Bayesian sampling procedure.}

### 2018-08-28 manuscript/manuscript.tex
first take writing

@@
 
-In the following, we summarize the recommended theoretical background knowledge of Markov state modeling for this tutorial.
+In the following, we summarize the recommended theory and background knowledge of Markov state modeling for this tutorial.
 Then, we address the software required to work through the lessons.
 
-\subsection{Background knowledge}
+\subsection{Essential theory}
+\label{sec:theory}
+
+Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
+In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
+The Markov property asserts that these states are ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
+In order to obtain Markovian dynamics, a lag time must be chosen at which the transitions between states are analyzed.
+This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
+
+In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
+Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a counts matrix.
+Then, the counts matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
+To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the counts matrix to the transition probability matrix.
+Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
+The requirement of detailed balance indicates that we assume our system is in equilibrium.
+We additionally require that the system is ergodic.
+
+We can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
+The CK property for a Markovian matrix is,
+
+\begin{equation}
+T(k \tau) = T^k(\tau).
+\end{equation}
+
+\noindent{}whe left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
+By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
+
+Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues,
+
+\begin{equation}
+\label{eq:transmat}
+T(\tau) \circ \phi_i = \lambda_i \phi_i,
+\end{equation}
+
+\noindent{}where the eigenvalues are indexed in decreasing order.
+The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
+The subsequent eigenvalues $\lambda_{i<1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
+The right eigenvectors, conventionally written as $\psi_i$, represent the flux into and out of the states of the MSM.
+The corresponding left eigenvectors contain the same information weighted by the stationary distribution.
+
+The timescale of a given dynamical process is a function of the relevant eigenvalue and the lag time at which the model was defined,
+
+\begin{equation}
+\label{eq:timescales}
+t_i \equiv -\frac{\tau}{\log(|\lambda_i|)}.
+\end{equation}
+
+\subsection{MSM construction the variational approach}
+\label{sec:construction}
+
+The theory described in the previous section required the decomposition of the phase or configuration space occupied by a dynamical system into discrete, disjoint states.
+Starting from the output of an MD simulation of a protein, there are several steps that can be taken to obtain an MSM from the original configuration space:
+
+\begin{itemize}
+	\item Featurization. The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
+	\item Dimensionality reduction. Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
+	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps} is used to transform the features into a set of slow coordinates.
+	\item Clustering. This is the step at which the state decomposition occurs. The features or TICs are grouped using a clustering algorithm such as $k$-means into a set of states.
+	\item Transition matrix approximation. At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
+\end{itemize}
+
+\subsection{Background knowledge and resources}
 \label{sec:background}
 
-For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
+For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
 Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
@@
 In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
-we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
+we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
 hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
@@
 
-To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
-\begin{equation}
-\label{eq:ck}
-T(k \tau) = T^k(\tau)
-\end{equation}
-where $T$ is the MSM transition matrix.
+To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test.
+% The CK test compares the right and the left side of the Chapman-Kolmogorov equation
+% \begin{equation}
+% \label{eq:ck}
+% T(k \tau) = T^k(\tau)
+% \end{equation}
+% where $T$ is the MSM transition matrix.
 The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.

### 2018-08-28 manuscript/manuscript.tex
vamp

@@
 Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
-The requirement of detailed balance indicates that we assume our system is in equilibrium.
+The requirement of detailed balance indicates that we assume our system is reversible.
 We additionally require that the system is ergodic.
@@
 
-\noindent{}whe left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
+\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
 By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
@@
 The right eigenvectors, conventionally written as $\psi_i$, represent the flux into and out of the states of the MSM.
-The corresponding left eigenvectors contain the same information weighted by the stationary distribution.
+The corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.
 
@@
 \begin{itemize}
-	\item Featurization. The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
-	\item Dimensionality reduction. Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
+	\item Featurization -- The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
+	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
 	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps} is used to transform the features into a set of slow coordinates.
-	\item Clustering. This is the step at which the state decomposition occurs. The features or TICs are grouped using a clustering algorithm such as $k$-means into a set of states.
-	\item Transition matrix approximation. At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
+	\item Clustering -- This is the step at which the state decomposition occurs.
+	The features or TICs are grouped using a clustering algorithm such as $k$-means into a set of states.
+	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
 \end{itemize}
+
+It is apparent that there are many choices involved in MSM construction such as what features should be used and how many states should be chosen.
+In 2013, the variational approach to conformational dynamics (VAC) was derived, which enabled an objective comparison among different state decomposition choices for models built with the same Markovian lag time~\cite{noe-vac}.
+More recently, the more general variational approach to Markov processes (VAMP) has been developed in order to facilitate the approximation and comparison of reversible models for basis sets that are continuous, as opposed to discrete states~\cite{vamp-preprint}.
+The VAMP can thus be used to perform model selection.
+Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model.
+However, the MSM lag time cannot be optimized using VAMP, and must be chosen using a separate validation as described previously.
 

### 2018-08-29 manuscript/manuscript.tex
line fixes

@@
 In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
-The Markov property asserts that these states are ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
+The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
 In order to obtain Markovian dynamics, a lag time must be chosen at which the transitions between states are analyzed.
-This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
+%This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
 
 In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
-Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a counts matrix.
-Then, the counts matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
-To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the counts matrix to the transition probability matrix.
-Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
+Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
+Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
+To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,
+
+\begin{equation}
+\label{eq:balance}
+p_t(i)p(i,j) = p_t(j)p(j,i),
+\end{equation}
+
+\noindent{}where $p_t(i)$ represents the probability of being in state $i$ at time $t$, and $p(i,j)$ is the probability of transitioning from state $i$ to $j$ at the next time step.
+% Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
 The requirement of detailed balance indicates that we assume our system is reversible.
@@
 
-We can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
+When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
+Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
+The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
+\begin{equation}
+\label{eq:its}
+t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
+\end{equation}
+
+Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
 The CK property for a Markovian matrix is,
@@
 
-\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
+\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
 By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
@@
 The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
-The subsequent eigenvalues $\lambda_{i<1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
-The right eigenvectors, conventionally written as $\psi_i$, represent the flux into and out of the states of the MSM.
+The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
+The right eigenvectors, conventionally written as $\psi_i$, each represent a dynamical process (for $i>1$), and the coefficients of the eigenvectors represent the flux into and out of the MSM states that characterizes that process.
 The corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.
@@
 	\item Clustering -- This is the step at which the state decomposition occurs.
-	The features or TICs are grouped using a clustering algorithm such as $k$-means into a set of states.
+	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
 	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
@@
 \subsection{MSM estimation and validation}
-When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
-Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
+% When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
+% Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
 The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
-\begin{equation}
-\label{eq:its}
-t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
-\end{equation}
+% \begin{equation}
+% \label{eq:its}
+% t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
+% \end{equation}
 A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.

### 2018-08-29 manuscript/manuscript.tex
amending/adding installation instructions (fixes #141)

--- 2018-08-28 manuscript/manuscript.tex
@@
 
-In the following, we summarize the recommended theory and background knowledge of Markov state modeling for this tutorial.
+In the following, we summarize the recommended theoretical background knowledge of Markov state modeling for this tutorial.
 Then, we address the software required to work through the lessons.
 
-\subsection{Essential theory}
-\label{sec:theory}
-
-Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
-In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
-The Markov property asserts that these states are ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
-In order to obtain Markovian dynamics, a lag time must be chosen at which the transitions between states are analyzed.
-This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
-
-In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
-Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a counts matrix.
-Then, the counts matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
-To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the counts matrix to the transition probability matrix.
-Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
-The requirement of detailed balance indicates that we assume our system is in equilibrium.
-We additionally require that the system is ergodic.
-
-We can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
-The CK property for a Markovian matrix is,
-
-\begin{equation}
-T(k \tau) = T^k(\tau).
-\end{equation}
-
-\noindent{}whe left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
-By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
-
-Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues,
-
-\begin{equation}
-\label{eq:transmat}
-T(\tau) \circ \phi_i = \lambda_i \phi_i,
-\end{equation}
-
-\noindent{}where the eigenvalues are indexed in decreasing order.
-The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
-The subsequent eigenvalues $\lambda_{i<1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
-The right eigenvectors, conventionally written as $\psi_i$, represent the flux into and out of the states of the MSM.
-The corresponding left eigenvectors contain the same information weighted by the stationary distribution.
-
-The timescale of a given dynamical process is a function of the relevant eigenvalue and the lag time at which the model was defined,
-
-\begin{equation}
-\label{eq:timescales}
-t_i \equiv -\frac{\tau}{\log(|\lambda_i|)}.
-\end{equation}
-
-\subsection{MSM construction the variational approach}
-\label{sec:construction}
-
-The theory described in the previous section required the decomposition of the phase or configuration space occupied by a dynamical system into discrete, disjoint states.
-Starting from the output of an MD simulation of a protein, there are several steps that can be taken to obtain an MSM from the original configuration space:
-
-\begin{itemize}
-	\item Featurization. The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
-	\item Dimensionality reduction. Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
-	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps} is used to transform the features into a set of slow coordinates.
-	\item Clustering. This is the step at which the state decomposition occurs. The features or TICs are grouped using a clustering algorithm such as $k$-means into a set of states.
-	\item Transition matrix approximation. At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
-\end{itemize}
-
-\subsection{Background knowledge and resources}
+\subsection{Background knowledge}
 \label{sec:background}
 
-For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
+For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
 Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
@@
 In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
-we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
+we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
 hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
@@
 We recommend Anaconda because it resolves and installs dependencies as well as provides pre-compiled versions of common packages.
+The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
 
-The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
+You can install the tutorial's dependencies in a new conda environment and start the notebook server via
+\begin{verbatim}
+conda create -n pyemma_tutorials
+conda activate pyemma_tutorials
+conda install -c conda-forge pyemma_tutorials
+pyemma_tutorials
+\end{verbatim}
+or refer to \githubrepository{} for more detailed installation and usage instructions.
+
 The data for the demonstrated test systems is downloaded upon the first use and is cached for future invocations of the tutorial.
@@
 
-To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test.
-% The CK test compares the right and the left side of the Chapman-Kolmogorov equation
-% \begin{equation}
-% \label{eq:ck}
-% T(k \tau) = T^k(\tau)
-% \end{equation}
-% where $T$ is the MSM transition matrix.
+To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
+\begin{equation}
+\label{eq:ck}
+T(k \tau) = T^k(\tau)
+\end{equation}
+where $T$ is the MSM transition matrix.
 The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.

### 2018-08-29 manuscript/manuscript.tex
manual word break for package name

@@
 \item mdshare -- A downloader for MD data from a public server
-\item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
+\item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\-\_contrib\_nbextensions and nbexamples
 \item matplotlib -- A plotting library~\cite{matplotlib}

### 2018-08-29 manuscript/manuscript.tex
[manuscript] append section about large systems

--- 2018-08-29 manuscript/manuscript.tex
@@
 We recommend Anaconda because it resolves and installs dependencies as well as provides pre-compiled versions of common packages.
+
 The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
-
-You can install the tutorial's dependencies in a new conda environment and start the notebook server via
-\begin{verbatim}
-conda create -n pyemma_tutorials
-conda activate pyemma_tutorials
-conda install -c conda-forge pyemma_tutorials
-pyemma_tutorials
-\end{verbatim}
-or refer to \githubrepository{} for more detailed installation and usage instructions.
-
 The data for the demonstrated test systems is downloaded upon the first use and is cached for future invocations of the tutorial.
@@
 
+\subsection{Markov models of large systems}
+For modeling large systems, challenges arise that are mostly system dependent. 
+One of them is that finding a state space discretization using e.g. $k$-means becomes difficult in very high dimensional spaces. 
+It is thus important to choose a feature space and coordinate transform that yield a low dimensional representation of the slow processes.
+Hidden Markov models might further help to overcome poor discretization to a certain extend.
+
+In addition, the slowest process in a system as identified by a Markov model might not be the one a modeler is interested in. 
+For instance, there might be a side chain flip that only occured once in a data set but has no biological relevance. 
+This problem can often be mitigated by choosing a more specific set of features.
+
+We explain how to deal with the technical challenges such as limited memory in the tutorials. 
+More details on how to model complex systems with the techniques presented here are described by e.g. ~\cite{plattner_protein_2015,plattner_complete_2017}.
+
 \subsection{Advanced Methods}

### 2018-08-29 manuscript/manuscript.tex
addressed new changes

--- 2018-08-29 manuscript/manuscript.tex
@@
 
-In the following, we summarize the recommended theoretical background knowledge of Markov state modeling for this tutorial.
+In the following, we summarize the recommended theory and background knowledge of Markov state modeling for this tutorial.
 Then, we address the software required to work through the lessons.
 
-\subsection{Background knowledge}
+\subsection{Essential theory}
+\label{sec:theory}
+
+Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
+In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
+The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
+In order to obtain Markovian dynamics, a lag time must be chosen at which the transitions between states are analyzed.
+%This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
+
+In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
+Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
+Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
+To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,
+
+\begin{equation}
+\label{eq:balance}
+p_t(i) p(i \rightarrow j) = p_t(j)p(j \rightarrow i),
+\end{equation}
+
+\noindent{}where $p_t(i)$ represents the probability of being in state $i$ at time $t$, and $p(i \rightarrow j)$ is the probability of transitioning from state $i$ to $j$ at the next time step.
+% Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
+The requirement of detailed balance indicates that we assume our system is reversible.
+We additionally require that the system is ergodic.
+
+When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
+Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
+The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
+\begin{equation}
+\label{eq:its}
+t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
+\end{equation}
+
+\noindent{}When the ITS become approximately constant with the lag time, we say that our timescales have converged and choose the smallest lag time with the converged timescales in order to maximize the model's temporal resolution.
+
+Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
+The CK property for a Markovian matrix is,
+
+\begin{equation}
+T(k \tau) = T^k(\tau),
+\end{equation}
+
+\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
+By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
+
+Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues,
+
+\begin{equation}
+\label{eq:transmat}
+T(\tau) \circ \phi_i = \lambda_i \phi_i,
+\end{equation}
+
+\noindent{}where the eigenvalues are indexed in decreasing order.
+The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
+The right eigenvector $\psi_1$ is a vector consisting of $1$'s.
+The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
+The right eigenvectors $\psi_i$ each represent a dynamical process (for $i>1$), and the coefficients of the eigenvectors represent the flux into and out of the MSM states that characterizes that process.
+The corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.
+
+The timescale of a given dynamical process is a function of the relevant eigenvalue and the lag time at which the model was defined,
+
+\begin{equation}
+\label{eq:timescales}
+t_i \equiv -\frac{\tau}{\log(|\lambda_i|)}.
+\end{equation}
+
+\subsection{MSM construction the variational approach}
+\label{sec:construction}
+
+The theory described in the previous section required the decomposition of the phase or configuration space occupied by a dynamical system into discrete, disjoint states.
+Starting from the output of an MD simulation of a protein, there are several steps that can be taken to obtain an MSM from the original configuration space:
+
+\begin{itemize}
+	\item Featurization -- The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
+	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
+	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps} is used to transform the features into a set of slow coordinates.
+	\item Clustering -- This is the step at which the state decomposition occurs.
+	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
+	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
+\end{itemize}
+
+It is apparent that there are many choices involved in MSM construction such as what features should be used and how many states should be chosen.
+In 2013, the variational approach to conformational dynamics (VAC) was derived, which enabled an objective comparison among different state decomposition choices for models built with the same Markovian lag time~\cite{noe-vac}.
+More recently, the more general variational approach to Markov processes (VAMP) has been developed in order to facilitate the approximation and comparison of reversible models for basis sets that are continuous, as opposed to discrete states~\cite{vamp-preprint}.
+The VAMP can thus be used to perform model selection.
+Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model.
+However, the MSM lag time cannot be optimized using VAMP, and must be chosen using a separate validation as described previously.
+
+\subsection{Background knowledge and resources}
 \label{sec:background}
 
-For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
+For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
 Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.

### 2018-08-30 manuscript/manuscript.tex
PR review requested changes cw, beh

--- 2018-08-29 manuscript/manuscript.tex
@@
 \subsection{Markov models of large systems}
-For modeling large systems, challenges arise that are mostly system dependent. 
-One of them is that finding a state space discretization using e.g. $k$-means becomes difficult in very high dimensional spaces. 
-It is thus important to choose a feature space and coordinate transform that yield a low dimensional representation of the slow processes.
-Hidden Markov models might further help to overcome poor discretization to a certain extend.
+For modeling large systems, challenges may arise that are mostly system dependent. 
+For example, one of them is that finding a state space discretization using e.g.~ $k$-means becomes difficult in very high dimensional spaces. 
+It is thus important to choose a feature space and coordinate transformion that yield a low dimensional representation of the slow processes.
+Hidden Markov models might further help to overcome poor discretization to a certain extend \cite{noe-proj-hid-msm}.
 
 In addition, the slowest process in a system as identified by a Markov model might not be the one a modeler is interested in. 
-For instance, there might be a side chain flip that only occured once in a data set but has no biological relevance. 
+For instance, there might be a side chain flip that only occurred once in a data set but has no biological relevance. 
 This problem can often be mitigated by choosing a more specific set of features.
 
-We explain how to deal with the technical challenges such as limited memory in the tutorials. 
+Further, large systems come with technical challenges such as huge memory requirements.
+We explain how to deal with those in the tutorials. 
 More details on how to model complex systems with the techniques presented here are described by e.g. ~\cite{plattner_protein_2015,plattner_complete_2017}.

### 2018-08-30 manuscript/manuscript.tex
simon edits

--- 2018-08-29 manuscript/manuscript.tex
@@
 In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
-The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
-In order to obtain Markovian dynamics, a lag time must be chosen at which the transitions between states are analyzed.
+The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ after a time $\tau$ is independent of the history of the system before it was in $i$.
 %This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
@@
 Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
-To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,
+To ensure that the transition probability matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,
 
@@
 The requirement of detailed balance indicates that we assume our system is reversible.
-We additionally require that the system is ergodic.
+We additionally require that the system is ergodic, which means that every state is accessible from every other state.
 
-When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
+When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
 Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.

### 2018-08-31 manuscript/manuscript.tex
[manuscript] large system section, incorporate review + major improvement

--- 2018-08-30 manuscript/manuscript.tex
@@
 
-\subsection{Markov models of large systems}
-For modeling large systems, challenges may arise that are mostly system dependent. 
-For example, one of them is that finding a state space discretization using e.g.~ $k$-means becomes difficult in very high dimensional spaces. 
-It is thus important to choose a feature space and coordinate transformion that yield a low dimensional representation of the slow processes.
-Hidden Markov models might further help to overcome poor discretization to a certain extend \cite{noe-proj-hid-msm}.
+\subsection{Modeling large systems}
+When estimating MSMs for large systems, challenges may arise that are mostly system dependent.
 
-In addition, the slowest process in a system as identified by a Markov model might not be the one a modeler is interested in. 
-For instance, there might be a side chain flip that only occurred once in a data set but has no biological relevance. 
+A case in point is the curse of dimensionality:
+it is hard to discretize a high dimensional feature space, not only because it is computationally demanding.
+More important, Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001} and, thus, cluster assignment based on that norm are prone to yield poor discretizations.
+Especially for large systems, it is thus particularly important for first find a suitable set of features, and to further apply dimension reduction techniques (e.g.
+TICA, VAMP, if applicable) to obtain a low-dimensional representation of the slow dynamics.
+Hidden Markov models might further mitigate poor discretization to a certain extend~\cite{noe-proj-hid-msm}.
+
+In addition, the slowest process in a system as identified by an MSM or HMM might not be the one a modeler is interested in.
+For instance, the slowest process might correspond to a biologically irrelevant side chain flip that only occurred once in the data set.
 This problem can often be mitigated by choosing a more specific set of features.
 
-Further, large systems come with technical challenges such as huge memory requirements.
-We explain how to deal with those in the tutorials. 
-More details on how to model complex systems with the techniques presented here are described by e.g. ~\cite{plattner_protein_2015,plattner_complete_2017}.
+The technical challenges with large systems are usually high demands in memory and computation time; we explain how to deal with those in the tutorials.
+
+More details on how to model complex systems with the techniques presented here are described e.g. by~\cite{plattner_protein_2015,plattner_complete_2017}.
 

### 2018-08-31 manuscript/manuscript.tex
renamed analysis figures
fixes #142

--- 2018-08-30 manuscript/manuscript.tex
@@
 Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model.
-However, the MSM lag time cannot be optimized using VAMP, and must be chosen using a separate validation as described previously.
+However, the MSM lag time cannot be optimized using VAMP, and must be chosen using a separate validation as described above~\cite{husic2017note}.
 
@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-ck}a), consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns.
+In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a), consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns.
 We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
@@
 (c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
-(d)~the time series of the first two ICs show rare transition events.
-(e)~The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
-(f) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of five metastable states accurately predicts the kinetic behavior on longer timescales.
-In~(e) and~(f), the shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
-\label{fig:io-to-ck}
+(d)~the time series of the first two ICs show rare transition events.}
+\label{fig:io-to-tica}
+\end{figure}
+
+\begin{figure}
+\includegraphics{figure_3}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
+(b) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of five metastable states accurately predicts the kinetic behavior on longer timescales.
+The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling scheme.
+The black line indicates the timescale cutoff and, thus, all ITS withion the grey-shaded area are not necessarily resolved.
+In both panels, the (non-grey) shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
+\label{fig:its-and-ck}
 \end{figure}
@@
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
-To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-ck}b).
+To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
 
@@
 Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
-The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-ck}c).
-Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-ck}d).
+The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-tica}c).
+Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
 We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.
@@
 The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
-In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:io-to-ck}e).
+In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:its-and-ck}a).
 Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.
@@
 An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
-For this analysis, we chose 5 metastable states.
-The CK test confirms that this is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (Fig.~\ref{fig:io-to-ck}f).
+For this analysis, we chose five metastable states.
+The CK test (Fig.~\ref{fig:its-and-ck}b) shows a good agreement between reestimation at higher lagtimes (black/solid lines) and higher powers of the original transition matrix (blue/dashed lines).
+Thus, it confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).
 
@@
 \begin{figure}
-\includegraphics{figure_3}
+\includegraphics{figure_4}
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
@@
 \begin{figure}
-\includegraphics{figure_4}
+\includegraphics{figure_5}
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
@@
 \begin{figure}
-\includegraphics{figure_5}
+\includegraphics{figure_6}
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however, (b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.

### 2018-09-03 manuscript/manuscript.tex
minor changes, review beh

--- 2018-08-31 manuscript/manuscript.tex
@@
 
-A case in point is the curse of dimensionality:
-it is hard to discretize a high dimensional feature space, not only because it is computationally demanding.
-More important, Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001} and, thus, cluster assignment based on that norm are prone to yield poor discretizations.
-Especially for large systems, it is thus particularly important for first find a suitable set of features, and to further apply dimension reduction techniques (e.g.
-TICA, VAMP, if applicable) to obtain a low-dimensional representation of the slow dynamics.
-Hidden Markov models might further mitigate poor discretization to a certain extend~\cite{noe-proj-hid-msm}.
+A case in point is the curse of dimensionality:~it is difficult to discretize a high dimensional feature space.
+While it is somewhat computationally demanding, more importantly, Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001} and thus cluster assignments based on that norm may yield a poor discretization.
+Especially for large systems, it is thus particularly important to first find a suitable set of features, and to further apply dimensionality reduction techniques (e.g.~TICA, VAMP, if applicable) to obtain a low dimensional representation of the slow dynamics.
+Hidden Markov models (HMMs) might further mitigate poor discretization to a certain extent~\cite{noe-proj-hid-msm}.
 
-In addition, the slowest process in a system as identified by an MSM or HMM might not be the one a modeler is interested in.
+Furthermore, the slowest process in a system as identified by an MSM or HMM might not be the one a modeler is interested in~\cite{banushkina_nonparametric_2015}.
 For instance, the slowest process might correspond to a biologically irrelevant side chain flip that only occurred once in the data set.
-This problem can often be mitigated by choosing a more specific set of features.
+This problem may be mitigated by choosing a more specific set of features.
 
-The technical challenges with large systems are usually high demands in memory and computation time; we explain how to deal with those in the tutorials.
+Additional technical challenges for large systems include high demands on memory and computation time; we explain how to deal with those in the tutorials.
 
-More details on how to model complex systems with the techniques presented here are described e.g. by~\cite{plattner_protein_2015,plattner_complete_2017}.
+More details on how to model complex systems with the techniques presented here are described e.g.~by~\cite{plattner_protein_2015,plattner_complete_2017}.
 

### 2018-09-04 manuscript/manuscript.tex

--- 2018-08-31 manuscript/manuscript.tex
@@
 When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
-Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
+Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
 The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
@@
 
-\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
+\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
 By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
@@
 The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
-The right eigenvector $\psi_1$ is a vector consisting of $1$'s.
-The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than $1$ and correspond to dynamical processes within the system.
+The right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
+The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than~$1$ and correspond to dynamical processes within the system.
 The right eigenvectors $\psi_i$ each represent a dynamical process (for $i>1$), and the coefficients of the eigenvectors represent the flux into and out of the MSM states that characterizes that process.
@@
 
-The tutorial software is currently supported for Python versions $3.5$ and $3.6$ on the operating systems Linux, OSX, and Windows.
+The tutorial software is currently supported for Python versions~$3.5$ and~$3.6$ on the operating systems Linux, OSX, and Windows.
 
@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a), consisting of $25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of $0.1$~ns.
-We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
+In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a), consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
+We present the results obtained in this notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
 The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
@@
 (b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$ ns shows multiple density maxima and
-(d)~the time series of the first two ICs show rare transition events.}
+(c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple density maxima and
+(d)~the time series of the first two ICs of the first trajectory.}
 \label{fig:io-to-tica}
@@
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The convergence behavior of the first four implied timescales indicates that a lag time of $\tau=0.5$ ns is suitable for MSM estimation.
-(b) A Chapman-Kolmogorov test shows that an MSM estimated at lag time $\tau=0.5$ ns under the assumption of five metastable states accurately predicts the kinetic behavior on longer timescales.
-The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling scheme.
-The black line indicates the timescale cutoff and, thus, all ITS withion the grey-shaded area are not necessarily resolved.
-In both panels, the (non-grey) shaded areas indicate $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
+(a)~The convergence behavior of the implied timescales associated with the four slowest processes.
+(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 meta-stable states..
+The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
+The black line indicates where implied timescales are equal to the lag time, whereas the grey area indicates all implied timescales faster than the lag time.
+In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}
 \label{fig:its-and-ck}
@@
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
-To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
+To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of~$0.5$~ns.
+We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
+This suggests that backbone torsions are the best of the options evaluated for MSM construction.
 
@@
 TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
-Here, performing TICA on the backbone torsions at lag time $0.5$ ns yields a four dimensional subspace using a $95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
+Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
 The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-tica}c).
@@
 % When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
-% Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
+% Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
 The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
@@
 A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
-The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
-In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:its-and-ck}a).
+The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with~$100$ samples.
+In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above~$0.5$~ns (Fig.~\ref{fig:its-and-ck}a).
 Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.
@@
 For this analysis, we chose five metastable states.
-The CK test (Fig.~\ref{fig:its-and-ck}b) shows a good agreement between reestimation at higher lagtimes (black/solid lines) and higher powers of the original transition matrix (blue/dashed lines).
-Thus, it confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).
+The CK test (Fig.~\ref{fig:its-and-ck}b) shows that predictions from our MSM (blue-dashed lines) agrees well with MSMs estimated with longer lag times (black-solid lines)
+Thus, the CK test confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).
 
@@
 In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening and their timescales.
-The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue $1$.
+The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue~$1$.
 The second right eigenvector, however, corresponds to the slowest process 
@@
 \includegraphics{figure_6}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however, (b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
-The solid/orange lines denote the maximum likelihood MSM result; the dashed/blue lines and the the shaded areas indicate sample means and $95\%$ confidence intervals computed with a Bayesian sampling procedure.}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however,
+(b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
+The solid/orange lines denote the maximum likelihood MSM result;
+the dashed/blue lines and the the shaded areas indicate sample means and~$95\%$ confidence intervals computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.}
 \label{fig:msm-exp-obs}
@@
 As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
-In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original $75$ MSM microstates.
+In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original~$75$ MSM microstates.
 To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
@@
 The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
-To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
+To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
 We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization, making it more likely to be experimentally measurable.

### 2018-09-04 manuscript/manuscript.tex
Added citation on osprey, discuss linear approach
fixes #148

[ci skip]

@@
 
+In the workflow there are multiple hyper parameters to be chosen by the modeler. In our approach we try to optimize a 
+parameter at the current stage of the pipeline and continue to the next stage, once a good choice was found. This 
+requires the researcher to understand the consequences of non optimal deciscions for the final result. For instance
+a non converged clustering could result in lumping states together which should be seperated from each other.
+
+There also exists automatized approaches to optimize all hyper parameters of the pipeline using a cross-validation 
+scheme \cite{husic-optimized}. In these approaches the researcher is still required to understand modeling choices like 
+sane ranges for parameters to avoid wasting computational time, which is spent to explore meaningless areas of the 
+hyperparameter space.
+In the sequential approach, one can fall back to the previous step, if one finds a bad result at any following stage. 
+This greatly reduces the computational effort and leads to a better understanding of the final model.
+
+%However one will not be able to find a good model based on partially bad modeling choices. E.g. a hidden Markov state 
+%model could partially correct bad clusterings, but
+
+\subsection{Feature selection}
+
 \begin{figure}
@@
 
+\subsection{Modeling large systems}
+When estimating MSMs for large systems, challenges may arise that are mostly system dependent.
+
+A case in point is the curse of dimensionality:~it is difficult to discretize a high dimensional feature space.
+While it is somewhat computationally demanding, more importantly, Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001} and thus cluster assignments based on that norm may yield a poor discretization.
+Especially for large systems, it is thus particularly important to first find a suitable set of features, and to further apply dimensionality reduction techniques (e.g.~TICA, VAMP, if applicable) to obtain a low dimensional representation of the slow dynamics.
+Hidden Markov models (HMMs) might further mitigate poor discretization to a certain extent~\cite{noe-proj-hid-msm}.
+
+Furthermore, the slowest process in a system as identified by an MSM or HMM might not be the one a modeler is interested in~\cite{banushkina_nonparametric_2015}.
+For instance, the slowest process might correspond to a biologically irrelevant side chain flip that only occurred once in the data set.
+This problem may be mitigated by choosing a more specific set of features.
+
+Additional technical challenges for large systems include high demands on memory and computation time; we explain how to deal with those in the tutorials.
+
+More details on how to model complex systems with the techniques presented here are described e.g.~by~\cite{plattner_protein_2015,plattner_complete_2017}.
+
 \subsection{Advanced Methods}

### 2018-09-04 manuscript/manuscript.tex

@@
 The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
-
-\subsection{Feature selection}
-
-In the workflow there are multiple hyper parameters to be chosen by the modeler. In our approach we try to optimize a 
-parameter at the current stage of the pipeline and continue to the next stage, once a good choice was found. This 
-requires the researcher to understand the consequences of non optimal deciscions for the final result. For instance
-a non converged clustering could result in lumping states together which should be seperated from each other.
-
-There also exists automatized approaches to optimize all hyper parameters of the pipeline using a cross-validation 
-scheme \cite{husic-optimized}. In these approaches the researcher is still required to understand modeling choices like 
-sane ranges for parameters to avoid wasting computational time, which is spent to explore meaningless areas of the 
-hyperparameter space.
-In the sequential approach, one can fall back to the previous step, if one finds a bad result at any following stage. 
-This greatly reduces the computational effort and leads to a better understanding of the final model.
-
-%However one will not be able to find a good model based on partially bad modeling choices. E.g. a hidden Markov state 
-%model could partially correct bad clusterings, but
 

### 2018-09-06 manuscript/manuscript.tex
review cw [ci skip]
addresses #154

@@
 
-While the present tutorial is intended to cover Markov State Modeling 101, we encourage the user to explore other, more recent extensions of the methodology.
-Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram} can be used to combine unbiased and biased simulations so as to probe kinetics of very rare events~\cite{trammbar}; MEMMs are implemented in PyEMMA.
+The present tutorial presents the basics of modern Markov state modelling with PyEMMA.
+However, recent years have seen many extensions of the methodology -- many of which are available within PyEMMA.
+We encourage interested readers to look into these methods in the software documentation and to check out the Jupyter notebooks distributed with PyEMMA.
+
+Conventional Markov state modelling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties.
+In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram}, we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence.
+MEMMs consequently enable users to combine enhanced sampling methods such as umbrella sampling or replica exchange with conventional molecular dynamics simulations to more efficiently study rare event kinetics~\cite{trammbar}.
+MEMMs are implemented in PyEMMA.
+
+Another issue often faced during Markov state modelling is a lack of quantitative agreement with complementary experimental data.
+This issue is not intrinsic to the Markov state modelling approach as such, but rather associated with systematic errors in the force field model used to conduct the simulation.
+Nevertheless, using Augmented Markov models (AMM) it is possible to build an integrative MSM which balances experimental and simulation data, taking into account their respective uncertainties~\cite{simon-amm}.
+AMMs are implemented in PyEMMA.
+
 Recently, there have been steps towards replacing the traditional user-directed pipeline (involving featurizing, reducing dimension, discretizing, MSM estimation and coarse-graining) by a single end-to-end deep learning method such as VAMPnets~\cite{vampnet}.

### 2018-09-06 manuscript/manuscript.tex
addresses #145 and elements of #136

@@
 
-Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues,
+Once validated, the transition matrix can be decomposed into left-eigenvectors and eigenvalues,
 
 \begin{equation}
-\label{eq:transmat}
-T(\tau) \circ \phi_i = \lambda_i \phi_i,
+\label{eq:spectral_left}
+\phi_i  T(\tau) = \lambda_i \phi_i,
 \end{equation}
 
-\noindent{}where the eigenvalues are indexed in decreasing order.
-The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution of the system.
-The right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
-The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than~$1$ and correspond to dynamical processes within the system.
-The right eigenvectors $\psi_i$ each represent a dynamical process (for $i>1$), and the coefficients of the eigenvectors represent the flux into and out of the MSM states that characterizes that process.
-The corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.
-
-The timescale of a given dynamical process is a function of the relevant eigenvalue and the lag time at which the model was defined,
+\noindent{}or equivalently into their right eigenvectors and eigenvalues,
 
 \begin{equation}
+\label{eq:spectral_right}
+T(\tau)\psi_i = \lambda_i \psi_i,
+\end{equation}
+
+\noindent{}where the eigenvalue-eigenvector pairs are indexed in decreasing order. The eigenvalues are the same in both cases, however, the left and right eigenvectors are related to each other as
+
+\begin{equation}
+\phi_i = \pi \circ \psi_i,
+\label{eq:left-right-eigenvalue-relation}
+\end{equation}
+\noindent{}where $\pi$ is the \emph{stationary distribution} of the MSM, and $\circ$ corresponds to an element-wise vector product.
+
+The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution, $\pi$.
+From the relationship between the left and right eigevenctors (eq.~\ref{eq:left-right-eigenvalue-relation}) we see that the right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
+
+The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} time-scales of dynamical processes within the system, through
+\begin{equation}
 	\label{eq:timescales}
-t_i \equiv -\frac{\tau}{\log(|\lambda_i|)}.
+	t_i \equiv -\frac{\tau}{\log|\lambda_i|}.
 \end{equation}
+
+The right eigenvectors $\psi_i$ each encode a dynamical process (for $i>1$), corresponding to the characteristic time-scale, $t_i$.
+The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterizes that process.
+Again, as can be seen from (eq.~\ref{eq:left-right-eigenvalue-relation}) the corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.
 
@@
 
-In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM as they contain information about what configurational changes are happening and their timescales.
+In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM.
+We here consider the right eigenvectors as they are normalized by the stationary distribution of the MSM.
+This enables us to dissect dynamic changes with finite time-scales from the equilbrium distribution of states.
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue~$1$.
-The second right eigenvector, however, corresponds to the slowest process 
-(the eigenvector components are real because of the detailed balance constraint enforced during MSM estimation).
+The second right eigenvector, on the other hand, corresponds to the slowest process. Note that the eigenvectors are real as detailed balance has been enforced during MSM estimation.
 The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.

### 2018-09-06 manuscript/manuscript.tex
fix notation
updates notations. removes redundant ITS equation.

@@
 \label{eq:spectral_left}
-\phi_i  T(\tau) = \lambda_i \phi_i,
+\phi_i^\top  T(\tau) = \lambda_i(\tau) \phi_i^\top,
 \end{equation}
@@
 \label{eq:spectral_right}
-T(\tau)\psi_i = \lambda_i \psi_i,
+T(\tau)\psi_i = \lambda_i(\tau) \psi_i,
 \end{equation}
@@
 
-The highest eigenvalue, $\lambda_1$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution, $\pi$.
-From the relationship between the left and right eigevenctors (eq.~\ref{eq:left-right-eigenvalue-relation}) we see that the right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
+The highest eigenvalue, $\lambda_1(\tau)$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution, $\pi$.
+From the relationship between the left and right eigenvectors (eq.~\ref{eq:left-right-eigenvalue-relation}) we see that the right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
 
-The subsequent eigenvalues $\lambda_{i>1}$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} time-scales of dynamical processes within the system, through
-\begin{equation}
-	\label{eq:timescales}
-	t_i \equiv -\frac{\tau}{\log|\lambda_i|}.
-\end{equation}
+The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} time-scales of dynamical processes within the system~\ref{eq:its}.
 

### 2018-09-06 manuscript/manuscript.tex

--- 2018-09-06 manuscript/manuscript.tex
@@
 
+Note that the modeler has to select hyper-parameters at most stages throughout the workflow.
+This selection must be done carefully as poor choices make it hard, or even impossible, to build a good MSM.
+
+While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter 
+space, we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized. This 
+approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling 
+choices.
+
 \subsection{Feature selection}
@@
 
-The present tutorial presents the basics of modern Markov state modelling with PyEMMA.
-However, recent years have seen many extensions of the methodology -- many of which are available within PyEMMA.
-We encourage interested readers to look into these methods in the software documentation and to check out the Jupyter notebooks distributed with PyEMMA.
-
-Conventional Markov state modelling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties.
-In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram}, we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence.
-MEMMs consequently enable users to combine enhanced sampling methods such as umbrella sampling or replica exchange with conventional molecular dynamics simulations to more efficiently study rare event kinetics~\cite{trammbar}.
-MEMMs are implemented in PyEMMA.
-
-Another issue often faced during Markov state modelling is a lack of quantitative agreement with complementary experimental data.
-This issue is not intrinsic to the Markov state modelling approach as such, but rather associated with systematic errors in the force field model used to conduct the simulation.
-Nevertheless, using Augmented Markov models (AMM) it is possible to build an integrative MSM which balances experimental and simulation data, taking into account their respective uncertainties~\cite{simon-amm}.
-AMMs are implemented in PyEMMA.
-
+While the present tutorial is intended to cover Markov State Modeling 101, we encourage the user to explore other, more recent extensions of the methodology.
+Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram} can be used to combine unbiased and biased simulations so as to probe kinetics of very rare events~\cite{trammbar}; MEMMs are implemented in PyEMMA.
 Recently, there have been steps towards replacing the traditional user-directed pipeline (involving featurizing, reducing dimension, discretizing, MSM estimation and coarse-graining) by a single end-to-end deep learning method such as VAMPnets~\cite{vampnet}.

### 2018-09-06 manuscript/manuscript.tex
Addresses #144.

--- 2018-09-06 manuscript/manuscript.tex
@@
 (a)~The convergence behavior of the implied timescales associated with the four slowest processes.
-(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 meta-stable states..
+(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 meta-stable states.
 The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
@@
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
-To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at a lag time of~$0.5$~ns.
+To choose among three different molecular features reflecting protein structure, we compute the (cross-validated) VAMP-2 score (Notebook 00).
+Although we cannot optimize lag times with a variational score, such as VAMP-2, it is important to ensure that properties that we optimize are robust as a function of lag time.
+Consequently, we compute the VAMP-2 score at several lag times (Notebook 00).
+We find that the relative rankings of the different molecular features are highly robust as a function of lag time.
+We show one example of this ranking and the absolute VAMP-2 scores for lag time~$0.5$~ns in Fig.~\ref{fig:io-to-tica}b.
 We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).

### 2018-09-06 manuscript/manuscript.tex
cw edits

@@
 
-Once validated, the transition matrix can be decomposed into left-eigenvectors and eigenvalues,
+Once validated, the transition matrix can be decomposed into left eigenvectors and eigenvalues,
 
@@
 
-\noindent{}where the eigenvalue-eigenvector pairs are indexed in decreasing order. The eigenvalues are the same in both cases, however, the left and right eigenvectors are related to each other as
+\noindent{}where the eigenvalue-eigenvector pairs are indexed in decreasing order. 
+The eigenvalues are the same in both cases, however, the left and right eigenvectors are related to each other as
 
@@
 
-The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} time-scales of dynamical processes within the system~\ref{eq:its}.
+The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} time-scales of dynamical processes within the system (eq.~\ref{eq:its}).
 
@@
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue~$1$.
-The second right eigenvector, on the other hand, corresponds to the slowest process. Note that the eigenvectors are real as detailed balance has been enforced during MSM estimation.
+The second right eigenvector, on the other hand, corresponds to the slowest process. 
+Note that the eigenvectors are real as detailed balance has been enforced during MSM estimation.
 The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.

### 2018-09-06 manuscript/manuscript.tex
beh edits

@@
 
-The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} time-scales of dynamical processes within the system (eq.~\ref{eq:its}).
+The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
 
@@
 (a)~The convergence behavior of the implied timescales associated with the four slowest processes.
-(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 meta-stable states.
+(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 metastable states.
 The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
@@
 To choose among three different molecular features reflecting protein structure, we compute the (cross-validated) VAMP-2 score (Notebook 00).
-Although we cannot optimize lag times with a variational score, such as VAMP-2, it is important to ensure that properties that we optimize are robust as a function of lag time. 
+Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2, it is important to ensure that properties that we optimize are robust as a function of lag time.
 Consequently, we compute the VAMP-2 score at several lag times (Notebook 00). 
@@
 In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM.
-We here consider the right eigenvectors as they are normalized by the stationary distribution of the MSM.
-This enables us to dissect dynamic changes with finite time-scales from the equilbrium distribution of states.
+We here chose the right eigenvectors as they are independent of the stationary distribution of the MSM.
+This enables us to specifically study what conformational changes are happening on a particular time scale independently of the equilbrium distribution.
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue~$1$.
-The second right eigenvector, on the other hand, corresponds to the slowest process. 
+The second right eigenvector, on the other hand, corresponds to the slowest process in the system. 
 Note that the eigenvectors are real as detailed balance has been enforced during MSM estimation.
@@
 
-The present tutorial presents the basics of modern Markov state modelling with PyEMMA. 
-However, recent years have seen many extensions of the methodology -- many of which are available within PyEMMA. 
-We encourage interested readers to look into these methods in the software documentation and to check out the Jupyter notebooks distributed with PyEMMA.
+The present tutorial presents the basics of modern Markov state modeling with PyEMMA. 
+However, recent years have seen many extensions of the methodology --- many of which are available within PyEMMA. 
+We encourage interested readers to look into these methods in the software documentation and to make use of the specific Jupyter notebooks distributed with PyEMMA.
 
-Conventional Markov state modelling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties. 
+Conventional Markov state modeling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties. 
 In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram}, we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence. 
@@
 
-Another issue often faced during Markov state modelling is a lack of quantitative agreement with complementary experimental data. 
-This issue is not intrinsic to the Markov state modelling approach as such, but rather associated with systematic errors in the force field model used to conduct the simulation. 
+Another issue often faced during Markov state modeling is a lack of quantitative agreement with complementary experimental data. 
+This issue is not intrinsic to the Markov state modeling approach as such, but rather associated with systematic errors in the force field model used to conduct the simulation. 
 Nevertheless, using Augmented Markov models (AMM) it is possible to build an integrative MSM which balances experimental and simulation data, taking into account their respective uncertainties~\cite{simon-amm}.

### 2018-09-06 manuscript/manuscript.tex
cw edit2

@@
 
-In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM.
-We here chose the right eigenvectors as they are independent of the stationary distribution of the MSM.
+In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM, as they are independent of the stationary distribution of the MSM.
 This enables us to specifically study what conformational changes are happening on a particular time scale independently of the equilbrium distribution.

### 2018-09-06 manuscript/manuscript.tex
cw fix

@@
 
-In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors of the MSM, as they are independent of the stationary distribution of the MSM.
+In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors, as they are independent of the stationary distribution.
 This enables us to specifically study what conformational changes are happening on a particular time scale independently of the equilbrium distribution.

### 2018-09-06 manuscript/manuscript.tex

@@
 The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
+
+Note that the modeler has to select hyper-parameters at most stages throughout the workflow.
+This selection must be done carefully as poor choices make it hard, or even impossible, to build a good MSM.
+
+While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter 
+space, we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized. This 
+approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling 
+choices.
 

### 2018-09-13 manuscript/manuscript.tex
requested text changes, line breaks, comments removed (#136)

@@
 \graphicspath{{figures/}}
+\usepackage{bm}
 
@@
 
-In this tutorial, we assume the reader's familiarity with the basic theory behind the MSM approach (see Sec.~\ref{sec:background}) and focus on usage of PyEMMA.
-Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
-We also assume the reader is familiar with MD analyses of proteins and peptides and commonly used structural components of these systems (for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}).
+In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances.
+(see [7] for a review).
+
+We further assume that the reader is familiar with the basic ideas and theory underlying Markov modeling and will only give a brief reminder of the basic concepts in Section 2.
+
+For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview,
+while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
+Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
+
+In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
+we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
+transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
+hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial},
+and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
+as these topics play important roles within the standard MSM workflow.
 
@@
 
-\subsection{Essential theory}
+\subsection{Markov state models}
 \label{sec:theory}
@@
 Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
-In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
-The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ after a time $\tau$ is independent of the history of the system before it was in $i$.
-%This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.
+In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states,
+and a transition matrix $P(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
+Let us make two remarks to avoid common misconceptions:
+\begin{enumerate}
+\item Equilibrium:
+While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
+Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
+they are useful to analyze short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
+\item Markovianity:
+An MSM is a memoryless model.
+Early MSM papers have argued that accurate MSMs can be found if a few states with high barriers are captured by the MSM states so as to achieve a Mori-Zwanzig projection with fast-decaying memory~\cite{swope-its,noe2007jcp,chodera2007jcp}.
+The modern view, however, is that MSMs can be highly accurate if the MSM states discretize the collective coordinates of the slowest processes well~\cite{msm-jhp}.
+This mainly requires that the system is characterized by only a few slow processes at lag time $\tau$,
+which is true for cooperative systems such as most proteins, but not for highly frustrated systems such as glasses.
+\end{enumerate}
 
@@
 Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
-Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
-To ensure that the transition probability matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,
-
+Then, the count matrix is converted to a row-stochastic transition probability matrix $P$, which is defined for the specified lag time.
+For MD simulations in equilibrium, $P$ should obey detailed balance which is enforced by constraining the estimation of $P$ to the following equations:
 \begin{equation}
 \label{eq:balance}
-p_t(i) p(i \rightarrow j) = p_t(j)p(j \rightarrow i),
+\pi_i P_{ij} = \pi_j P_{ji},
 \end{equation}
-
-\noindent{}where $p_t(i)$ represents the probability of being in state $i$ at time $t$, and $p(i \rightarrow j)$ is the probability of transitioning from state $i$ to $j$ at the next time step.
-% Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
-The requirement of detailed balance indicates that we assume our system is reversible.
-We additionally require that the system is ergodic, which means that every state is accessible from every other state.
+where $\pi_i$ is the stationary probability of state $i$ and $P_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
+The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
+for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
+For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.
 
@@
 \end{equation}
+When the ITS become approximately constant with the lag time, we say that our timescales have converged and choose the smallest lag time with the converged timescales in order to maximize the model's temporal resolution.
 
-\noindent{}When the ITS become approximately constant with the lag time, we say that our timescales have converged and choose the smallest lag time with the converged timescales in order to maximize the model's temporal resolution.
-
-Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{msm-jhp}.
+Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
 The CK property for a Markovian matrix is,
-
 \begin{equation}
@@
 \end{equation}
-
-\noindent{}where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
+where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
 By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
 
-Once validated, the transition matrix can be decomposed into left eigenvectors and eigenvalues,
-
+Once validated, the transition matrix can be decomposed into eigenvectors and eigenvalues.
+The highest eigenvalue, $\lambda_1(\tau)$, is unique and equal to $1$.
+Its corresponding left eigenvector is the stationary distribution, $\bm{\pi}$:
 \begin{equation}
-\label{eq:spectral_left}
-\phi_i^\top  T(\tau) = \lambda_i(\tau) \phi_i^\top,
+\bm{\pi}^\top  P(\tau) = \bm{\pi}^\top.
 \end{equation}
 
-\noindent{}or equivalently into their right eigenvectors and eigenvalues,
+The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
+The dynamical process themself (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
+\begin{equation}
+P(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
+\end{equation}
+where the eigenvalue-eigenvector pairs are indexed in decreasing order.
+The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
+The right eigenvector $\bm{\psi}_1$ is a vector consisting of~$1$'s.
 
-\begin{equation}
-\label{eq:spectral_right}
-T(\tau)\psi_i = \lambda_i(\tau) \psi_i,
-\end{equation}
-
-\noindent{}where the eigenvalue-eigenvector pairs are indexed in decreasing order. 
-The eigenvalues are the same in both cases, however, the left and right eigenvectors are related to each other as
-
-\begin{equation}
-\phi_i = \pi \circ \psi_i,
-\label{eq:left-right-eigenvalue-relation}
-\end{equation}
-\noindent{}where $\pi$ is the \emph{stationary distribution} of the MSM, and $\circ$ corresponds to an element-wise vector product.
-
-The highest eigenvalue, $\lambda_1(\tau)$, is unique and is equal to $1$, and its corresponding left eigenvector $\phi_1$ corresponds to the stationary distribution, $\pi$.
-From the relationship between the left and right eigenvectors (eq.~\ref{eq:left-right-eigenvalue-relation}) we see that the right eigenvector $\psi_1$ is a vector consisting of~$1$'s.
-
-The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
-
-The right eigenvectors $\psi_i$ each encode a dynamical process (for $i>1$), corresponding to the characteristic time-scale, $t_i$.
-The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterizes that process.
-Again, as can be seen from (eq.~\ref{eq:left-right-eigenvalue-relation}) the corresponding left eigenvectors $\phi_i$ contain the same information weighted by the stationary distribution.
-
-\subsection{MSM construction the variational approach}
-\label{sec:construction}
+\subsection{Variational approach and TICA}
 
@@
 It is apparent that there are many choices involved in MSM construction such as what features should be used and how many states should be chosen.
-In 2013, the variational approach to conformational dynamics (VAC) was derived, which enabled an objective comparison among different state decomposition choices for models built with the same Markovian lag time~\cite{noe-vac}.
-More recently, the more general variational approach to Markov processes (VAMP) has been developed in order to facilitate the approximation and comparison of reversible models for basis sets that are continuous, as opposed to discrete states~\cite{vamp-preprint}.
+In 2013, the variational approach to conformational dynamics (VAC) was derived,
+which enabled an objective comparison among different state decomposition choices for models built with the same Markovian lag time~\cite{noe-vac}.
+More recently, the more general variational approach to Markov processes (VAMP) has been developed in order to facilitate the approximation and comparison of reversible models for basis sets that are continuous,
+as opposed to discrete states~\cite{vamp-preprint}.
 The VAMP can thus be used to perform model selection.
 Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model.
-However, the MSM lag time cannot be optimized using VAMP, and must be chosen using a separate validation as described above~\cite{husic2017note}.
+However, the MSM lag time cannot be optimized using VAMP,
+and must be chosen using a separate validation as described above~\cite{husic2017note}.
 
-\subsection{Background knowledge and resources}
-\label{sec:background}
-
-For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
-Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
-
-In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
-we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
-hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.
-
-\subsection{Software/system requirements}
+\subsection{Software and installation}
 
@@
 A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
-By iterating between data processing and MSM estimation/validation, a dynamical model is obtained that can be analyzed (last row).}
+By iterating between data processing and MSM estimation/validation,
+a dynamical model is obtained that can be analyzed (last row).}
 \label{fig:workflowchart}
@@
 
-\section{Content and links}
+\section{PyEMMA tutorials}
 
 This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
-The first notebook (00), which we will summarize in the following, showcases the entire estimation, validation, and analysis workflow for a small example system.
+The first notebook (00), which we will summarize in the following, showcases the entire estimation,
+validation, and analysis workflow for a small example system.
 The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
-The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics, and the last notebook (08) contains guidelines on how to deal with common problems during MSM estimation.
+The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics,
+and the last notebook (08) contains guidelines on how to deal with common problems during MSM estimation.
 
@@
 For the remainder of this manuscript we will walk through the first notebook (00).
-In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a), consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
-We present the results obtained in this notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
+In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
+consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
+We present the results obtained in this notebook,
+thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
 The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
@@
 The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
-The black line indicates where implied timescales are equal to the lag time, whereas the grey area indicates all implied timescales faster than the lag time.
+The black line indicates where implied timescales are equal to the lag time,
+whereas the grey area indicates all implied timescales faster than the lag time.
 In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}
@@
 In Markov state modeling our objective is to model the slow dynamics of a molecular process.
-In order to approximate the slow dynamics in a statistically efficient manner, a lower dimensional representation of our simulation data is necessary.
+In order to approximate the slow dynamics in a statistically efficient manner,
+a lower dimensional representation of our simulation data is necessary.
 However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system are unknown a priori~\cite{NoeClementiReview}.
-Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk} and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint} provide a systematic means to quantitatively compare multiple representations of the simulation data.
+Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
+and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
+provide a systematic means to quantitatively compare multiple representations of the simulation data.
 In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
@@
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
-To choose among three different molecular features reflecting protein structure, we compute the (cross-validated) VAMP-2 score (Notebook 00).
-Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2, it is important to ensure that properties that we optimize are robust as a function of lag time.
+To choose among three different molecular features reflecting protein structure,
+we compute the (cross-validated) VAMP-2 score (Notebook 00).
+Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
+it is important to ensure that properties that we optimize are robust as a function of lag time. 
 Consequently, we compute the VAMP-2 score at several lag times (Notebook 00). 
@@
 
-Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space, which typically contains many degrees of freedom, to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
+Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space,
+which typically contains many degrees of freedom,
+to a lower dimensional space that can be discretized with higher resolution and better statistical efficiency.
 TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
-Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
+Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff
+(note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
 The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-tica}c).
@@
 \subsection{Discretization}
-TICA yields a representation of our molecular simulation data with a reduced dimensionality, which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
+
+TICA yields a representation of our molecular simulation data with a reduced dimensionality,
+which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation.
 Here, we use the $k$-means algorithm to segment the four dimensional TICA space into $k=75$ cluster centers.
-The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above, which is shown in the showcase notebook (00).
+The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
+which is shown in the showcase notebook (00).
 
 \subsection{MSM estimation and validation}
-% When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
-% Plotting the implied timescales (ITS) as a function of~$\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
-The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
-% \begin{equation}
-% \label{eq:its}
-% t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
-% \end{equation}
-A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
+
+A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$;
+accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty.
 The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with~$100$ samples.
@@
 
-To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test.
-% The CK test compares the right and the left side of the Chapman-Kolmogorov equation
-% \begin{equation}
-% \label{eq:ck}
-% T(k \tau) = T^k(\tau)
-% \end{equation}
-% where $T$ is the MSM transition matrix.
-The left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power.
-Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
+To test the validity of our MSM, we perform a Chapman-Kolmogorov (CK) test.
+Visualizing the full transition probability matrix $T$ is difficult;
+we therefore coarse-grain $T$ into a smaller number of metastable states before performing the test.
 An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot.
 For this analysis, we chose five metastable states.
-The CK test (Fig.~\ref{fig:its-and-ck}b) shows that predictions from our MSM (blue-dashed lines) agrees well with MSMs estimated with longer lag times (black-solid lines)
-Thus, the CK test confirms that five metastable states is an appropriate choice and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system within error (blue/shaded area).
+The CK test (Fig.~\ref{fig:its-and-ck}b) shows that predictions from our MSM (blue-dashed lines)
+agrees well with MSMs estimated with longer lag times (black-solid lines)
+Thus, the CK test confirms that five metastable states is an appropriate choice
+and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the
+long-timescale behavior of our system within error (blue/shaded area).
 
@@
 (b)~PCCA++ identifies as five metastable states.
-(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states, in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
+(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
+in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
 (d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
@@
 We can now directly extract several thermodynamic and kinetic properties from the estimated and validated model.
-An example of the former is the free energy surface in the projection onto the first two TICA components (Fig.~\ref{fig:msm-analysis}a) reweighted by the MSM stationary distribution.
+An example of the former is the free energy surface in the projection onto the first two TICA components
+(Fig.~\ref{fig:msm-analysis}a) reweighted by the MSM stationary distribution.
 
-A spectral clustering using the PCCA++ algorithm~\cite{pcca++,Deuflhard2005-pcca,Kube2007-pcca+} allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
+A spectral clustering using the PCCA++ algorithm~\cite{pcca++,Deuflhard2005-pcca,Kube2007-pcca+}
+allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates
+(Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$,
+for which we then approximate the stationary probabilities and relative free energies
+(defined up to an additive constant)
 \[ \begin{array}{ccc}
@@
 
-In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors, as they are independent of the stationary distribution.
+In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors,
+as they are independent of the stationary distribution.
 This enables us to specifically study what conformational changes are happening on a particular time scale independently of the equilbrium distribution.
@@
 The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.
-The relaxation timescale of this exchange process is exactly the corresponding implied timescale, which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
-In the projection onto the first two TICA components, we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system, with macrostates $\mathcal{S}_4$ and $\mathcal{S}_5$ in particular (Fig.~\ref{fig:msm-analysis}c).
+The relaxation timescale of this exchange process is exactly the corresponding implied timescale,
+which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
+In the projection onto the first two TICA components,
+we identify the slowest MSM process as a probability shift between macrostate $\mathcal{S}_1$ and the rest of the system,
+with macrostates $\mathcal{S}_4$ and $\mathcal{S}_5$ in particular (Fig.~\ref{fig:msm-analysis}c).
 
@@
 Connecting MSM analysis to experimental data can both serve as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
-Since we have both the stationary and dynamic properties of the molecular system encoded in the MSM transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.
+Since we have both the stationary and dynamic properties of the molecular system encoded in the MSM transition probability matrix,
+we can compute observables that involve both stationary ensemble averages as well as correlation functions.
 
-As an example, here we look at the fluorescence correlation of Trp-1, since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
-In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each of the original~$75$ MSM microstates.
-To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
-Now that we have an approximation of the fluorescence in each of our MSM states, we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-exp-obs}a).
+As an example, here we look at the fluorescence correlation of Trp-1,
+since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
+In order to compute the fluorescence correlation functions we require a microscopic,
+instantaneous value of the tryptophan fluorescence for each of the original~$75$ MSM microstates.
+To approximate the fluorescence signal in our pentapeptide system,
+we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
+Now that we have an approximation of the fluorescence in each of our MSM states,
+we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-exp-obs}a).
 Note how the computed ACF has a very small response (i.e., signal amplitude).
@@
 The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
-To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
-We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization, making it more likely to be experimentally measurable.
+To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$
+and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
+We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization,
+making it more likely to be experimentally measurable.
 
@@
 \subsection{Modeling large systems}
+
 When estimating MSMs for large systems, challenges may arise that are mostly system dependent.
@@
 A case in point is the curse of dimensionality:~it is difficult to discretize a high dimensional feature space.
-While it is somewhat computationally demanding, more importantly, Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001} and thus cluster assignments based on that norm may yield a poor discretization.
-Especially for large systems, it is thus particularly important to first find a suitable set of features, and to further apply dimensionality reduction techniques (e.g.~TICA, VAMP, if applicable) to obtain a low dimensional representation of the slow dynamics.
+While it is somewhat computationally demanding, more importantly,
+Euclidean distances become less meaningful with increasing dimensionality~\cite{aggarwal_surprising_2001}
+and thus cluster assignments based on that norm may yield a poor discretization.
+Especially for large systems, it is thus particularly important to first find a suitable set of features,
+and to further apply dimensionality reduction techniques (e.g., TICA, VAMP, if applicable)
+to obtain a low dimensional representation of the slow dynamics.
 Hidden Markov models (HMMs) might further mitigate poor discretization to a certain extent~\cite{noe-proj-hid-msm}.
@@
 
-Additional technical challenges for large systems include high demands on memory and computation time; we explain how to deal with those in the tutorials.
+Additional technical challenges for large systems include high demands on memory and computation time;
+we explain how to deal with those in the tutorials.
 
@@
 The present tutorial presents the basics of modern Markov state modeling with PyEMMA. 
-However, recent years have seen many extensions of the methodology --- many of which are available within PyEMMA. 
+However, recent years have seen many extensions of the methodology---many of which are available within PyEMMA. 
 We encourage interested readers to look into these methods in the software documentation and to make use of the specific Jupyter notebooks distributed with PyEMMA.
@@
 Conventional Markov state modeling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties. 
-In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram}, we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence. 
-MEMMs consequently enable users to combine enhanced sampling methods such as umbrella sampling or replica exchange with conventional molecular dynamics simulations to more efficiently study rare event kinetics~\cite{trammbar}.
+In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram},
+we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence. 
+MEMMs consequently enable users to combine enhanced sampling methods such as umbrella sampling or replica exchange
+with conventional molecular dynamics simulations to more efficiently study rare event kinetics~\cite{trammbar}. 
 MEMMs are implemented in PyEMMA.
@@
 Another issue often faced during Markov state modeling is a lack of quantitative agreement with complementary experimental data. 
-This issue is not intrinsic to the Markov state modeling approach as such, but rather associated with systematic errors in the force field model used to conduct the simulation. 
-Nevertheless, using Augmented Markov models (AMM) it is possible to build an integrative MSM which balances experimental and simulation data, taking into account their respective uncertainties~\cite{simon-amm}.
+This issue is not intrinsic to the Markov state modeling approach as such,
+but rather associated with systematic errors in the force field model used to conduct the simulation. 
+Nevertheless, using Augmented Markov models (AMM) it is possible to build an integrative MSM which balances experimental and simulation data,
+taking into account their respective uncertainties~\cite{simon-amm}. 
 AMMs are implemented in PyEMMA.
 
-Recently, there have been steps towards replacing the traditional user-directed pipeline (involving featurizing, reducing dimension, discretizing, MSM estimation and coarse-graining) by a single end-to-end deep learning method such as VAMPnets~\cite{vampnet}.
-Other deep learning methods for performing the dimension reduction~\cite{tae}, finding reaction coordinates for enhanced sampling~\cite{hernandez-vde,Sultan2018-vde-enhanced-sampling,Ribeiro2018-rave}, and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research on its own right.
+Recently, there have been steps towards replacing the traditional user-directed pipeline (involving featurizing,
+reducing dimension, discretizing, MSM estimation and coarse-graining) by a single end-to-end deep learning method such as VAMPnets~\cite{vampnet}.
+Other deep learning methods for performing the dimension reduction~\cite{tae},
+finding reaction coordinates for enhanced sampling~\cite{hernandez-vde,Sultan2018-vde-enhanced-sampling,Ribeiro2018-rave},
+and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research on its own right.
 Implementations of some of these methods are available or are under development in the deeptime package \url{github.com/markovmodel/deeptime}.

### 2018-09-13 manuscript/manuscript.tex
updated article tex files

@@
 % Use the 'lineno' option for adding line numbers.
+% Use the 'pubversion' option for adding the citation and publication information to the document footer.
 % The 'bestpractices' option for indicates that this is a best practices guide.
@@
 
-\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository{}; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
+\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% PUBLICATION INFORMATION
+%%% Fill out these parameters when available
+%%% These are used when the "pubversion" option is invoked
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\pubDOI{10.XXXX/YYYYYYY}
+\pubvolume{<volume>}
+\pubyear{<year>}
+\articlenum{<number>}
+\datereceived{July, 27, 2018}
+\dateaccepted{Month, Day, Year}
 

### 2018-09-13 manuscript/manuscript.tex
update to figure 2 (#136)

@@
 (b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The sample density projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple density maxima and
-(d)~the time series of the first two ICs of the first trajectory.}
+(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
+(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
 \label{fig:io-to-tica}
@@
 
-In Markov state modeling our objective is to model the slow dynamics of a molecular process.
+In Markov state modeling, our objective is to model the slow dynamics of a molecular process.
 In order to approximate the slow dynamics in a statistically efficient manner,
@@
 (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
-The sample density projected onto the first two independent components (ICs) exhibits several maxima (Fig.~\ref{fig:io-to-tica}c).
-Discrete jumps between the maxima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
+The sample free energy projected onto the first two independent components (ICs) exhibits several minima (Fig.~\ref{fig:io-to-tica}c).
+Discrete jumps between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
 We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.

### 2018-09-13 manuscript/manuscript.tex
minor changes

@@
 
-\newcommand{\versionnumber}{0.3}
+\newcommand{\versionnumber}{1.0}
 \newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}
@@
 \label{eq:balance}
-\pi_i P_{ij} = \pi_j P_{ji},
+\pi_i p_{ij} = \pi_j p_{ji},
 \end{equation}
-where $\pi_i$ is the stationary probability of state $i$ and $P_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
+where $\pi_i$ is the stationary probability of state $i$ and $p_{ij}$ is the probability of transitioning to state $j$ conditional on being in state $i$.
 The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
@@
 \begin{equation}
-T(k \tau) = T^k(\tau),
+P(k \tau) = P^k(\tau),
 \end{equation}
-where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~1, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
+where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
 By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
@@
 
-While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter 
-space, we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized. This 
-approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling 
-choices.
+While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter space,
+we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized.
+This approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling choices.
 
@@
 \includegraphics{figure_2}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone: (a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
 (b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
@@
 
-More details on how to model complex systems with the techniques presented here are described e.g.~by~\cite{plattner_protein_2015,plattner_complete_2017}.
+More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
 

### 2018-09-13 manuscript/manuscript.tex
minor changes

@@
 \item mdshare -- A downloader for MD data from a public server
-\item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\-\_contrib\_nbextensions and nbexamples
+\item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
 \item matplotlib -- A plotting library~\cite{matplotlib}
@@
 
-For a more detailed description of author contributions, see the GitHub issue tracking and changelog at\\\githubrepository{}.
+For a more detailed description of author contributions, see the GitHub issue tracking and changelog at \githubrepository{}.
 

### 2018-09-13 manuscript/manuscript.tex
adding TICA text (#136)

@@
 In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states,
-and a transition matrix $P(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
+and a transition matrix $\mathbf{P}(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
 Let us make two remarks to avoid common misconceptions:
@@
 Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
-Then, the count matrix is converted to a row-stochastic transition probability matrix $P$, which is defined for the specified lag time.
-For MD simulations in equilibrium, $P$ should obey detailed balance which is enforced by constraining the estimation of $P$ to the following equations:
+Then, the count matrix is converted to a row-stochastic transition probability matrix $\mathbf{P}$, which is defined for the specified lag time.
+For MD simulations in equilibrium, $\mathbf{P}$ should obey detailed balance which is enforced by constraining the estimation of $\mathbf{P}$ to the following equations:
 \begin{equation}
@@
 
-Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $T(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
+Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $\mathbf{P}(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
 The CK property for a Markovian matrix is,
 \begin{equation}
-P(k \tau) = P^k(\tau),
+\mathbf{P}(k \tau) = \mathbf{P}^k(\tau),
 \end{equation}
@@
 \begin{equation}
-\bm{\pi}^\top  P(\tau) = \bm{\pi}^\top.
+\bm{\pi}^\top  \mathbf{P}(\tau) = \bm{\pi}^\top.
 \end{equation}
@@
 \begin{equation}
-P(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
+\mathbf{P}(\tau)\bm{\psi}_i = \lambda_i(\tau) \bm{\psi}_i,
 \end{equation}
@@
 and must be chosen using a separate validation as described above~\cite{husic2017note}.
+
+Our recommended method for dimensionality reduction, TICA, is a particular implementation of the VAC.
+To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
+\begin{eqnarray}
+c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \\
+c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t,
+\end{eqnarray}
+where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
+Then, we can solve the generalized eigenvalue problem
+\begin{equation}
+\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i
+\end{equation}
+to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
+where the pairs of eigenvalues and independent components are in descending order.
+
+Dimensionality reduction is achieved by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
+onto the leading $d$ independent components $\mathbf{U}=[\mathbf{u}_1 \dots \mathbf{u}_d]$,
+\begin{equation}
+\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
+\end{equation}
+while retaining the kinetic variance
+\begin{equation}
+\textrm{KV}_d = \sum\limits_{i=1}^d \lambda_i^2(\tau);
+\end{equation}
+the total kinetic variance is the sum of the squares of all eigenvalues.
 

### 2018-09-17 manuscript/manuscript.tex
addressing @brookehus' change requests

@@
 In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances.
-(see [7] for a review).
+(see~\cite{dror2012biomolecular} for a review).
 
@@
 while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
-Additionally, two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
+Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
 
-In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
+In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
 we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
@@
 Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
-In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states,
+In its standard formulation, the creation of an MSM involves decomposing the phase or configuration space occupied by a dynamical system into a set of disjoint, discrete states,
 and a transition matrix $\mathbf{P}(\tau) = [p_{ij}(\tau)]$ denoting the conditional probability of finding the system in state $j$ at time $t+\tau$ given that it was in state $i$ at time $t$.
@@
 Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
-they are useful to analyze short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
+they are useful for the analysis of short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
 \item Markovianity:
@@
 In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors,
-as they are independent of the stationary distribution.
-This enables us to specifically study what conformational changes are happening on a particular time scale independently of the equilbrium distribution.
+as they are independent of the stationary distribution (see Section~\ref{sec:theory)}.
+This enables us to specifically study what conformational changes are happening on a particular timescale independently of the equilbrium distribution.
 The first right eigenvector corresponds to the stationary process and its eigenvalue is the Perron eigenvalue~$1$.
@@
 To approximate the fluorescence signal in our pentapeptide system,
-we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.
+we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA)~\cite{sasa-calculation} of Trp-1.
 Now that we have an approximation of the fluorescence in each of our MSM states,
@@
 This issue is not intrinsic to the Markov state modeling approach as such,
-but rather associated with systematic errors in the force field model used to conduct the simulation. 
+but rather is associated with systematic errors in the force field model used to conduct the simulation. 
 Nevertheless, using Augmented Markov models (AMM) it is possible to build an integrative MSM which balances experimental and simulation data,
@@
 finding reaction coordinates for enhanced sampling~\cite{hernandez-vde,Sultan2018-vde-enhanced-sampling,Ribeiro2018-rave},
-and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research on its own right.
+and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research in its own right.
 Implementations of some of these methods are available or are under development in the deeptime package \url{github.com/markovmodel/deeptime}.

### 2018-09-18 manuscript/manuscript.tex
fixing section reference

@@
 In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors,
-as they are independent of the stationary distribution (see Section~\ref{sec:theory)}.
+as they are independent of the stationary distribution (see Section~\ref{sec:theory}).
 This enables us to specifically study what conformational changes are happening on a particular timescale independently of the equilbrium distribution.

### 2018-09-18 manuscript/manuscript.tex
added HMM subsection

@@
 transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
-hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial},
+hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint},
 and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
@@
 the total kinetic variance is the sum of the squares of all eigenvalues.
+
+\subsection*{Hidden Markov state models}
+
+The estimation of an MSM requires the dynamics between microstates to be Markovian.
+However, in case of a poor discretization or short trajectories,
+it might be hard or even impossible to meet this requirement (as we will see in Notebook~07).
+
+An alternative, which is much less sensitive to poor discretization,
+is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
+HMMs require the dynamics in some hidden (unobserved) state space to be Markovian.
+
+To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
+in practice, a timescale separation $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
+Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
+\begin{equation}
+\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
+\end{equation}
+Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
+$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m$ hidden states' stationary probabilities,
+$\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
+and $\bm{\chi}$ is an $m\times n$-dimensional row stochastic matrix
+where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.
+
+An HMM estimation usually yields a small model where each hidden state is considered to be metastable and,
+thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully.
+Given that an HMM is a generalization of an MSM,
+we can compute all physico-chemically relevant quantities available for MSMs, too.
 

### 2018-09-18 manuscript/manuscript.tex
better figure placing and hyphenation

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \title{Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \pubDOI{10.XXXX/YYYYYYY}
@@
 $\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
-and $\bm{\chi}$ is an $m\times n$-dimensional row stochastic matrix
+and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
 where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.
@@
 
-While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter space,
-we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized.
+While there exist automated schemes~\cite{husic-optimized} for cross-val\-\-i\-dat\-ed optimization in the full hyper-parameter space,
+we chose to adopt a sequential approach where only the hyper-pa\-ra\-me\-ters of the current stage are optimized.
 This approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling choices.
-
-\subsection{Feature selection}
 
@@
 
-\begin{figure}
-\includegraphics{figure_3}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The convergence behavior of the implied timescales associated with the four slowest processes.
-(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 metastable states.
-The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
-The black line indicates where implied timescales are equal to the lag time,
-whereas the grey area indicates all implied timescales faster than the lag time.
-In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}
-\label{fig:its-and-ck}
-\end{figure}
+\subsection{Feature selection}
 
@@
 
+\begin{figure}
+\includegraphics{figure_3}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The convergence behavior of the implied timescales associated with the four slowest processes.
+(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 metastable states.
+The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
+The black line indicates where implied timescales are equal to the lag time,
+whereas the grey area indicates all implied timescales faster than the lag time.
+In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}
+\label{fig:its-and-ck}
+\end{figure}
+
 A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$;
@@
 \subsection{Analyzing the MSM}
-
-\begin{figure}
-\includegraphics{figure_4}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
-(b)~PCCA++ identifies as five metastable states.
-(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
-in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
-(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
-\label{fig:msm-analysis}
-\end{figure}
 
@@
 \begin{figure}
+\includegraphics{figure_4}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
+(b)~PCCA++ identifies as five metastable states.
+(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
+in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
+(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
+\label{fig:msm-analysis}
+\end{figure}
+
+\begin{figure}
 \includegraphics{figure_5}
@@
 (a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however,
-(b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
+(b)~can be enhanced if the system is prepared in the non\-equi\-lib\-ri\-um condition $\mathcal{S}_1$.
 The solid/orange lines denote the maximum likelihood MSM result;

### 2018-09-18 manuscript/manuscript.tex
fixed subsection numbering [ci skip]
addressing @brookehus' comments

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\hyphenation{val-i-dat-ed}
 
@@
 
-\subsection*{Hidden Markov state models}
+\subsection{Hidden Markov state models}
 
@@
 is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
-HMMs require the dynamics in some hidden (unobserved) state space to be Markovian.
+HMMs require the dynamics in some hidden (i.e., unobserved) state space to be Markovian.
 
 To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
-in practice, a timescale separation $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
+in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
 Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
@@
 
-An HMM estimation usually yields a small model where each hidden state is considered to be metastable and,
-thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully.
+An HMM estimation usually yields a model with a small number of (hidden) states
+where each state is considered to be metastable and,
+thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
 Given that an HMM is a generalization of an MSM,
@@
 \subsection{Software and installation}
-
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{figure_1}
-\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
-A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
-By iterating between data processing and MSM estimation/validation,
-a dynamical model is obtained that can be analyzed (last row).}
-\label{fig:workflowchart}
-\end{figure}
 
@@
 
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{figure_1}
+\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
+A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
+By iterating between data processing and MSM estimation/validation,
+a dynamical model is obtained that can be analyzed (last row).}
+\label{fig:workflowchart}
+\end{figure}
+
 \section{PyEMMA tutorials}
@@
 
-Note that the modeler has to select hyper-parameters at most stages throughout the workflow.
-This selection must be done carefully as poor choices make it hard, or even impossible, to build a good MSM.
-
-While there exist automated schemes~\cite{husic-optimized} for cross-val\-\-i\-dat\-ed optimization in the full hyper-parameter space,
-we chose to adopt a sequential approach where only the hyper-pa\-ra\-me\-ters of the current stage are optimized.
-This approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling choices.
-
 \begin{figure}
@@
 \end{figure}
+
+Note that the modeler has to select hyper-parameters at most stages throughout the workflow.
+This selection must be done carefully as poor choices make it hard, or even impossible, to build a good MSM.
+
+While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter space,
+we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized.
+This approach is not only computationally cheaper but allows us to discuss the significance of the necessary modeling choices.
 
@@
 
+\begin{figure}
+\includegraphics{figure_4}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
+(b)~PCCA++ identifies as five metastable states.
+(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
+in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
+(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
+\label{fig:msm-analysis}
+\end{figure}
+
 The mean first passage times (MFPTs) out of and into the macrostate $\mathcal{S}_1$ compute to
@@
 \begin{figure}
-\includegraphics{figure_4}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
-(b)~PCCA++ identifies as five metastable states.
-(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
-in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
-(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
-\label{fig:msm-analysis}
-\end{figure}
-
-\begin{figure}
 \includegraphics{figure_5}
@@
 (a)~the Trp-1 SASA autocorrelation function yields a weak signal which, however,
-(b)~can be enhanced if the system is prepared in the non\-equi\-lib\-ri\-um condition $\mathcal{S}_1$.
+(b)~can be enhanced if the system is prepared in the nonequilibrium condition $\mathcal{S}_1$.
 The solid/orange lines denote the maximum likelihood MSM result;

### 2018-09-19 manuscript/manuscript.tex
go back to in-place hyphenation

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\hyphenation{val-i-dat-ed}
 
@@
 
-While there exist automated schemes~\cite{husic-optimized} for cross-validated optimization in the full hyper-parameter space,
+While there exist automated schemes~\cite{husic-optimized} for cross-val\-i\-dat\-ed optimization in the full hyper-parameter space,
 we chose to adopt a sequential approach where only the hyper-parameters of the current stage are optimized.

### 2018-09-20 manuscript/manuscript.tex
addressing @psolsson's comments [ci skip]

@@
 The estimation of an MSM requires the dynamics between microstates to be Markovian.
-However, in case of a poor discretization or short trajectories,
-it might be hard or even impossible to meet this requirement (as we will see in Notebook~07).
+However, in case of a poor dimension reduction and/or discretization or short trajectories,
+we cannot anticipate this to be the case.
+We illustrate this point in Notebook~07.
 
@@
 is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
-HMMs require the dynamics in some hidden (i.e., unobserved) state space to be Markovian.
+HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space.
+Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
+and gives rise to our observed data, e.g., the discretized trajectories.
+This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
+our molecular dynamics trajectories.
 
@@
 Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
-$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m$ hidden states' stationary probabilities,
-$\tilde{\mathbf{P}}(\tau)$ is a (small) transition matrix between the $m$ hidden states at lag time $\tau$,
+$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m<n$ hidden states' stationary probabilities,
+$\tilde{\mathbf{P}}(\tau)$ is a transition matrix between the $m<n$ hidden states at lag time $\tau$,
 and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
@@
 
-An HMM estimation usually yields a model with a small number of (hidden) states
+An HMM estimation always yields a model with a small number of (hidden) states
 where each state is considered to be metastable and,
 thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
-Given that an HMM is a generalization of an MSM,
-we can compute all physico-chemically relevant quantities available for MSMs, too.
+As the HMMs---like MSMs---approximate the full phase-space dynamics,
+we can similarly compute physical observables from directly.
 

### 2018-09-20 manuscript/manuscript.tex
incrementing figure numbers
adding new figure 1 (HMM scheme)

@@
 
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{figure_1}
+\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ is governed by the emission probability $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
+\label{fig:hmm-scheme}
+\end{figure}
+
 The estimation of an MSM requires the dynamics between microstates to be Markovian.
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{figure_1}
+\includegraphics[width=0.48\textwidth]{figure_2}
 \caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
@@
 \begin{figure}
-\includegraphics{figure_2}
+\includegraphics{figure_3}
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
@@
 \begin{figure}
-\includegraphics{figure_5}
+\includegraphics{figure_6}
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
@@
 \begin{figure}
-\includegraphics{figure_6}
+\includegraphics{figure_7}
 \caption{Example analysis of the conformational dynamics of a pentapeptide backbone:

### 2018-09-20 manuscript/manuscript.tex
addressing @franknoe's change requests

@@
 
-Our recommended method for dimensionality reduction, TICA, is a particular implementation of the VAC.
+A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC.
 To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
@@
 \begin{equation}
-\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i
+\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n,
 \end{equation}
 to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
-where the pairs of eigenvalues and independent components are in descending order.
+where the pairs of eigenvalues and independent components are sorted in descending order,
+and we define a cumulative kinetic variance fraction
+\begin{equation}
+c_d = \frac{\sum_{i=2}^d \lambda_i^2(\tau)}{\textrm{TKV}},
+\end{equation}
+where
+\begin{equation}
+\textrm{TKV} = \sum\limits_{i=2}^n \lambda_i^2(\tau)
+\end{equation}
+is the total kinetic variance explained by all $n$ features.
 
-Dimensionality reduction is achieved by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
-onto the leading $d$ independent components $\mathbf{U}=[\mathbf{u}_1 \dots \mathbf{u}_d]$,
+If we further scale the independent components $\mathbf{u}_i$ by the corresponding eigenvectors $\lambda_i(\tau)$,
+we obtain a \emph{kinetic map} which is the default behavior in PyEMMA.
+
+Note, though, that TICA requires the data to be in equilibrium.
+To use TICA with nonequilibrium data, we can either symmetrize the time-lagged covariance matrix $\mathbf{C}(\tau)$
+or apply a Koopman reweighting~\cite{vamp-preprint}.
+For short trajectories and nonequilibrium data we generally recommend to use VAMP~\cite{vamp-preprint}.
+
+For all these approaches,
+dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
+onto the leading $d$ independent components $\mathbf{U}_d=\left[\mathbf{u}_1 \dots \mathbf{u}_d\right]$,
 \begin{equation}
@@
 \end{equation}
-while retaining the kinetic variance
-\begin{equation}
-\textrm{KV}_d = \sum\limits_{i=1}^d \lambda_i^2(\tau);
-\end{equation}
-the total kinetic variance is the sum of the squares of all eigenvalues.
+where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., $95\%$).
 
@@
 Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
-and gives rise to our observed data, e.g., the discretized trajectories.
+and gives rise to our observed data, e.g., the discretized trajectories (see Fig.~\ref{fig:hmm-scheme}).
 This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
@@
 and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
-where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.
+where each row encodes the emission probabilities into the $n$ microstates conditioned on being in the corresponding hidden state~\cite{noe-proj-hid-msm}.
 
@@
 As the HMMs---like MSMs---approximate the full phase-space dynamics,
-we can similarly compute physical observables from directly.
+we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.
 

### 2018-09-20 manuscript/manuscript.tex
amending figure placement [ci skip]

@@
 \includegraphics[width=0.48\textwidth]{figure_1}
-\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ is governed by the emission probability $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
+\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ is governed by the emission probabilities $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
 \label{fig:hmm-scheme}
@@
 
-\begin{figure}
+\section{PyEMMA tutorials}
+
+\begin{figure}[bt]
 \includegraphics[width=0.48\textwidth]{figure_2}
@@
 \end{figure}
-
-\section{PyEMMA tutorials}
 
@@
 
-\begin{figure}
-\includegraphics{figure_3}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
-(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
-(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
-\label{fig:io-to-tica}
-\end{figure}
-
 Note that the modeler has to select hyper-parameters at most stages throughout the workflow.
@@
 In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
+
+\begin{figure}
+\includegraphics{figure_3}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
+(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
+(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
+(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
+\label{fig:io-to-tica}
+\end{figure}
 
@@
 
+\begin{figure}
+\includegraphics{figure_5}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
+(b)~PCCA++ identifies as five metastable states.
+(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
+in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
+(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
+\label{fig:msm-analysis}
+\end{figure}
+
+\begin{figure}
+\includegraphics{figure_6}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+visualization of the transition paths from $\mathcal{S}_2$ to $\mathcal{S}_4$.
+Metastable states $\mathcal{S}_{(1-5)}$ are represented by an ensemble of representative structures and are arranged along the horizonal axis according to their committor probabilities.
+The three main transition pathways starting from $\mathcal{S}_2$ and ending in $\mathcal{S}_4$ are depicted by gray arrows with thickness proportional to the transition flux.
+The dominant pathway proceeds through $\mathcal{S}_5$.}
+\label{fig:tpt-network}
+\end{figure}
+
 We can now directly extract several thermodynamic and kinetic properties from the estimated and validated model.
@@
 
-\begin{figure}
-\includegraphics{figure_5}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The reweighted free energy surface projected onto the first two independent components exhibits five minima which
-(b)~PCCA++ identifies as five metastable states.
-(c)~The second right eigenvector shows that the slowest process shifts probability between the least probable state ($\mathcal{S}_1$) and the other states,
-in particular states ($\mathcal{S}_4$, $\mathcal{S}_5$), whereas
-(d)~the committor $\mathcal{S}_2\to\mathcal{S}_4$ indicates that states $\mathcal{S}_{(1,3,5)}$ act as a transition region between states $\mathcal{S}_2$ and $\mathcal{S}_4$.}
-\label{fig:msm-analysis}
-\end{figure}
-
 The mean first passage times (MFPTs) out of and into the macrostate $\mathcal{S}_1$ compute to
@@
 It is easy to see from this depiction that the dominant pathway from $\mathcal{S}_2$ to $\mathcal{S}_4$ proceeds through $\mathcal{S}_5$.
-
-\begin{figure}
-\includegraphics{figure_6}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-visualization of the transition paths from $\mathcal{S}_2$ to $\mathcal{S}_4$.
-Metastable states $\mathcal{S}_{(1-5)}$ are represented by an ensemble of representative structures and are arranged along the horizonal axis according to their committor probabilities.
-The three main transition pathways starting from $\mathcal{S}_2$ and ending in $\mathcal{S}_4$ are depicted by gray arrows with thickness proportional to the transition flux.
-The dominant pathway proceeds through $\mathcal{S}_5$.}
-\label{fig:tpt-network}
-\end{figure}
 

### 2018-09-20 manuscript/manuscript.tex

@@
 
-PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming} (MSMs).
+PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming,noe2007jcp,chodera2007jcp,buchete-msm-2008} (MSMs).
 The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.
@@
 \begin{eqnarray}
-c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \\
-c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t,
+c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \label{eq:c0}\\
+c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t, \label{eq:ct}
 \end{eqnarray}
 where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
-Then, we can solve the generalized eigenvalue problem
+By default, PyEMMA estimates (\ref{eq:c0},\ref{eq:ct}) using symmetrization~\cite{tica}.
+This symmetrization induces a significant bias when using non-equilibrium data from short trajectories~\cite{hao-variational-koopman-models}.
+As an alternative, the so-called Koopman reweighting estimator is available which avoids this bias,
+but comes at the cost of a large variance~\cite{hao-variational-koopman-models}.
+
+After estimating the covariance matrices, TICA solves the generalized eigenvalue problem
 \begin{equation}
@@
 to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
-where the pairs of eigenvalues and independent components are sorted in descending order,
-and we define a cumulative kinetic variance fraction
+where the pairs of eigenvalues and independent components are sorted in descending order.
+A way to measure the contribution of each independent component to the kinetics
+is obtained by the kinetic distance~\cite{kinetic-maps}
+which assigns a cumulative variance fraction to the first $d$ independent components:
 \begin{equation}
@@
 If we further scale the independent components $\mathbf{u}_i$ by the corresponding eigenvectors $\lambda_i(\tau)$,
-we obtain a \emph{kinetic map} which is the default behavior in PyEMMA.
+we obtain a \emph{kinetic map}~\cite{kinetic-maps} which is the default behavior in PyEMMA.
 
-Note, though, that TICA requires the data to be in equilibrium.
-To use TICA with nonequilibrium data, we can either symmetrize the time-lagged covariance matrix $\mathbf{C}(\tau)$
-or apply a Koopman reweighting~\cite{vamp-preprint}.
-For short trajectories and nonequilibrium data we generally recommend to use VAMP~\cite{vamp-preprint}.
+Note, though, that TICA requires the dynamics to be simulated at equilibrium conditions.
+To use TICA with nonequilibrium MD, e.g., subject to external forces,
+or simply to perform dimension reduction on short trajectory data without worrying about reweighting,
+we recommend to use VAMP~\cite{vamp-preprint}.
 
@@
 \includegraphics[width=0.48\textwidth]{figure_1}
-\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ is governed by the emission probabilities $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
+\caption{The HMM transition matrix $\tilde{\mathbf{P}}(\tau)$ propagates the hidden state trajectory $\tilde{s}(t)$ (orange circles) and, at each time step $t$, the emission into the observable state $s(t)$ (cyan circles) is governed by the emission probabilities $\bm{\chi}\left( s(t) \middle| \tilde{s}(t) \right)$.}
 \label{fig:hmm-scheme}
@@
 An alternative, which is much less sensitive to poor discretization,
-is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
-HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space.
+is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
+HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
 Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
-and gives rise to our observed data, e.g., the discretized trajectories (see Fig.~\ref{fig:hmm-scheme}).
+and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
 This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
@@
 
-To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
+To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ timescale;
 in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
-Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
-\begin{equation}
-\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
-\end{equation}
-Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
-$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m<n$ hidden states' stationary probabilities,
-$\tilde{\mathbf{P}}(\tau)$ is a transition matrix between the $m<n$ hidden states at lag time $\tau$,
-and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
-where each row encodes the emission probabilities into the $n$ microstates conditioned on being in the corresponding hidden state~\cite{noe-proj-hid-msm}.
+The HMM then consists of a transition matrix $\tilde{\mathbf{P}}(\tau)$ between $m<n$ hidden states
+and a row-stochastic matrix ($\bm{\chi}$) of probabilities $\chi\left( s \middle| \tilde{s} \right)$
+to emit the discrete state $s$ conditional on being in the hidden state $\tilde{s}$.
+
+We can further compute a reversal of the emission matrix $\bm{\chi}\in\mathbb{R}^{m \times n}$:
+the membership matrix $\mathbf{M}\in\mathbb{R}^{n \times m}$ which encodes
+a fuzzy assignment of each of the $n$ observable microstates $s$ to the $m$ hidden states $\tilde{s}$ and,
+thus, defines the \emph{coarse graining} of microstate.
 
@@
 
-\begin{figure}[bt]
-\includegraphics[width=0.48\textwidth]{figure_2}
-\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
-A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
-By iterating between data processing and MSM estimation/validation,
-a dynamical model is obtained that can be analyzed (last row).}
-\label{fig:workflowchart}
-\end{figure}
-
 This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
@@
 
+\begin{figure}[bt]
+\includegraphics[width=0.48\textwidth]{figure_2}
+\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
+A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
+By iterating between data processing and MSM estimation/validation,
+a dynamical model is obtained that can be analyzed (last row).}
+\label{fig:workflowchart}
+\end{figure}
+
 \subsection{Feature selection}
@@
 In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
-
-\begin{figure}
-\includegraphics{figure_3}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
-(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
-(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
-\label{fig:io-to-tica}
-\end{figure}
 
@@
 We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.
+
+\begin{figure}
+\includegraphics{figure_3}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
+(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
+(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
+(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
+\label{fig:io-to-tica}
+\end{figure}
 

### 2018-11-12 manuscript/manuscript.tex
link sections to notebooks 01-08

@@
 In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
+In Notebook (01), we present in detail how to extract features from MD datasets and how to systematically compare them.
 
-Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
+Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
@@
 
+We demonstrate how to apply TICA, suggest how to interpret the projected coordinates and compare the results to other dimension reduction techniques in Notebook (02).
+
 \begin{figure}
@@
 The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
-which is shown in the showcase notebook (00).
+which is shown in the showcase Notebook (00).
+A detailed comparison between different clustering techniques is provided in Notebook (02).
 
@@
 long-timescale behavior of our system within error (blue/shaded area).
+
+In Notebook (03), we demonstrate in detail how to estimate and validate MSMs with PyEMMA.
 
@@
 
+More details about (spectral) properties of MSMs and how to analyze them with PyEMMA are discussed in Notebook (04) and Notebook (05).
+
 \subsection{Connecting the MSM with experimental data}
@@
 
+Besides a detailed demonstration of the above, Notebook (06) demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.
+
 \subsection{Summary}
@@
 More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
+We further demonstrate the symptoms of difficult data situations and how to deal with them in Notebook (08).
 

### 2018-11-12 manuscript/manuscript.tex

--- 2018-09-20 manuscript/manuscript.tex
@@
 Additional technical challenges for large systems include high demands on memory and computation time;
-we explain how to deal with those in the tutorials.
+we explain how to deal with those in the tutorials (Notebook 00 and 02).
 

### 2018-11-13 manuscript/manuscript.tex
BH's suggestion

@@
 
-We demonstrate how to apply TICA, suggest how to interpret the projected coordinates and compare the results to other dimension reduction techniques in Notebook (02).
+We demonstrate how to apply TICA, suggest how to interpret the projected coordinates, and compare the results to other dimension reduction techniques in Notebook (02).
 

### 2018-11-13 manuscript/manuscript.tex
BH's suggestion

@@
 
-Besides a detailed demonstration of the above, Notebook (06) demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.
+In addition to a detailed demonstration of the above, Notebook (06) demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.
 
@@
 More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
-We further demonstrate the symptoms of difficult data situations and how to deal with them in Notebook (08).
+We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in Notebook (08).
 

### 2018-11-14 manuscript/manuscript.tex
completed funding information (addresses #171)

--- 2018-11-12 manuscript/manuscript.tex
@@
 In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
-In Notebook (01), we present in detail how to extract features from MD datasets and how to systematically compare them.
 
-Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
+Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
@@
 
-We demonstrate how to apply TICA, suggest how to interpret the projected coordinates and compare the results to other dimension reduction techniques in Notebook (02).
-
 \begin{figure}
@@
 The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
-which is shown in the showcase Notebook (00).
-A detailed comparison between different clustering techniques is provided in Notebook (02).
+which is shown in the showcase notebook (00).
 
@@
 long-timescale behavior of our system within error (blue/shaded area).
-
-In Notebook (03), we demonstrate in detail how to estimate and validate MSMs with PyEMMA.
 
@@
 
-More details about (spectral) properties of MSMs and how to analyze them with PyEMMA are discussed in Notebook (04) and Notebook (05).
-
 \subsection{Connecting the MSM with experimental data}
@@
 
-Besides a detailed demonstration of the above, Notebook (06) demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.
-
 \subsection{Summary}
@@
 More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
-We further demonstrate the symptoms of difficult data situations and how to deal with them in Notebook (08).
 
@@
 %%%%%%%
-TH acknowledges financial support by SFB/TRR 186. SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.
+TH acknowledges financial support from Deutsche Forschungsgemeinschaft (SFB/TRR 186, Project A12).
+FN and BEH acknowledge funding from European Commission (ERC CoG 772230 "ScaleCell").
+FN acknowledges funding from Deutsche Forschungsgemeinschaft (SFB 1114, Projects A04 and C03, NO 825/2-2).
+SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.
 

### 2018-11-14 manuscript/manuscript.tex
typo (addresses #171)

@@
 
-In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances.
-(see~\cite{dror2012biomolecular} for a review).
+In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see~\cite{dror2012biomolecular} for a review).
 

### 2018-11-14 manuscript/manuscript.tex
decapitalize all instances of Notebook and fix notebook number typo (addresses #171)

@@
 we cannot anticipate this to be the case.
-We illustrate this point in Notebook~07.
+We illustrate this point in notebook~07.
 
@@
 where each state is considered to be metastable and,
-thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
+thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see notebook~07).
 As the HMMs---like MSMs---approximate the full phase-space dynamics,
@@
 To choose among three different molecular features reflecting protein structure,
-we compute the (cross-validated) VAMP-2 score (Notebook 00).
+we compute the (cross-validated) VAMP-2 score (notebook 00).
 Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
 it is important to ensure that properties that we optimize are robust as a function of lag time. 
-Consequently, we compute the VAMP-2 score at several lag times (Notebook 00). 
+Consequently, we compute the VAMP-2 score at several lag times (notebook 00). 
 We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
@@
 Additional technical challenges for large systems include high demands on memory and computation time;
-we explain how to deal with those in the tutorials (Notebook 00 and 02).
+we explain how to deal with those in the tutorials (notebooks~00 and~01).
 

### 2018-11-14 manuscript/manuscript.tex
fixed mistake in last commit and added the emma-project url (addresses #171) [ci skip]

@@
 Additional technical challenges for large systems include high demands on memory and computation time;
-we explain how to deal with those in the tutorials (notebooks~00 and~01).
+we explain how to deal with those in the tutorials (notebook~01).
 
@@
 However, recent years have seen many extensions of the methodology---many of which are available within PyEMMA. 
-We encourage interested readers to look into these methods in the software documentation and to make use of the specific Jupyter notebooks distributed with PyEMMA.
+We encourage interested readers to look into these methods in the software documentation
+and to make use of the specific Jupyter notebooks distributed with PyEMMA (\url{http://emma-project.org}).
 

### 2018-11-14 manuscript/manuscript.tex

@@
 
+We demonstrate how to apply TICA, suggest how to interpret the projected coordinates, and compare the results to other dimension reduction techniques in Notebook (02).
+
 \begin{figure}
@@
 
+In addition to a detailed demonstration of the above, Notebook (06) demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.
+
 \subsection{Summary}
@@
 More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
+We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in Notebook (08).
 

### 2018-11-14 manuscript/manuscript.tex
[manuscript] unify notation for notebook numbers

--- 2018-11-14 manuscript/manuscript.tex
@@
 This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
-The first notebook (00), which we will summarize in the following, showcases the entire estimation,
+The first notebook~00, which we will summarize in the following, showcases the entire estimation,
 validation, and analysis workflow for a small example system.
-The goal of this introductory notebook (00) is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
-The seven subsequent notebooks (01--07) provide in-depth lessons on specific topics,
-and the last notebook (08) contains guidelines on how to deal with common problems during MSM estimation.
+The goal of this introductory notebook~00 is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
+The seven subsequent notebooks~01--07 provide in-depth lessons on specific topics,
+and the last notebook~08 contains guidelines on how to deal with common problems during MSM estimation.
 
@@
 
-For the remainder of this manuscript we will walk through the first notebook (00).
+For the remainder of this manuscript we will walk through the first notebook~00.
 In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
@@
 thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
-The figures that will be displayed in the following are created in the showcase notebook (00) and can be easily reproduced.
+The figures that will be displayed in the following are created in the showcase notebook~00 and can be easily reproduced.
 
@@
 In particular, we can use a scalar score obtained using VAMP to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
+In notebook~01, we present in detail how to extract features from MD datasets and how to systematically compare them.
 
-Here, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
+Throughout this tutorial, we utilize the VAMP-2 score, which maximizes the kinetic variance contained in the features~\cite{kinetic-maps}.
 We should always evaluate the score in a cross-validated manner to ensure that we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
 To choose among three different molecular features reflecting protein structure,
-we compute the (cross-validated) VAMP-2 score (notebook 00).
+we compute the (cross-validated) VAMP-2 score (notebook~00).
 Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
 it is important to ensure that properties that we optimize are robust as a function of lag time. 
-Consequently, we compute the VAMP-2 score at several lag times (notebook 00). 
+Consequently, we compute the VAMP-2 score at several lag times (notebook~00). 
 We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
@@
 
+We demonstrate how to apply TICA, suggest how to interpret the projected coordinates, and compare the results to other dimension reduction techniques in notebook~02.
+
 \begin{figure}
@@
 The number of cluster centers has been chosen to optimize the VAMP-2 score in a manner identical to how the feature selection was carried out above,
-which is shown in the showcase notebook (00).
+which is shown in the showcase notebook~00.
+A detailed comparison between different clustering techniques is provided in notebook~02.
 
@@
 long-timescale behavior of our system within error (blue/shaded area).
+
+In notebook~03, we demonstrate in detail how to estimate and validate MSMs with PyEMMA.
 
@@
 
+More details about (spectral) properties of MSMs and how to analyze them with PyEMMA are discussed in notebook~04 and notebook~05.
+
 \subsection{Connecting the MSM with experimental data}
@@
 
+In addition to a detailed demonstration of the above, notebook~06 demonstrates how to compute J-couplings and dynamic fingerprints from MSMs.
+
 \subsection{Summary}
@@
 In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
-For the full analysis, please refer to the first notebook (00).
+For the full analysis, please refer to the first notebook~00.
 All notebooks as well as detailed installation instructions are available on \githubrepository{}.
@@
 More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
+We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in notebook~08.
 

### 2018-11-14 manuscript/manuscript.tex
[manuscript] simplify HMM section, restructure citations

@@
 An alternative, which is much less sensitive to poor discretization,
-is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
+is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
 HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
@@
 to emit the discrete state $s$ conditional on being in the hidden state $\tilde{s}$.
-
-We can further compute a reversal of the emission matrix $\bm{\chi}\in\mathbb{R}^{m \times n}$:
-the membership matrix $\mathbf{M}\in\mathbb{R}^{n \times m}$ which encodes
-a fuzzy assignment of each of the $n$ observable microstates $s$ to the $m$ hidden states $\tilde{s}$ and,
-thus, defines the \emph{coarse graining} of microstate.
+For more details about HMM properties and the estimation algorithm, we suggest Ref.~\cite{hmm-tutorial}.
 

### 2018-11-15 manuscript/manuscript.tex
hmm edits

@@
 transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
-hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint},
+hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
 and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
@@
 to emit the discrete state $s$ conditional on being in the hidden state $\tilde{s}$.
-For more details about HMM properties and the estimation algorithm, we suggest Ref.~\cite{hmm-tutorial}.
 
@@
 we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.
+
+For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
+For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
+A generalized extension for estimating this type of low-dimensional projection from the data is given in Ref.~\cite{wu2015projected}.
 

### 2018-11-15 manuscript/manuscript.tex
more literature changes

@@
 
-In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see~\cite{dror2012biomolecular} for a review).
+In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see Ref.~\cite{dror2012biomolecular} for a review on the MD simulation of biomolecules, and Ref.~\cite{mdtutorial} for a tutorial on MD simulations).
 
@@
 
-For those seeking further resources, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview,
-while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.
+For those seeking further resources, the recent perspective ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a timeline of methods advances with relevant citations,
+while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology and provides the underlying mathematics in detail.
 Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
@@
 In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
-we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
+we also recommend the literature on TICA~\cite{tica,tica3,kinetic-maps,mdtutorial},
 transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
@@
 	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
-	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps} is used to transform the features into a set of slow coordinates.
+	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,kinetic-maps} is used to transform the features into a set of slow coordinates.
 	\item Clustering -- This is the step at which the state decomposition occurs.
@@
 
-Subsequently, we perform TICA~\cite{tica,kinetic-maps} in order to reduce the dimension from the feature space,
+Subsequently, we perform TICA~\cite{tica,tica3,kinetic-maps} in order to reduce the dimension from the feature space,
 which typically contains many degrees of freedom,
@@
 we explain how to deal with those in the tutorials (notebook~01).
-
 More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
@@
 MEMMs are implemented in PyEMMA.
+Since the many publications associated with the development of these methods are beyond the scope of this tutorial, we refer the reader to Sec.~8.3 of Ref.~\cite{msm-brooke} and the references therein.
 

### 2018-11-15 manuscript/manuscript.tex
typo

@@
 In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
-we also recommend the literature on TICA~\cite{tica,tica3,kinetic-maps,mdtutorial},
+we also recommend the literature on TICA~\cite{tica,tica3,kinetic-maps,tica2},
 transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},

### 2018-11-15 manuscript/manuscript.tex

--- 2018-11-15 manuscript/manuscript.tex
@@
 \author[1\authfn{1}]{Tim Hempel}
-\author[2]{Brooke E. Husic}
+\author[1,2]{Brooke E. Husic}
 \author[1]{Simon Olsson}
@@
 transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
-hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
+hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint},
 and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
@@
 An alternative, which is much less sensitive to poor discretization,
-is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
+is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
 HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
@@
 
+We can further compute a reversal of the emission matrix $\bm{\chi}\in\mathbb{R}^{m \times n}$:
+the membership matrix $\mathbf{M}\in\mathbb{R}^{n \times m}$ which encodes
+a fuzzy assignment of each of the $n$ observable microstates $s$ to the $m$ hidden states $\tilde{s}$ and,
+thus, defines the \emph{coarse graining} of microstate.
+
 An HMM estimation always yields a model with a small number of (hidden) states
@@
 we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.
-
-For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
-For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
-A generalized extension for estimating this type of low-dimensional projection from the data is given in Ref.~\cite{wu2015projected}.
 

### 2018-11-16 manuscript/manuscript.tex
[manuscript] make caption fig 4 self-containing #171 [ci skip]

--- 2018-11-15 manuscript/manuscript.tex
@@
 (a)~The convergence behavior of the implied timescales associated with the four slowest processes.
+The solid lines refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
+The black line (marking equality of timescale and lag time) with grey area indicates the timescale horizon beyond which the MSM cannot resolve processes.
+As implied timescales are well-converged at $\tau=0.5$~ns, this lag time is chosen for subsequent MSM estimation.
 (b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 metastable states.
-The solid lines in (a) refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
-The black line indicates where implied timescales are equal to the lag time,
-whereas the grey area indicates all implied timescales faster than the lag time.
+Predictions from this model agree with higher lag time estimates within confidence intervals.
+Implied timescales convergence as well as a passing Chapman-Kolmogorov test are a necessary condition in MSM validation.
 In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}

### 2018-11-16 manuscript/manuscript.tex

@@
 The solid lines refer to the maximum likelihood result while the dashed lines show the ensemble mean computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.
-The black line (marking equality of timescale and lag time) with grey area indicates the timescale horizon beyond which the MSM cannot resolve processes.
+The black line (marking equality of timescale and lag time) with grey area indicates the timescale horizon below which the MSM cannot resolve processes.
 As implied timescales are well-converged at $\tau=0.5$~ns, this lag time is chosen for subsequent MSM estimation.

### 2018-11-19 manuscript/manuscript.tex
[manuscript] update @franknoe `s affiliation, #171
adding citations for CK test discussion (this fixes #171)

@@
 \author[1]{Simon Olsson}
-\author[1*]{Frank Noé}
+\author[1,3*]{Frank Noé}
 \affil[1]{Department of Mathematics and Computer Science, Freie Universität Berlin, Arnimallee 6, 14195 Berlin, Germany}
 \affil[2]{Department of Chemistry, Stanford University, 333 Campus Drive, Stanford, California 94305, USA}
+\affil[3]{Department of Chemistry, Rice University, 6100 Main Street, Houston, Texas 77005, USA}
 
@@
 where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
-By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model.
+By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model~\cite{msm-jhp,noe-folding-pathways}.
 

### 2018-11-19 manuscript/manuscript.tex
specify section in CK reference (this addresses #171)

@@
 
-Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $\mathbf{P}(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways}.
+Once we have used the ITS to choose the lag time, we can check whether a given transition probability matrix $\mathbf{P}(\tau)$ is approximately Markovian using the Chapman-Kolmogorov (CK) test~\cite{noe-folding-pathways,msm-jhp}.
 The CK property for a Markovian matrix is,
@@
 where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
-By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model~\cite{msm-jhp,noe-folding-pathways}.
+By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model (see Sec.~IV.F in~\cite{msm-jhp}).
 

### 2018-11-19 manuscript/manuscript.tex
use recommended float placing
fixing overfull boxes [ci skip]
added href to proper repository for binder
[ci skip]

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\hyphenation{Mar-kov}
+\setlength{\emergencystretch}{3em}
 
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics[width=0.48\textwidth]{figure_1}
@@
 We utilize Jupyter~\cite{jupyter} notebooks to show code examples along with figures and interactive widgets to display molecules.
-The user can install all necessary packages in one step using the \texttt{conda} command provided by the Anaconda Python stack (\url{https://anaconda.com}).
+The user can install all necessary packages in one step using the \texttt{conda} command provided by the Anaconda Python 
+stack (\url{https://conda.io/miniconda.html}).
 We recommend Anaconda because it resolves and installs dependencies as well as provides pre-compiled versions of common packages.
@@
 Should the user prefer not to use Anaconda, a manual installation via the pip installer is possible.
-Alternatively, one can use the Binder service (\url{https://mybinder.org}) to view and run the tutorials online in any browser.
+Alternatively, one can use the Binder service 
+(\href{https://mybinder.org/v2/gh/markovmodel/pyemma_tutorials/master?filepath=notebooks}{https://mybinder.org}) to view 
+and run the tutorials online in any browser.
 
@@
 \subsection{The PyEMMA workflow}
+
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]{figure_2}
+\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
+A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
+By iterating between data processing and MSM estimation/validation,
+a dynamical model is obtained that can be analyzed (last row).}
+\label{fig:workflowchart}
+\end{figure}
 
@@
 
-\begin{figure}[bt]
-\includegraphics[width=0.48\textwidth]{figure_2}
-\caption{The PyEMMA workflow: MD trajectories are processed and discretized (first row).
-A Markov state model is estimated from the resulting discrete trajectories and validated (middle row).
-By iterating between data processing and MSM estimation/validation,
-a dynamical model is obtained that can be analyzed (last row).}
-\label{fig:workflowchart}
+\subsection{Feature selection}
+
+\begin{figure}[bht]
+\includegraphics{figure_3}
+\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
+(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
+(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
+(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
+(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
+\label{fig:io-to-tica}
 \end{figure}
-
-\subsection{Feature selection}
 
@@
 
-\begin{figure}
-\includegraphics{figure_3}
-\caption{Example analysis of the conformational dynamics of a pentapeptide backbone:
-(a)~The Trp-Leu-Ala-Leu-Leu pentapeptide in licorice representation~\cite{vmd}.
-(b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
-(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
-\label{fig:io-to-tica}
-\end{figure}
-
 \subsection{Discretization}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics{figure_4}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics{figure_5}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics{figure_6}
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics{figure_7}

### 2018-11-28 manuscript/manuscript.tex
updating template [ci skip]

@@
 
-\title{Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber}
+\title{Introduction to Markov state modeling with the PyEMMA software [Article v\versionnumber]}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \pubDOI{10.XXXX/YYYYYYY}
@@
 \articlenum{<number>}
-\datereceived{July, 27, 2018}
-\dateaccepted{Month, Day, Year}
+\datereceived{Day Month Year}
+\dateaccepted{Day Month Year}
 
@@
 
-For a more detailed description of author contributions, see the GitHub issue tracking and changelog at \githubrepository{}.
+For a more detailed description of author contributions,
+see the GitHub issue tracking and changelog at \githubrepository.
 
@@
 
-For a more detailed description of contributions from the community and others, see the GitHub issue tracking and changelog at \githubrepository{}.
+For a more detailed description of contributions from the community and others,
+see the GitHub issue tracking and changelog at \githubrepository.
 

### 2018-11-28 manuscript/manuscript.tex
replace text-based reference with proper label

@@
 
-We further assume that the reader is familiar with the basic ideas and theory underlying Markov modeling and will only give a brief reminder of the basic concepts in Section 2.
+We further assume that the reader is familiar with the basic ideas and theory underlying Markov modeling and will only give a brief reminder of the basic concepts in Section~\ref{sec:prereq}.
 
@@
 \section{Prerequisites}
+\label{sec:prereq}
 

### 2018-11-28 manuscript/manuscript.tex
breaking up long lines and use consistent indentation
use proper siunitx [ci skip]

@@
 \end{equation}
-where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$, whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
+where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$,
+whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
 By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model (see Sec.~IV.F in~\cite{msm-jhp}).
@@
 \end{equation}
-where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., $95\%$).
+where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).
 
@@
 In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
-consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~$0.1$~ns.
+consisting of~$25$ independent MD trajectories conducted in implicit solvent with frames saved at an interval of~\SI{0.1}{\nano\second}.
 We present the results obtained in this notebook,
@@
 (b)~The VAMP-2 score indicates which of the tested featurizations contains the highest kinetic variance.
-(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=0.5$~ns shows multiple minima and
+(c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=\SI{0.5}{\nano\second}$ shows multiple minima and
 (d)~the time series of the first two ICs of the first trajectory show rare jumps.}
@@
 We find that the relative rankings of the different molecular features are highly robust as a function of lag time. 
-We show one example of this ranking and the absolute VAMP-2 scores for lag time~$0.5$~ns in Fig.~\ref{fig:io-to-tica}b. 
+We show one example of this ranking and the absolute VAMP-2 scores for lag time~\SI{0.5}{\nano\second} in Fig.~\ref{fig:io-to-tica}b. 
 We find that backbone torsions contain more kinetic variance than the backbone heavy atom positions or the distances between them (Fig.~\ref{fig:io-to-tica}b).
@@
 TICA is a special case of the variational principle~\cite{noe-vac,nueske-vamk} and is designed to find a projection preserving the long-timescale dynamics in the dataset.
-Here, performing TICA on the backbone torsions at lag time~$0.5$~ns yields a four dimensional subspace using a~$95\%$ kinetic variance cutoff
+Here, performing TICA on the backbone torsions at lag time~\SI{0.5}{\nano\second} yields a four dimensional subspace using a~\SI{95}{\percent} kinetic variance cutoff
 (note that we perform a $\cos/\sin$-transformation of the torsions before TICA in order to preserve their periodicity).
@@
 The black line (marking equality of timescale and lag time) with grey area indicates the timescale horizon below which the MSM cannot resolve processes.
-As implied timescales are well-converged at $\tau=0.5$~ns, this lag time is chosen for subsequent MSM estimation.
-(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=0.5$~ns assuming~5 metastable states.
+As implied timescales are well-converged at $\tau=\SI{0.5}{\nano\second}$, this lag time is chosen for subsequent MSM estimation.
+(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=\SI{0.5}{\nano\second}$ assuming~5 metastable states.
 Predictions from this model agree with higher lag time estimates within confidence intervals.
 Implied timescales convergence as well as a passing Chapman-Kolmogorov test are a necessary condition in MSM validation.
-In both panels, the (non-grey) shaded areas indicate~$95\%$ confidence intervals computed with the aforementioned Bayesian sampling procedure.}
+In both panels, the (non-grey) shaded areas indicate~\SI{95}{\percent} confidence intervals computed with the aforementioned Bayesian sampling procedure.}
 \label{fig:its-and-ck}
@@
 The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with~$100$ samples.
-In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above~$0.5$~ns (Fig.~\ref{fig:its-and-ck}a).
-Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5$~ns.
+In our example, we find that the four slowest ITS converge quickly and are constant within a \SI{95}{\percent} confidence interval for lag times above~\SI{0.5}{\nano\second} (Fig.~\ref{fig:its-and-ck}a).
+Using this lag time we can now estimate a (Bayesian) MSM with $\tau=\SI{0.5}{\nano\second}$.
 
@@
 Thus, the CK test confirms that five metastable states is an appropriate choice
-and shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the
+and shows that the MSM we have estimated at lag time $\tau=\SI{0.5}{\nano\second}$ indeed predicts the
 long-timescale behavior of our system within error (blue/shaded area).
@@
 \[ \begin{array}{crcr}
-\textrm{direction} & \textrm{mean / ns} && \textrm{std / ns} \\
+\textrm{direction} & \textrm{mean / \si{\nano\second}} && \textrm{std / \si{\nano\second}} \\
 \hline
@@
 The solid/orange lines denote the maximum likelihood MSM result;
-the dashed/blue lines and the the shaded areas indicate sample means and~$95\%$ confidence intervals computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.}
+the dashed/blue lines and the the shaded areas indicate sample means and~\SI{95}{\percent} confidence intervals computed with a Bayesian sampling procedure~\cite{ben-rev-msm}.}
 \label{fig:msm-exp-obs}

### 2018-11-29 manuscript/manuscript.tex
some edits

@@
 
-PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models~\cite{schuette-msm,singhal-msm-naming,noe2007jcp,chodera2007jcp,buchete-msm-2008} (MSMs).
+PyEMMA~\cite{pyemma} (\url{http://emma-project.org}) is a software for the analysis of molecular dynamics (MD) simulations using Markov state models (MSMs)~\cite{schuette-msm,singhal-msm-naming,noe2007jcp,chodera2007jcp,buchete-msm-2008}.
 The package is written in Python (\url{http://python.org}), relies heavily on NumPy/SciPy~\cite{numpy,scipy}, and is compatible with the scikit-learn~\cite{sklearn} framework for machine learning.
@@
 
-In this tutorial, we assume that the reader is familiar with MD simulation and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see Ref.~\cite{dror2012biomolecular} for a review on the MD simulation of biomolecules, and Ref.~\cite{mdtutorial} for a tutorial on MD simulations).
+In this tutorial, we assume that the reader is familiar with MD and standard analysis of MD simulations of peptides and proteins, such as computation of torsion angles and distances (see Ref.~\cite{dror2012biomolecular} for a review on the MD simulation of biomolecules, and Ref.~\cite{mdtutorial} for a tutorial on MD simulations).
 
-We further assume that the reader is familiar with the basic ideas and theory underlying Markov modeling and will only give a brief reminder of the basic concepts in Section~\ref{sec:prereq}.
+We further assume that the reader is familiar with the basic ideas and theory underlying Markov state modeling and will only give a brief reminder of the basic concepts in Sec.~\ref{sec:prereq}.
 
 For those seeking further resources, the recent perspective ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a timeline of methods advances with relevant citations,
-while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology and provides the underlying mathematics in detail.
+while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology and details the underlying mathematics.
 Additionally, two textbooks have been published that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.
@@
 In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
-we also recommend the literature on TICA~\cite{tica,tica3,kinetic-maps,tica2},
+we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,kinetic-maps,tica2},
 transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
@@
 
-Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
+Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high dimensional MD simulation datasets.
 In its standard formulation, the creation of an MSM involves decomposing the phase or configuration space occupied by a dynamical system into a set of disjoint, discrete states,
@@
 While most analysis techniques require simulation trajectories to be long enough to sample from the equilibrium distribution, this is not required for MSMs.
-Because MSMs are using the \emph{conditional} probability $p_{ij}(\tau)$,
-they are useful for the analysis of short simulation trajectories with arbitrary starting points---see~\cite{oom-feliks} for restrictions.
+Because MSMs use the \emph{conditional} probability $p_{ij}(\tau)$,
+they are useful for the analysis of short simulation trajectories with arbitrary starting points---see Ref.~\cite{oom-feliks} for restrictions.
 \item Markovianity:
@@
 Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
-Then, the count matrix is converted to a row-stochastic transition probability matrix $\mathbf{P}$, which is defined for the specified lag time.
-For MD simulations in equilibrium, $\mathbf{P}$ should obey detailed balance which is enforced by constraining the estimation of $\mathbf{P}$ to the following equations:
+Then, the count matrix is converted to a row-stochastic transition probability matrix $\mathbf{P}(\tau)$, which is defined for the specified lag time.
+For MD simulations in equilibrium, $\mathbf{P}(\tau)$ should obey detailed balance which is enforced by constraining the estimation of $\mathbf{P}(\tau)$ to the following equations:
 \begin{equation}
@@
 The constraints (\ref{eq:balance}) are omitted if MD simulations are not conducted in equilibrium, e.g.,
-for systems experiencing a pulling force or an external potential---see~\cite{Koltai2018} for a recent review on nonequilibrium MSMs.
-For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces and (\ref{eq:balance}) to hold.
+for systems experiencing a pulling force or an external potential (see Ref.~\cite{Koltai2018} for a recent review on nonequilibrium MSMs).
+For the remainder of this section we will simplify the matter by assuming the more common scenario of MD simulations without external forces, such that Eqn.~(\ref{eq:balance}) is assumed to hold.
 
@@
 \end{equation}
-where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$, where $k$ is an integer larger than~$1$,
-whereas the right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
-By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model (see Sec.~IV.F in~\cite{msm-jhp}).
+where the left-hand side of the equation corresponds to an MSM estimated at lag time $k\tau$ and $k$ is an integer larger than~$1$.
+The right-hand side of the equation is our estimated MSM transition probability matrix to the $k^\textrm{th}$ power.
+By assessing how well the approximated transition probability matrix adheres to the CK property, we can validate the appropriateness of the Markovian assumption for the model (see Sec.~IV.F in Ref.~\cite{msm-jhp}).
 
@@
 The subsequent eigenvalues $\lambda_{i>1}(\tau)$ are real with absolute values less than~$1$ and are related to the \emph{characteristic} or \emph{implied} timescales of dynamical processes within the system (eq.~\ref{eq:its}).
-The dynamical process themself (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
+The dynamical processes themselves (for $i>1$) are encoded by the right eigenvectors $\bm{\psi}_i$,
 \begin{equation}
@@
 \end{equation}
-where the eigenvalue-eigenvector pairs are indexed in decreasing order.
+where the eigenvalue-eigenvector pairs are indexed in decreasing order according to the eigenvalues.
 The coefficients of the eigenvectors represent the flux into and out of the Markov states that characterize the corresponding process.
@@
 \begin{itemize}
-	\item Featurization -- The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
-	\item Dimensionality reduction -- Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
-	Frequently, time-lagged independent component analysis (TICA)~\cite{tica,tica3,kinetic-maps} is used to transform the features into a set of slow coordinates.
-	\item Clustering -- This is the step at which the state decomposition occurs.
+	\item Featurization: The Cartesian coordinates characterizing each frame of the MD trajectory are transformed into an intuitive basis such as the protein's dihedral angles or contact distance pairs.
+	\item Dimensionality reduction: Optionally, a basis set transformation can be performed that produces a linear (or nonlinear) combination of the features in the previous step.
+	Frequently, TICA~\cite{tica,tica3,kinetic-maps} is used to transform the features into a set of slow coordinates.
+	\item Clustering: This is the step at which the state decomposition occurs.
 	The features or TICs are grouped into a set of states using a clustering algorithm such as $k$-means.
-	\item Transition matrix approximation -- At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
+	\item Transition matrix approximation: At this stage, transitions are counted at a pre-specified lag time, and the estimation and validation described in the previous section are performed.
 \end{itemize}
@@
 
-A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC.
+A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC~\cite{tica}.
 To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
@@
 where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
-By default, PyEMMA estimates (\ref{eq:c0},\ref{eq:ct}) using symmetrization~\cite{tica}.
-This symmetrization induces a significant bias when using non-equilibrium data from short trajectories~\cite{hao-variational-koopman-models}.
+By default, PyEMMA estimates Eqns.~(\ref{eq:c0}) and~(\ref{eq:ct}) using symmetrization~\cite{tica}.
+This symmetrization induces a significant bias when using nonequilibrium data from short trajectories~\cite{hao-variational-koopman-models}.
 As an alternative, the so-called Koopman reweighting estimator is available which avoids this bias,
@@
 \begin{equation}
-\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n,
+\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n
 \end{equation}
@@
 or simply to perform dimension reduction on short trajectory data without worrying about reweighting,
-we recommend to use VAMP~\cite{vamp-preprint}.
+we recommend to use VAMP, which does not require reversible dynamics~\cite{vamp-preprint}.
 
-For all these approaches,
+For all of these approaches,
 dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
@@
 HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
-Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
+Instead, HMMs assume that there is an underlying (hidden) dynamic process that is Markovian
 and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
-This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
+This is a powerful principle as we know that there is indeed an underlying process that is Markovian:
 our molecular dynamics trajectories.
@@
 An HMM estimation always yields a model with a small number of (hidden) states
-where each state is considered to be metastable and,
+in which each state is considered to be metastable and,
 thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see notebook~07).
@@
 For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
-A generalized extension for estimating this type of low-dimensional projection from the data is given in Ref.~\cite{wu2015projected}.
+A generalized extension for estimating this type of low dimensional projection from the data is given in Ref.~\cite{wu2015projected}.
 
@@
 stack (\url{https://conda.io/miniconda.html}).
-We recommend Anaconda because it resolves and installs dependencies as well as provides pre-compiled versions of common packages.
+We recommend Anaconda because it resolves and installs dependencies and provides pre-compiled versions of common packages.
 The tutorial installation contains a launcher command to start the Jupyter notebook server as well as the notebook files.
 
-You can install the tutorial's dependencies in a new conda environment and start the notebook server via
+The user can install the tutorial's dependencies in a new conda environment and start the notebook server via
 \begin{verbatim}
@@
 \begin{itemize}
-    \item \textbf{PyEMMA} -- MSM/HMM estimation, validation, analysis, and visualization, and its dependencies~\cite{pyemma}
-    \item mdshare -- A downloader for MD data from a public server
-    \item notebook -- The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
-    \item matplotlib -- A plotting library~\cite{matplotlib}
-    \item nglview -- Widget for active viewing of molecular structures in Jupyter environments~\cite{nglview}
+    \item \textbf{PyEMMA}: MSM/HMM estimation, validation, analysis, and visualization, and its dependencies~\cite{pyemma}
+    \item mdshare: A downloader for MD data from a public server
+    \item notebook: The Jupyter~\cite{jupyter} notebook tool used for running the tutorials, along with extension packages jupyter\_contrib\_nbextensions and nbexamples
+    \item matplotlib: A plotting library~\cite{matplotlib}
+    \item nglview: Widget for active viewing of molecular structures in Jupyter environments~\cite{nglview}
 \end{itemize}
@@
 This tutorial consists of nine Jupyter notebooks which introduce the basic features of PyEMMA.
-The first notebook~00, which we will summarize in the following, showcases the entire estimation,
+The first notebook~(00), which we will summarize in the following, showcases the entire estimation,
 validation, and analysis workflow for a small example system.
-The goal of this introductory notebook~00 is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
-The seven subsequent notebooks~01--07 provide in-depth lessons on specific topics,
-and the last notebook~08 contains guidelines on how to deal with common problems during MSM estimation.
+The goal of this introductory notebook is to provide the user with the typical steps required to obtain a validated MSM analysis of protein or peptide simulation data.
+The seven subsequent notebooks~(01--07) provide in-depth lessons on specific topics,
+and the last notebook~(08) contains guidelines on how to deal with common problems during MSM estimation.
 
@@
 
-In short, the workflow (Fig.~\ref{fig:workflowchart}) for a full analysis of an MD dataset might consist of,
+In short, the workflow (Fig.~\ref{fig:workflowchart}) for a full analysis of an MD dataset might consist of
 \begin{itemize}
@@
 
-For the remainder of this manuscript we will walk through the first notebook~00.
+For the remainder of this manuscript we will walk through the first notebook~(00).
 In notebook~00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide (Fig.~\ref{fig:io-to-tica}a),
@@
 thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
-The figures that will be displayed in the following are created in the showcase notebook~00 and can be easily reproduced.
+The figures that will be displayed in the following sections are created in the showcase notebook~00 and can be easily reproduced.
 
@@
 (c)~The sample free energy projected onto the first two time-lagged independent components (ICs) at lag time $\tau=\SI{0.5}{\nano\second}$ shows multiple minima and
-(d)~the time series of the first two ICs of the first trajectory show rare jumps.}
+(d)~the time series of the first two ICs of the first trajectory show rare transitions.}
 \label{fig:io-to-tica}
@@
 Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk}
-and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint}
+and the more general VAMP~\cite{vamp-preprint}
 provide a systematic means to quantitatively compare multiple representations of the simulation data.
@@
 we compute the (cross-validated) VAMP-2 score (notebook~00).
-Although we cannot MSM optimize lag times with a variational score\cite{husic2017note}, such as VAMP-2,
-it is important to ensure that properties that we optimize are robust as a function of lag time. 
+Although we cannot optimize MSM lag times with a variational score\cite{husic2017note}, such as VAMP-2,
+it is important to ensure that the properties we optimize are robust as a function of lag time. 
 Consequently, we compute the VAMP-2 score at several lag times (notebook~00). 
@@
 The sample free energy projected onto the first two independent components (ICs) exhibits several minima (Fig.~\ref{fig:io-to-tica}c).
-Discrete jumps between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
+Discrete transitions between the minima can be observed by visualizing the transformation of the first trajectory into these ICs (Fig.~\ref{fig:io-to-tica}d).
 We thus assume that our TICA-transformed backbone torsion features describe one or more metastable processes.
@@
 
-\subsection{Analyzing the MSM}
+\subsection{MSM Analysis}
 
@@
 In order to interpret the slowest relaxation timescales, we refer to the (right) eigenvectors,
-as they are independent of the stationary distribution (see Section~\ref{sec:theory}).
+as they are independent of the stationary distribution (see Sec.~\ref{sec:theory}).
 This enables us to specifically study what conformational changes are happening on a particular timescale independently of the equilbrium distribution.
@@
 The second right eigenvector, on the other hand, corresponds to the slowest process in the system. 
-Note that the eigenvectors are real as detailed balance has been enforced during MSM estimation.
+Note that the eigenvectors are real because detailed balance has been enforced during MSM estimation.
 The minimal and maximal components of the second right eigenvector indicate the microstates between which the process shifts probability density.
 The relaxation timescale of this exchange process is exactly the corresponding implied timescale,
-which can be computed from its corresponding eigenvalue using~\eqref{eq:its}.
+which can be computed from its corresponding eigenvalue using Eqn.~(\ref{eq:its}).
 In the projection onto the first two TICA components,
@@
 TPT~\cite{weinan-tpt,metzner-msm-tpt} is a method used to analyze the statistics of transition pathways.
-The TPT version of~\cite{noe-folding-pathways} can be conveniently applied to the estimated MSM.
+TPT as implemented in Ref.~\cite{noe-folding-pathways} can be conveniently applied to the estimated MSM.
 Here, we compute the TPT flux between macrostates $\mathcal{S}_2$ and $\mathcal{S}_4$ (Fig.~\ref{fig:msm-analysis}d).
@@
 
-As an example, here we look at the fluorescence correlation of Trp-1,
+As an example, we look at the fluorescence correlation of Trp-1,
 since this terminal tryptophan is a realistic experimental observable for our pentapeptide system.
@@
 Now that we have an approximation of the fluorescence in each of our MSM states,
-we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (\ref{fig:msm-exp-obs}a).
+we can use PyEMMA to compute the fluorescence autocorrelation function (ACF) from our MSM (Fig.~\ref{fig:msm-exp-obs}a).
 Note how the computed ACF has a very small response (i.e., signal amplitude).
@@
 To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$
-and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-exp-obs}b).
+and follow the predicted fluorescence signal as it relaxes to equilibrium (Fig.~\ref{fig:msm-exp-obs}b).
 We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization,
@@
 In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA.
-For the full analysis, please refer to the first notebook~00.
+For the full analysis, please refer to the first notebook~(00).
 All notebooks as well as detailed installation instructions are available on \githubrepository{}.
@@
 and thus cluster assignments based on that norm may yield a poor discretization.
-Especially for large systems, it is thus particularly important to first find a suitable set of features,
+Especially for large systems, it is particularly important to first find a suitable set of features,
 and to further apply dimensionality reduction techniques (e.g., TICA, VAMP, if applicable)
@@
 we explain how to deal with those in the tutorials (notebook~01).
-More details on how to model complex systems with the techniques presented here are described, e.g., by~\cite{plattner_protein_2015,plattner_complete_2017}.
+More details on how to model complex systems with the techniques presented here are described, e.g., by Refs.~\cite{plattner_protein_2015,plattner_complete_2017}.
 We further examine some symptoms that may indicate problematic or difficult datasets, and demonstrate how to deal with them in notebook~08.
@@
 The present tutorial presents the basics of modern Markov state modeling with PyEMMA. 
-However, recent years have seen many extensions of the methodology---many of which are available within PyEMMA. 
+However, recent years have seen many extensions of the methodology, many of which are available within PyEMMA. 
 We encourage interested readers to look into these methods in the software documentation
@@
 Conventional Markov state modeling often relies on large simulation datasets to ensure proper convergence of thermodynamic and kinetic properties. 
-In one extension, Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram},
+In one extension, multi-ensemble Markov models (MEMMs)~\cite{dtram,tram},
 we can integrate unbiased and biased simulations in a systematic manner to speed up the convergence. 
@@
 taking into account their respective uncertainties~\cite{simon-amm}. 
-AMMs are implemented in PyEMMA.
+AMMs are also implemented in PyEMMA.
 

### 2018-11-29 manuscript/manuscript.tex
chapman-kolmogorov > ck

@@
 As implied timescales are well-converged at $\tau=\SI{0.5}{\nano\second}$, this lag time is chosen for subsequent MSM estimation.
-(b)~Chapman-Kolmogorov test computed using an MSM estimated with lag time $\tau=\SI{0.5}{\nano\second}$ assuming~5 metastable states.
+(b)~CK test computed using an MSM estimated with lag time $\tau=\SI{0.5}{\nano\second}$ assuming~5 metastable states.
 Predictions from this model agree with higher lag time estimates within confidence intervals.
-Implied timescales convergence as well as a passing Chapman-Kolmogorov test are a necessary condition in MSM validation.
+Implied timescales convergence as well as a passing CK test are a necessary condition in MSM validation.
 In both panels, the (non-grey) shaded areas indicate~\SI{95}{\percent} confidence intervals computed with the aforementioned Bayesian sampling procedure.}
@@
 
-To test the validity of our MSM, we perform a Chapman-Kolmogorov (CK) test.
+To test the validity of our MSM, we perform a CK test.
 Visualizing the full transition probability matrix $T$ is difficult;

### 2018-11-29 manuscript/manuscript.tex

@@
 The VAMP can thus be used to perform model selection.
-Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model.
+Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model~\cite{kinetic-maps}.
 However, the MSM lag time cannot be optimized using VAMP,
@@
 \end{equation}
-where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).
+where, in practice, $d$ is preferably chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).
 
@@
 we cannot anticipate this to be the case.
-We illustrate this point in notebook~07.
 
@@
 Instead, HMMs assume that there is an underlying (hidden) dynamic process that is Markovian
-and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
+and gives rise to our observed data, i.e., the ($n$~states) discretized trajectories $s(t)$.
 This is a powerful principle as we know that there is indeed an underlying process that is Markovian:
@@
 in which each state is considered to be metastable and,
-thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see notebook~07).
+thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully.
 As the HMMs---like MSMs---approximate the full phase-space dynamics,
@@
 A generalized extension for estimating this type of low dimensional projection from the data is given in Ref.~\cite{wu2015projected}.
+One of our tutorial notebooks, to be discussed in the next section, provides an example of HMM analysis.
 
@@
 
-Using PyEMMA, we can simulate the relaxation of an observable if we had prepared our molecular system in a nonequilibrium initial condition.
+Using PyEMMA, we can simulate the relaxation of an observable from a nonequilibrium initial condition.
 The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
@@
 
-\subsection{Summary}
+\subsection{Summary of the showcase notebook}
 
@@
 %%%%%%%
+MKS acknowledges financial support from European Commission (ERC StG 307494 "pcCell").
 TH acknowledges financial support from Deutsche Forschungsgemeinschaft (SFB/TRR 186, Project A12).
+SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.
+FN and MKS acknowledge funding from Deutsche Forschungsgemeinschaft (SFB 1114, Projects A04 and C03, NO 825/2-2).
 FN and BEH acknowledge funding from European Commission (ERC CoG 772230 "ScaleCell").
-FN acknowledges funding from Deutsche Forschungsgemeinschaft (SFB 1114, Projects A04 and C03, NO 825/2-2).
-SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.
 

### 2018-12-10 manuscript/manuscript.tex
adding doi + pubversion option
Update manuscript.tex
Fix the URL for the github repo.

@@
 %%% PREAMBLE
-\documentclass[9pt,tutorial]{livecoms}
+\documentclass[9pt,tutorial,pubversion]{livecoms}
 % Use the 'onehalfspacing' option for 1.5 line spacing
@@
 \newcommand{\versionnumber}{1.0}
-\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}
+\newcommand{\githubrepository}{\url{https://github.com/markovmodel/pyemma_tutorials}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\pubDOI{10.XXXX/YYYYYYY}
-\pubvolume{<volume>}
-\pubyear{<year>}
-\articlenum{<number>}
-\datereceived{Day Month Year}
-\dateaccepted{Day Month Year}
+\pubDOI{10.33011/livecoms.1.1.5965}
+\pubvolume{1}
+\pubyear{2018}
+\articlenum{1}
+\datereceived{27 July 2018}
+\dateaccepted{27 November 2018}
 
@@
 and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research in its own right.
-Implementations of some of these methods are available or are under development in the deeptime package \url{github.com/markovmodel/deeptime}.
+Implementations of some of these methods are available or are under development in the deeptime package \url{https://github.com/markovmodel/deeptime}.
 
